### 2014-06-18 doc/pseudocode.tex
Added pseudocode (from Eduard).

\documentclass[10pt,onecolumn]{article}

\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}

\begin{document}

\begin{algorithm}
  \caption{Starcode algorithm}
  \label{alg:starcode}
  \begin{algorithmic}[1]
    \State Define: $\tau$
    \State \textbf{Initialize variables:} $seed$, $start = 0$,
    $height$, $seq$, $trie$
    \State \textbf{Initialize containers:} $hits$, $pebbles$
    \State \textsc{read} sequence file
    \State $height \gets$ \textsc{determine} maximum sequence length
    \State \textsc{pad} sequences up to $height$
    \State \textsc{sort} sequences alphabetically
    \State $trie \gets$ \textsc{create} an empty trie of height $height$
    \State \textsc{insert} root node of $trie$ in $pebbles$ at depth 0
    \ForAll{sequences}
    \State $seq \gets$ \textsc{get} next sequence
    \State $seed \gets$ \textsc{length} of shared prefix
    between current and next $seq$
    \State \textsc{clear} hits
    \State \textsc{clear} $pebbles$ at depth $>start$
    \ForAll{$pebbles$ at depth $start$}
    \State $node \gets$ \textsc{get} next node from $pebbles$
    \State \textbf{call} \textsc{poucet}($node$, $seed$, $hits$)
    \EndFor
    \State \textsc{insert} $seq$ path in $trie$
    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
    \State $start \gets seed$
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}
  \caption{Poucet search algorithm}
  \label{alg:poucet}
  \begin{algorithmic}[1]
    \Procedure{poucet}{$node$, $seed$, $hits$}:
    \State \textsc{compute} $node$-specific column values following NW algorithm
    \ForAll{$child$ nodes in $node$}
    \State \textsc{compute} $child$-specific row values following NW algorithm
    \State \textsc{compute} center value using row and column
    values
    \If{center value $>\tau$} \Comment{Mismatches exceeded.}
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth = $height$} \Comment{Hit found.}
    \State \textsc{save} $node$ sequence in $hits$
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth $\leq seed$}
    \State \textsc{save} $node$ in pebbles at current depth
    \EndIf
    \State \textbf{call} poucet($child$, $seed$, $hits$)
    \EndFor
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\end{document}

### 2014-06-18 main-tquery.c

@@ -1,217 +1,0 @@
-#include "starcode.h"
-
-char *USAGE = "Usage:\n"
-"  tquery [-v] [-d dist] [-i input] [-q query] [-o output]\n"
-"    -v --verbose: verbose\n"
-"    -d --dist: maximum Levenshtein distance\n"
-"    -t --threads: number of concurrent threads (default 2)\n"
-"    -l --level: number of subtries (default 3*threads)\n"
-"    -i --index: index file\n"
-"    -q --query: query file\n"
-"    -o --output: output file (default stdout)";
-
-void say_usage(void) { fprintf(stderr, "%s\n", USAGE); }
-
-void SIGSEGV_handler(int sig) {
-  void *array[10];
-  size_t size;
-
-  // get void*'s for all entries on the stack
-  size = backtrace(array, 10);
-
-  // print out all the frames to stderr
-  fprintf(stderr, "Error: signal %d:\n", sig);
-  backtrace_symbols_fd(array, size, STDERR_FILENO);
-  exit(1);
-}
-
-int
-main(
-   int argc,
-   char **argv
-)
-{
-   signal(SIGSEGV, SIGSEGV_handler);
-
-   // Unset flags (value -1).
-   int verbose_flag = -1;
-   int dist_flag = -1;
-   int threads_flag = -1;
-   int ntries_flag = -1;
-   // Unset options (value 'UNSET').
-   char _u; char *UNSET = &_u;
-   char *index = UNSET;
-   char *query = UNSET;
-   char *output = UNSET;
-
-   int c;
-   while (1) {
-      int option_index = 0;
-      static struct option long_options[] = {
-         {"verbose", no_argument,       0, 'v'},
-         {"help",    no_argument,       0, 'h'},
-         {"dist",    required_argument, 0, 'd'},
-         {"index",   required_argument, 0, 'i'},
-         {"query",   required_argument, 0, 'q'},
-         {"output",  required_argument, 0, 'o'},
-         {0, 0, 0, 0}
-      };
-
-      c = getopt_long(argc, argv, "t:l:d:i:ho:q:v",
-            long_options, &option_index);
- 
-      /* Detect the end of the options. */
-      if (c == -1) break;
-  
-      switch (c) {
-         case 'd':
-            if (dist_flag < 0) {
-               dist_flag = atoi(optarg);
-            }
-            else {
-               fprintf(stderr, "distance option set more than once\n");
-               say_usage();
-               return 1;
-            }
-            break;
-
[…]

### 2014-06-18 trie.c

@@ -33,24 +33,24 @@ search
    const int         tau,
 //         hstack_t ** hits,
          gstack_t ** hits,
-         int         start,
-   const int         trail
+         int         start_depth,
+   const int         seed_depth
 )
 // SYNOPSIS:                                                              
-//   Front end query of a trie with the "trail search" algorithm. Search  
+//   Front end query of a trie with the "poucet search" algorithm. Search 
 //   does not start from root. Instead, it starts from a given depth      
 //   corresponding to the length of the prefix from the previous search.  
 //   Since initial computations are identical for queries starting with   
 //   the same prefix, the search can restart from there. Each call to     
-//   the function starts ahead and leaves a trail for the next query.     
+//   the function starts ahead and seeds pebbles for the next query.      
 //                                                                        
 // PARAMETERS:                                                            
 //   trie: the trie to query                                              
 //   query: the query as an ascii string                                  
 //   tau: the maximum edit distance                                       
 //   hits: a hit stack to push the hits                                   
-//   start: the depth to start the search                                 
-//   trail: how deep to trail                                             
+//   start_depth: the depth to start the search                           
+//   seed_depth: how deep to seed pebbles                                 
 //                                                                        
 // RETURN:                                                                
 //   A pointer to 'hits' node array.                                      
@@ -84,8 +84,8 @@ search
    info_t *info = trie->info;
 
    // Reset the pebbles that will be overwritten.
-   start = max(start, 0);
-   for (int i = start+1 ; i <= min(trail, height) ; i++) {
+   start_depth = max(start_depth, 0);
+   for (int i = start_depth+1 ; i <= min(seed_depth, height) ; i++) {
       info->pebbles[i]->nitems = 0;
    }
 
@@ -94,7 +94,7 @@ search
    int translated[M];
    translated[0] = length;
    translated[length+1] = EOS;
-   for (int i = max(0, start-maxtau) ; i < length ; i++) {
+   for (int i = max(0, start_depth-maxtau) ; i < length ; i++) {
       translated[i+1] = altranslate[(int) query[i]];
    }
 
@@ -105,15 +105,15 @@ search
       .tau     = tau,
       .maxtau  = maxtau,
       .pebbles = info->pebbles,
-      .trail   = trail,
[…]

### 2014-06-18 trie.h

@@ -66,7 +66,7 @@ struct arg_t {
    char        tau;
    char        maxtau;
    int       * query;
-   int         trail;
+   int         seed_depth;
    int         height;
    int         err;
 };

### 2014-06-19 doc/document.tex
Added LaTeX manuscript (Bioinformatics).

\documentclass{bioinfo}
\copyrightyear{2014}
\pubyear{2014}

\begin{document}
\firstpage{1}

\title[Exact sequence clustering]{Starcode: an exact algorithm for sequence clustering}
\author[Valera Zorita \textit{et~al}]{Eduard Valera Zorita\,$^{1,2}$, Pol Cusc\'o\,$^{1,2}$ and Guillaume Filion\,$^{1,2}$\footnote{to whom correspondence should be addressed}}
\address{$^{1}$Genome Architecture, Gene Regulation, Stem Cells and Cancer Programme, Centre for Genomic Regulation (CRG), Dr. Aiguader 88, 08003 Barcelona, Spain.\\
$^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\maketitle

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new applications
and challenges for computational biology. One such application is the use of
random barcodes to trace and quantify transcripts or lineages in experimentals
setups. The high error rate of modern sequencers calls for additional
post-processing techniques capable of detecting and reverting the misreads.
However, in the absence of a reference population, the problem amounts to
performing a pairwise comparison of all the barcodes, which is unfeasible for
excessive computationally complexity.

\section{Results:}
Here we address this problem and describe an exact algorithm to determine which
pairs of sequences lie within a given Levenshtein distance. The matched pairs
are merged into clusters represented by a canonical sequence. The speed of
starcode is attributable to the poucet search, a novel  implementation of the
Needleman-Wunsch algorithm performed on the nodes of a trie. Parallelization
achieves linear scaling of performance on multi-core machines. On the task of
clustering random barcodes, starcode outperforms short read mappers and sequence
clustering algorithms in both speed and precision.

\section{Availability and implementation:}
The C source code is available at http://github.com/gui11aume/starcode.

\section{Contact:} \href{guillaume.filion@gmail.com}{guillaume.filion@gmail.com}
\end{abstract}

\section{Introduction}

Sequence clustering is the process of grouping similar biological sequences.
It has been traditionally applied to identify related protein families and to
reduce sequence redundancy in databases. Recently, the advent of high throughput
sequencing has created additional needs for efficient clustering algorithms, in
particular because of the high error rate of such technologies. For instance,
the Illumina platform \citep{pmid16056220} shows a 1-2\% error rate consisting ofs
 substitutions near the 3' end of the read \citep{pmid18660515, pmid21576222}. The
PacBio platform shows a 15\% error rate that mostly correspond to insertions
and deletions \citep{pmid19023044}. As a consequence, the same sequence is often
decoded in different ways, which artificially increases the diversity of the
output.

Sequencing errors can be discovered by mapping the reads onto a reference, if it
is available. But when the sequences are random or drawn from an unknown
reference, clustering is the best option to tell real from spurious reads. One
such case is the use of random barcodes to track cells or transcripts
\citep{pmid18809713, pmid23953119}. Sequencing errors will create new spurious
barcodes that have to
be reverted to the original sequence by clustering. Hands on experience with real
datasets produced in our laboratory showed that this step can become limiting when
the number of unique sequences is high. In search for a solution to this problem, we
realized that heuristic approaches rely on assumptions that will perhaps not hold
as technologies evolve, we therefore set out to find an exact algorithm.

The first task of clustering is a matching phase where similar barcodes are paired,
similarly to linked nodes on a graph. The second task is the clusterng proper,
where communities in this graph are merged. We called our algorithm ``starcode'', in
reference to the star shape of the graph formed by the barcodes in the same
cluster. It occurred to us that the first version of our algorithm had the same
expected running time when barcodes were sorted or not, because it was not
exploiting data structuring of any kind. We reasoned that there should be a way to
use the prefix redundancy of sorted barcodes to gain speed. This is how we came up
with the poucet search algorithm at the heart of the matching step.

Here we describe the starcode algorithm and we benchmark it against existing
software that perform related tasks. We show that starcode is faster than the
alternatives, and that as an exact algorithm it makes no mistake on ideal datasets.
We show that starcode can also be used more generally to cluster k-mers from large
datasets from bacterial and even metazoan genomes.

\begin{methods}
\section{Methods}
\subsection{Inexact string matching using tries}
The matching method of starcode is a variation of the Needleman-Wunsch (NW)
algorithm \citep{pmid5420325}. In the original algorithm (Figure~\ref{fig:01}a),
the Levenshtein distance between two sequences is found by applying a recurrence
relation throughout a matrix of $mn$ terms, where $m$ and $n$ are the respective
sequence lengths. The complexity of this dynamic programming approach is $O(mn)$.

\begin{figure}[!tpb]%figure1
\centerline{\includegraphics{Fig1.pdf}}
\caption{Needleman-Wunsch (NW) alignment. \textbf{a}
  Alignment of GTTGCA and GATCCA using
the NW alignment matrix. The margins (purple) are initialized and the cells of
the matrix (yellow) are computed from left to right and from top to bottom by a
dynamic programming algorithm. The Levenshtein distance distance between the two
sequences is found in the bottom right cell. \textbf{b}
Lower complexity algorithm to
determine whether GTTGCA and GATCCA are 2-neighbors. The values in purple cells
are set during initialization. The dynamic programming algorithm proceeds as above,
with the difference that it is interrupted if the value of a diagonal cell (bold
borders) is larger than 2. The values in purple cells are not always the same as
their equivalent in a (purple arrow), but the values in the yellow cells are
nevertheless identical. The values of the white cells are never computed, which
contributes to making the algorithm less complex.}\label{fig:01}
\end{figure}

In many instances, the information of interest is whether the sequences are
$d$-neighbors (separated by a distance less than or equal to a threshold $d$).
In that case, the complexity can be reduced to $O(d\cdot \min(m,n))$ as described
below \citep{ukkonen}. Instead of initializing the margins and computing all the
terms, the matrix is initialized as shown on Figure~\ref{fig:01}b and only the
terms around
the diagonal are computed. If the computation of a diagonal term yields a value
greater than $d$, the process is halted (and the distance is known to be greater
than $d$). Otherwise, the distance is the bottom-right term, as in the NW
algorithm.

This method can be used for inexact matching of a sequence against a reference
set indexed as a trie, also called prefix trie \citep{ukkonen}. The terms of the
matrix are updated row-wise as a depth-first search traverses the trie from the
root, as illustrated in Figure~\ref{fig:02}. Every time a node is visited, the row
corresponding to its depth is recomputed. If a diagonal term exceeds the threshold
$d$, the depth-first search backtracks to the parent node because the Levenshtein
distance between the query and all the downstream sequences is greater than $d$,
so there are no hits to discover. When the process halts, every tail node
(corresponding to a sequence of the database) on the path of this search is a
$d$-neighbor of the query. This method is efficient because it eliminates large
areas of the search space, and because the NW alignment of the query with each
prefix of the database is computed only once.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics{Fig2.pdf}}
\caption{Trie query with NW alignment. Each sequence of the index is a path in
the trie. The query GTTGCA is written at the top of a NW matrix, which is
initialized as shown on Figure~\ref{fig:01}b. The trie is traversed by a
depth-first search (red path) from the root. At each depth, the node added to
the path is written on the left of the NW matrix and the row is computed.
Checkpoints from 1 to 4 (circled red numbers) show the state of the NW matrix
as the search proceeds. The node labelled 3 is a leaf and thus corresponds to a
2-neighbor of the query. The search path then backtracks to the node labelled 2
and the last rows of the NW are erased. The search path then goes to node labelled
4, in which case the newly computed diagonal cell exceeds the threshold (circled
in red). Even if this node has children, they are not visited (red crosses).}
\label{fig:02}
\end{figure}

\subsection{The poucet search algorithm}
In the case that many sequences are queried only once, this strategy can be
further improved. Notice that if two consecutive queries share a prefix of length
$k$, the sequences of computations up to row $k$ of the NW matrix will be exactly
the same in both queries. Since processing the queries in dictionary order
maximizes prefix sharing, the main idea of the algorithm is to store the
computational intermediates in the nodes of the trie. This way, the search can
start at depth $k$ whenever a query shares a prefix of length $k$ with the previous
one.

However, storing the rows of the NW matrix in the nodes, as suggested by the
approach described in the previous section is not optimal. At depth equal to the
length of the shared prefix, say $k$, the terms on the right of the diagonal were
computed in a previous query using characters that do not belong to the current
query, which means that they would have to be recomputed. Since they also depend
on the terms computed at lower depth, the search cannot restart at depth $k$. This
issue is solved by storing in each node the terms along an angle shape, looking
like a horizontally flipped L character as shown on Figure~\ref{fig:03}. This way, the
computation intermediates stored in a node of depth $k$ depend only on the first
$k$ characters of the query. This means that each query can restart at a depth
equal to the length of the shared prefix with the previous query, thereby
averting the need to recompute the first terms of the NW alignments.

In the fairy tale ``Le Petit Poucet'', the hero seeds white pebbles for his older
brothers to find their way home, which is reminiscent of the way queries pave
 the way for the next in this algorithm. We therefore called this search
algorithm ``poucet''.

\begin{figure}[!tpb]%figure3
\centerline{\includegraphics{Fig3.pdf}}
\caption{Poucet search algorithm. The algorithm proceeds with the same principles
as shown on Figure~\ref{fig:02} with the difference that the NW matrix is not
updated row-wise, but along horizontally flipped L shapes. As the depth-first
search proceeds, these values are stored in the nodes of the trie. Only the nodes
at the top contain initialized values; for the other nodes, the values at the
border are implicitly known to be 3. Since the values in the vertical part of
the flipped L are the same for every child of the same node, they are computed
only once (purple arrow). The values in the grey cells will be computed as the
search path (red) visits the node. Storing the intermediates in the nodes allows
the next query to restart at depth $k$ if it shares a common prefix of length
$k$ with the current query.}
\label{fig:03}
\end{figure}

\subsection{Seek and construct}
To reduce the size of the search space, we use a dynamic ``seek and construct''
approach whereby queries are processed meanwhile the trie is built. In other
words, each sequence is matched against the trie before it is inserted. An example
will explain why the trie does not need to contain all the sequences upon query.
Assume that two sequences A and B are $d$-neighbors. A is processed first. Since
B is not yet inserted, A yields not hit. It is then inserted in the trie. At the
time B is processed, A is a hit for the query and the match A-B is discovered.
This approach guarantees that every hit is discovered, while maintaining the
trie as ``thin'' as possible, which reduces search time.
The whole matching process is summarized in the pseudocode shown Figure~\ref{fig:04}

\subsection{Parallelization}
To parallelize the search, queries are separated into contiguous blocks after
sorting, and matching proceeds in two phases. During the first phase, each block
is used as input for the seek and construct process described above. In the second
phase, blocks are queried against the tries built in the first phase. If the
queries are segregated into $N$ blocks, the first phase consists of $N$ seek and
construct jobs, and the second consists of $N(N-1)/2$ query jobs. Since the jobs
show little dependence on each other, the matching algorithm can be efficiently
parallelized provided $N$ is larger than the number of independent threads.

\subsection{Clustering}
Starcode implements a multi-purpose clustering algorithm called ``sphere
clustering'', and a message passing algorithm \citep{mackay} tailored for the
task of clustering random barcodes. In sphere clustering, barcodes are sorted
by frequency and each barcode, starting from the most frequent, can claim its
$d$-neighbors that were not already claimed. This is the method used by
\cite{pmid23953119}.

In message passing clustering, read counts are distributed equally among the
closest neighbors of each barcode if they are at least 5 times more frequent.
The sequences with positive counts at the end of the process are considered
canonical, and their associated count is the estimated cluster size. A barcode
is assigned to a cluster if all its read count are eventually given to the
corresponding canonical barcode. The barcodes for which the read counts are
split between different canonical barcodes are not assigned to any cluster. The
reason for imposing a factor 5 or larger in order to transfer the read counts
is that barcodes with similar frequencies are not likely derived from each other
through sequencing errors. More likely they are either unrelated, or they both
derive from another more abundant barcode.

\subsection{Benchmark options}
The benchmark dataset is described in the Results section. Starcode was run with
options \texttt{-t 20}. CD-HIT version 4.6.1 was run with options
\texttt{-c 0.925 -T 20 -M 0}. USEARCH 32 bit version 7.0.1090 was run with options
\texttt{-threads 20 -cluster\_fast -id 0.925}. The Bowtie2 indexer was run as
\texttt{bowtie2-build --quiet}, and Bowtie2 version 2.1.0 was then run with
options \texttt{-f -a -p 20 --no-hd --very-sensitive}. The BWA indexer was run
as \texttt{bwa index -a is} and then BWA version 0.7.9a was run with options
\texttt{mem -t 20 -a -k1 -B0}. The GEM indexer was first run as
\texttt{gem-indexer -T 20}. GEM version 1.423 (beta) was then run with options
\texttt{-e3 -s -T 20 --granularity 100000}. The sequential algorithm of 
\cite{pmid23953119} was implemented in R and relied on the
\texttt{stringdist}
function of the \texttt{stringdist} package, used with options
\texttt{method='lv', maxDist=2}.

\end{methods}

\section{Results}
\subsection{Benchmark against sequence clustering algorithms}
Sequence clustering is routinely used to curate databases of non redundant
nucleotide or protein sequences. We benchmarked starcode against the two popular
sequence clustering algorithms CD-HIT \citep{pmid23060610} and USEARCH 
\citep{pmid20709691}.

We generated 1 million random 40-mers that were duplicated 47 times. For each
40-mer, we added 3 mutant sequences to the pool by sampling again 3 nucleotides
at random positions. The probability that such a mutant sequence is within a
distance 3 of another 40-mer is of the order of 10-12 and can be neglected.
Each cluster thus consists of 50 sequences, of which 3 have a distance at most
3 from the canonical representant.

Using 20 cores, starcode clustered the shuffled 50 million sequences in 4 minutes
without a single error (the output consisted of 1 million clusters of size 50).
We submitted the unique sequences of the dataset to CD-HIT and USEARCH. The
results are summarized in Figure~\ref{fig:05}. The running times of CD-HIT and USEARCH
were 59 and 64 minutes respectively, and they identified 2,362,352 and 2,739,954
clusters respectively. Both algorithms rely on heuristics, which is why the
number of clusters is off target. These results show that starcode outperforms
other clustering algorithms in speed and accuracy on this particular task.

\begin{figure}[!tpb]%figure5
\centerline{\includegraphics{Fig5.pdf}}
\caption{Benchmark of starcode against sequence clustering algorithms and short
read mappers. \textbf{a} Running time compared to CD-HIT and UCLUST on the
dataset described in the text. \textbf{b} Error rate compared to CD-HIT and UCLUST,
expressed as the percentage of clusters failing to be merged. \textbf{c} Running
time compared to Bowtie2, BWA and GEM on this dataset. The running time is
decomposed in extraction of unique sequences (purple), Burrows-Wheeler indexing
(yellow), and query proper (white). The extraction was performed with the Linux
sort command, so the time is the same in every case. \textbf{d} Error rate compared
to Bowtie2, BWA and GEM, expressed as the proportion of matches with a single hit.
This is an underestimate of the true error rate.}
\label{fig:05}
\end{figure}

\subsection{Benchmark against short read mappers}
A potential strategy to cluster short sequences is to use last generation short
read mappers to match a set of sequences against itself. Once matches are
available, several community detection algorithms can be used to identify the
clusters. We benchmarked starcode against the short read mappers Bowtie2
\citep{pmid22388286}, BWA \citep{pmid19451168} and GEM \citep{pmid23103880}.
Because short read
mappers do not perform clustering, we only evaluated their performance on the
matching problem.

For this test, we used again the dataset described above, in which every
sequence has at least 2 distinct matches at a distance 3. This means that we
can use matches with a single hit (themselves) as a lower bound on the inaccuracy.
We used unique sequences to build an index and queried the index with the same
sequences. Figure~\ref{fig:05} shows the running time of starcode along with the
decomposition of the running time for the three mappers. Note that time to
extract unique sequences is always the same because we used the Linux
\texttt{sort -u} command in every case. It appears that the running time of
starcode is about the same as the time required by BWA (the fastest aligner in
this test) to index and run the query proper. Figure~\ref{fig:05} shows the
lower bound on
the error rate, as estimated by the proportion of single hits. BWA was found to
be the most accurate aligner with a lower bound of 19\% errors. This measure
is an underestimate of the true error rate, but given that starcode makes no
mistake on this dataset, a more accurate measure is not necessary. In conclusion,
starcode runs slightly faster than short read mappers when all steps are taken
into account, and clearly outperforms them in precision on this particular problem.

\subsection{Clustering TRIP barcodes}
In the course of setting up the TRIP technology in our lab \citep{pmid23953119}, we
realized the need to develop efficient algorithms to cluster similar sequences.
Briefly, the principle of TRIP (Thousands of Reporters Integrated in Parallel)
is to tag reporter transcripts with random barcodes and measure the abundance
of barcodes in the RNA as a proxy for gene expression. There is no reference to
match aberrant barcodes against, because the tagging sequences are unknown.
Instead, barcodes are matched against each other and clustered by similarity to
infer canonical sequences.

We tested the efficiency of starcode on the TRIP dataset from
\cite{pmid23953119}. In the experiment labelled mPGKA, we identified
24.1 million barcode-containing reads out of 26.6 million (91\%), consisting
of approximately 223,000 and 220,000 unique barcode sequences for replicates
1 and 2 respectively. Following the authors, we kept only the barcodes with
at least 5 counts and performed clustering as described: ``First we sorted
barcodes according to their counts. Then, for each barcode (starting from the
most frequent one), we identified and removed all its mutant versions, defined
as barcodes within a Hamming distance of 2.'' This method is here on referred
to as the ``sequential algorithm''. When replacing the Hamming distance by the
Levenshtein distance, starcode produced exactly the same output as the sequential
algorithm. The running time of starcode averaged over the replicates was
2.90 seconds, instead of more than 3 minutes 40 seconds, which represents a
75 fold speedup.

The performance of the sequential algorithm relies on the arbitrary exclusion
of reads with less than 5 counts, the main purpose of which is to reduce the
computational burden. When all barcodes were kept, the average running time of
starcode was 10.27 seconds, versus $>$ 6 hours for the sequential algorithm.
Using the Hamming distance as the authors originally suggested decreased the
average running time of the sequential algorithm to 3 hours and 30 minutes (and
the output differed from that of starcode). Note that in all the cases
mentioned above, starcode was run with a single core to compare the algorithms
based on similar computer resources. In conclusion, the barcode clustering
problem can be simplified by various heuristics, but starcode brings down the
running time to nearly instantaneous, and thereby obviates the need for such
arbitrary heuristics.

\subsection{Identifying enriched sequence motifs}
Sequence motifs are thought to play an important role in DNA metabolism. Key
regulators such as transcription factors, nucleosomes and non coding RNAs have
sequence preferences targeting them to the sites where they act. Identifying
those sequences is a way to pinpoint the regulators and the mechanisms they
are involved in. However, the sequence motifs are not strictly identical
between sites, which is why they are better identified by inexact matching.
This problem becomes computationally difficult for long motifs (above 12-13
nucleotides) because of the combinatorial explosion. But as motifs become
longer, the problem of identifying abundant inexact matches becomes similar
to barcode clustering.

We set up a test based on the meningitis-causing agent
\textit{Neisseria meningitidis}. The genome of this bacterium is interspersed
with a frequent 12 bp sequence known as DNA uptake sequence \citep{pmid10673000}. We
extracted the 12-mers from both orientations of  the 2.19 Mb genome, yielding
4.39 million 12-mers, consisting of 2.77 million unique sequences. Using
starcode to cluster the 12-mers within a Levenshtein distance of 2 took less
than 3 minutes with 20 cores. We identified the known DNA uptake sequence of
Neisseria meningitidis (ATGCCGTCTGAA) as the most abundant 12-mer, comprising
1,466 exact matches and 4,625 inexact matches. This shows that starcode can
be used to identify biologically relevant motifs in bacterial genomes.

To test starcode on the scale of a metazoan genome, we used a Drosophila
dataset from \cite{pmid20888037}. That study describes two
signatures of chromatin proteins called Red and Yellow, that both correspond
to transcribed regions of the genome. The distinction is based on epigenomic
features only, which prompted us to identify sequence features that could
discriminate Red versus Yellow regions. To this end, we used starcode to
cluster 18-mers from both regions.

The total coverage of Red and Yellow chromatin in the Drosophila genome is
10.7 and 21.3 Mb respectively, corresponding to 21.5 million and 42.5 million
18-mers respectively. We used starcode to identify clusters of 18-mers within
a Levenshtein distance of 3 (the running time was approximately 35 minutes for
Red 18-mers and 70 minutes for Yellow 18-mers using 20 cores). The frequencies
of the most abundant 18-mers showed that Red chromatin has a lower sequence
complexity than Yellow chromatin (not shown). The most discriminating 18-mer
was (GA)$_9$, which showed a 4.8 fold enrichment in Red versus Yellow chromatin.
We counted 832 exact matches and 6,749 inexact matches (over 3,880 distinct
18-mers) in Red chromatin, showing that a larger portion of the signal comes
from inexact matches. Taken together, these results show that starcode can be
used to cluster short k-mers from metazoan-scale datasets.

\section{Discussion and conclusion}
Through the parallel poucet search algorithm, starcode implements an exact
sequence clustering algorithm that is even faster than widely used heuristics.
By design, starcode is tailored to process high throughput sequencing data on
multi-core platforms. Our benchmark shows that starcode is superior in speed
and accuracy to other methods on the problem of clustering short random
sequences. Part of the reason is that this task is stretching the tested
software far from their initial design. Sequence clustering algorithms are
meant to to cluster long biological sequences, while short read mappers are
meant to map reads on potentially large genomes. In this respect, starcode
fills a need arising from the development of barcoding technologies.

The speed of starcode also makes it proficient at more general clustering
problems, such as identifying enriched k-mers in genomes and in experimental
data. Here we have given two examples of such applications. In the first, we
recover a known enriched 12-mer in the genome of \textit{Neisseria meningitidis}.
In the second, we identify a sequence feature that discriminates genomic
regions found in two different chromatin types in Drosophila. GA repeats are
known to be bound in vivo by the protein GAF \citep{pmid12601174}, which suggests
that this transcription factor is instrumental in setting the distinction
between Red and Yellow chromatin.

The current version of starcode has been mildly optimized for memory
consumption. The memory footprint depends on the number of sequences to
cluster (because the sequences of the input set are loaded in memory as a
group of tries) and on the mean number of matches per sequence. Every match
has to be stored until the clustering phase, which can represent a heavy
load. As counterintuitive as it may seem, long queries will usually impose
a lower memory footprint because the matches between sequences are less
frequent.

The speed of starcode could benefit other applications, such as searching
shared subsequences in genome assembly. However, starcode supports only global
and not local alignment. A work around is to compare k-mers as we have done
in the examples above, but this approach is expected to be memory demanding.
Extending the starcode algorithm to support clustering of protein sequences
is straightforward but the speed is not expected to meet the level of the
algorithm shown here. On the one hand, there are more letters in the protein
alphabet, so the shared prefix between consecutive queries is expected to be
shorter on average, which defeats the poucet search. On the other hand, if
a threshold distance of the order of 1 per 10 nucleotides is sufficient to
identify similar DNA sequences, a distance of 1 per 10 residues is
unrealistically low for protein sequences, and this parameter critically
determines the speed of starcode. More generally, storing computational
intermediates for shared prefixes could find some applications in other
algorithms such as short read mappers. The idea of the poucet search seems
simple in retrospect, but it is a powerful way to tap into the data
structuration provided by string sorting.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
%     as it may ignore the figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Acknowledgement}
We would like to thank...

\paragraph{Funding\textcolon}
The research leading to these results has received funding from the
Government of Catalonia (Dept. of Economy and Knowledge) and the Spanish
Ministry of Economy and Competitiveness (Centro de Excelencia Severo Ochoa
2013-2017' (SEV-2012-0208). P.C. fellowship is partly financed by the
Spanish Ministry of Economy and Competitiveness (State Training Subprogram:
predoctoral fellowships for the training of PhD students (FPI) 2013).

\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
\bibliography{document}

%\begin{thebibliography}{}
%\end{thebibliography}
\end{document}

### 2014-06-19 doc/getref.py

@@ -1,0 +1,10 @@
+#!/usr/bin/env python
+# -*- coding:utf-8 -*-
+
+import sys
+import eUtils
+
+with open(sys.argv[1]) as f:
+   pmids = [line.rstrip() for line in f]
+
+sys.stdout.write(eUtils.eFetch_query(id=','.join(pmids)).read())

### 2014-06-19 doc/document.tex
Corrected typos in the manuscript.

@@
 \pubyear{2014}
+\usepackage{amssymb}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 
-\title[Exact sequence clustering]{Starcode: an exact algorithm for sequence clustering}
-\author[Valera Zorita \textit{et~al}]{Eduard Valera Zorita\,$^{1,2}$, Pol Cusc\'o\,$^{1,2}$ and Guillaume Filion\,$^{1,2}$\footnote{to whom correspondence should be addressed}}
-\address{$^{1}$Genome Architecture, Gene Regulation, Stem Cells and Cancer Programme, Centre for Genomic Regulation (CRG), Dr. Aiguader 88, 08003 Barcelona, Spain.\\
+\title[Exact sequence clustering with starcode]{Starcode: an exact algorithm for
+sequence clustering}
+\author[Valera Zorita \textit{et~al}]{Eduard Valera Zorita\,$^{1,2}$, Pol
+Cusc\'o\,$^{1,2}$ and Guillaume Filion\,$^{1,2}$\footnote{to whom correspondence
+should be addressed}}
+\address{$^{1}$Genome Architecture, Gene Regulation, Stem Cells and Cancer Programme,
+Centre for Genomic Regulation (CRG), Dr. Aiguader 88, 08003 Barcelona, Spain.\\
 $^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain.}
@@
 pairs of sequences lie within a given Levenshtein distance. The matched pairs
-are merged into clusters represented by a canonical sequence. The speed of
+are merged into clusters represented by a canonical sequence. The effiency of
 starcode is attributable to the poucet search, a novel  implementation of the
@@
 particular because of the high error rate of such technologies. For instance,
-the Illumina platform \citep{pmid16056220} shows a 1-2\% error rate consisting ofs
+the Illumina platform \citep{pmid16056220} shows a 1-2\% error rate consisting of
  substitutions near the 3' end of the read \citep{pmid18660515, pmid21576222}. The
-PacBio platform shows a 15\% error rate that mostly correspond to insertions
+PacBio platform shows a 15\% error rate that mostly corresponds to insertions
 and deletions \citep{pmid19023044}. As a consequence, the same sequence is often
@@
 Sequencing errors can be discovered by mapping the reads onto a reference, if it
-is available. But when the sequences are random or drawn from an unknown
+is available. When the sequences are random or drawn from an unknown
 reference, clustering is the best option to tell real from spurious reads. One
 such case is the use of random barcodes to track cells or transcripts
-\citep{pmid18809713, pmid23953119}. Sequencing errors will create new spurious
-barcodes that have to
-be reverted to the original sequence by clustering. Hands on experience with real
-datasets produced in our laboratory showed that this step can become limiting when
+\citep{pmid18809713, pmid23953119}. Sequencing errors will create erroneous
+barcodes that have to be reverted to the original sequence. Hands on experience
+with real datasets show that this step becomes limiting when
 the number of unique sequences is high. In search for a solution to this problem, we
-realized that heuristic approaches rely on assumptions that will perhaps not hold
-as technologies evolve, we therefore set out to find an exact algorithm.
+realized that heuristic approaches rely on assumptions that may not hold
+as technologies evolve. We therefore set out to find an exact algorithm.
 
-The first task of clustering is a matching phase where similar barcodes are paired,
-similarly to linked nodes on a graph. The second task is the clusterng proper,
+The first task of clustering is a matching phase where closely related barcodes are
+paired, similarly to linked nodes on a graph. The second task is the clustering proper,
 where communities in this graph are merged. We called our algorithm ``starcode'', in
 reference to the star shape of the graph formed by the barcodes in the same
-cluster. It occurred to us that the first version of our algorithm had the same
-expected running time when barcodes were sorted or not, because it was not
-exploiting data structuring of any kind. We reasoned that there should be a way to
-use the prefix redundancy of sorted barcodes to gain speed. This is how we came up
-with the poucet search algorithm at the heart of the matching step.
+cluster. The first version of the algorithm had the same performance regardless
+of the order in which the input sequences were processed, because it did not
+exploit data structuring of any kind. Exploiting the prefix redundancy of
+alphabetically sorted sequences allowed us to avoid unnecessary recomputations
+and gain speed. This is the rationale behind the poucet search algorithm at the
+heart of the matching step.
 
 Here we describe the starcode algorithm and we benchmark it against existing
-software that perform related tasks. We show that starcode is faster than the
-alternatives, and that as an exact algorithm it makes no mistake on ideal datasets.
-We show that starcode can also be used more generally to cluster k-mers from large
-datasets from bacterial and even metazoan genomes.
+software that perform related tasks. We show that starcode is both faster and
+more precise than the alternatives, achieving perfect clustering on ideal datasets.
+We further show that starcode is a good candidate to cluster k-mers from large
+datasets from bacterial or even metazoan genomes.
 
@@
 \subsection{Inexact string matching using tries}
-The matching method of starcode is a variation of the Needleman-Wunsch (NW)
-algorithm \citep{pmid5420325}. In the original algorithm (Figure~\ref{fig:01}a),
-the Levenshtein distance between two sequences is found by applying a recurrence
-relation throughout a matrix of $mn$ terms, where $m$ and $n$ are the respective
-sequence lengths. The complexity of this dynamic programming approach is $O(mn)$.
+The matching method of starcode is based on a variation of the Needleman-Wunsch
+(NW) algorithm \citep{pmid5420325}. In the original algorithm
+(Figure~\ref{fig:01}a), the Levenshtein distance between two sequences is found
+by applying a recurrence relation throughout a matrix of $mn$ terms, where $m$
+and $n$ are the respective sequence lengths. The complexity of this dynamic
+programming approach is $O(mn)$.
 
@@
 
-In many instances, the information of interest is whether the sequences are
-$d$-neighbors (separated by a distance less than or equal to a threshold $d$).
-In that case, the complexity can be reduced to $O(d\cdot \min(m,n))$ as described
-below \citep{ukkonen}. Instead of initializing the margins and computing all the
+In many instances, the only information of interest is to find out whether the
+sequences are $d$-neighbors (separated by a distance less than or equal to a
+threshold $d$).  In that case, the complexity then reduces to
+$O(d \min(m,n))$ as described below \citep{ukkonen}. Instead
+of initializing the margins of the matrix and computing all the
 terms, the matrix is initialized as shown on Figure~\ref{fig:01}b and only the
@@
 the diagonal are computed. If the computation of a diagonal term yields a value
-greater than $d$, the process is halted (and the distance is known to be greater
-than $d$). Otherwise, the distance is the bottom-right term, as in the NW
-algorithm.
+greater than $d$, the distance is known to be greater than $d$ and the process
+is halted. Otherwise, the bottom-right term indicates the actual distance
+between the sequences, as in the NW algorithm.
 
-This method can be used for inexact matching of a sequence against a reference
-set indexed as a trie, also called prefix trie \citep{ukkonen}. The terms of the
+This method can be used for inexact matching of sequences against a reference
+set indexed as prefix tree, also known as a trie \citep{ukkonen}. The terms of the
 matrix are updated row-wise as a depth-first search traverses the trie from the
 root, as illustrated in Figure~\ref{fig:02}. Every time a node is visited, the row
-corresponding to its depth is recomputed. If a diagonal term exceeds the threshold
-$d$, the depth-first search backtracks to the parent node because the Levenshtein
-distance between the query and all the downstream sequences is greater than $d$,
-so there are no hits to discover. When the process halts, every tail node
+corresponding to its depth is recomputed. If the threshold value $d$ is exceeded
+for a diagonal term, the Levenshtein distance for all the downstream sequences is
+also necessarily greater than $d$. Therefore, no more hits are to be discovered
+in this path and the depth-first search backtracks to the parent node. When the
+process halts, every tail node
 (corresponding to a sequence of the database) on the path of this search is a
@@
 \subsection{The poucet search algorithm}
-In the case that many sequences are queried only once, this strategy can be
+This strategy can be
 further improved. Notice that if two consecutive queries share a prefix of length
-$k$, the sequences of computations up to row $k$ of the NW matrix will be exactly
-the same in both queries. Since processing the queries in dictionary order
-maximizes prefix sharing, the main idea of the algorithm is to store the
-computational intermediates in the nodes of the trie. This way, the search can
-start at depth $k$ whenever a query shares a prefix of length $k$ with the previous
+$k$, the succession of computations up to the $k$-th row of the NW matrix will be exactly
+the same in both queries. To take advantage of this property, the input sequences are
+sorted alphabetically in order to maximize the prefix sharing. The main idea of the
+algorithm is to store the computational intermediates in the nodes of the trie and
+use them later to resume the computation in the next query. This way, the search can
+start at depth $k$ whenever a query shares a prefix of such length with the previous
 one.
@@
 length of the shared prefix, say $k$, the terms on the right of the diagonal were
-computed in a previous query using characters that do not belong to the current
-query, which means that they would have to be recomputed. Since they also depend
+computed using characters that belong to the previous query. Since they also depend
 on the terms computed at lower depth, the search cannot restart at depth $k$. This
-issue is solved by storing in each node the terms along an angle shape, looking
+issue is solved by storing in each node a combination of row and column terms that
+form an angle shape, looking
 like a horizontally flipped L character as shown on Figure~\ref{fig:03}. This way, the
-computation intermediates stored in a node of depth $k$ depend only on the first
-$k$ characters of the query. This means that each query can restart at a depth
-equal to the length of the shared prefix with the previous query, thereby
+computation intermediates stored in a node of depth $k$ depend only on the previous
+characters of the query. Using this technique, the search can effectively restart
+at a depth equal to the length of the shared prefix with the previous query, thereby
 averting the need to recompute the first terms of the NW alignments.
@@
 approach whereby queries are processed meanwhile the trie is built. In other
-words, each sequence is matched against the trie before it is inserted. An example
-will explain why the trie does not need to contain all the sequences upon query.
-Assume that two sequences A and B are $d$-neighbors. A is processed first. Since
-B is not yet inserted, A yields not hit. It is then inserted in the trie. At the
+words, each sequence is matched against the trie before it is inserted. To
+illustrate why the trie does not need to contain all the sequences upon query,
+assume that two sequences A and B are $d$-neighbors. A is processed first. Since
+B is not yet inserted, A yields no hit. It is then inserted in the trie. At the
 time B is processed, A is a hit for the query and the match A-B is discovered.
 This approach guarantees that every hit is discovered, while maintaining the
-trie as ``thin'' as possible, which reduces search time.
-The whole matching process is summarized in the pseudocode shown Figure~\ref{fig:04}
+trie as ``thin'' as possible, thereby reducing the search time.
+%The whole matching process is summarized in the pseudocode shown Figure~\ref{fig:04}
+The whole matching process is summarized in the pseudocode shown in
+Algorithms~\ref{alg:starcode} and \ref{alg:poucet}.
+
+\begin{algorithm}
+  \caption{Starcode algorithm}
+  \label{alg:starcode}
+  \begin{algorithmic}[1]
+    \State Define: $\tau$
+    \State \textbf{Initialize variables:} $seed$, $start = 0$,
+    $height$, $seq$, $trie$
+    \State \textbf{Initialize containers:} $hits$, $pebbles$
+    \State \textsc{read} sequence file
+    \State $height \gets$ \textsc{determine} maximum sequence length
+    \State \textsc{pad} sequences up to $height$
+    \State \textsc{sort} sequences alphabetically
+    \State $trie \gets$ \textsc{create} an empty trie of height $height$
+    \State \textsc{insert} root node of $trie$ in $pebbles$ at depth 0
+    \ForAll{sequences}
+    \State $seq \gets$ \textsc{get} next sequence
+    \State $seed \gets$ \textsc{length} of shared prefix
+    between current and next $seq$
+    \State \textsc{clear} $hits$
+    \State \textsc{clear} $pebbles$ at depth $>start$
+    \ForAll{$pebbles$ at depth $start$}
+    \State $node \gets$ \textsc{get} next node from $pebbles$
+    \State \textbf{call} \textsc{poucet}($node$, $seed$, $hits$)
+    \EndFor
+    \State \textsc{insert} $seq$ path in $trie$
+    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
+    \State $start \gets seed$
+    \EndFor
+  \end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+  \caption{Poucet search algorithm}
+  \label{alg:poucet}
+  \begin{algorithmic}[1]
+    \Procedure{poucet}{$node$, $seed$, $hits$}:
+    \State \textsc{compute} $node$-specific column values following NW algorithm
+    \ForAll{$child$ nodes in $node$}
+    \State \textsc{compute} $child$-specific row values following NW algorithm
+    \State \textsc{compute} center value using row and column
+    values
+    \If{center value $>\tau$} \Comment{Mismatches exceeded.}
+    \State \textbf{continue} with next $child$
+    \EndIf
+    \If{$node$ depth = $height$} \Comment{Hit found.}
+    \State \textsc{save} $node$ sequence in $hits$
+    \State \textbf{continue} with next $child$
+    \EndIf
+    \If{$node$ depth $\leq seed$}
+    \State \textsc{save} $node$ in pebbles at current depth
+    \EndIf
+    \State \textbf{call} \textsc{poucet}($child$, $seed$, $hits$)
+    \EndFor
+    \EndProcedure
+  \end{algorithmic}
+\end{algorithm}
 
@@
 To parallelize the search, queries are separated into contiguous blocks after
-sorting, and matching proceeds in two phases. During the first phase, each block
-is used as input for the seek and construct process described above. In the second
-phase, blocks are queried against the tries built in the first phase. If the
+sorting. The matching algorithm proceeds in two phases. During the first phase,
+an independent trie is created for each block and filled with its sequence using
+the seek and construct process described above. In the second
+phase, blocks are only queried against but not inserted into the tries built in
+the first phase. If the
 queries are segregated into $N$ blocks, the first phase consists of $N$ seek and
@@
 task of clustering random barcodes. In sphere clustering, barcodes are sorted
-by frequency and each barcode, starting from the most frequent, can claim its
-$d$-neighbors that were not already claimed. This is the method used by
-\cite{pmid23953119}.
+by frequency of occurrence and each barcode, starting from the most frequent,
+can claim its $d$-neighbors that were not already claimed. This is the method
+used by \cite{pmid23953119}.
 
@@
 closest neighbors of each barcode if they are at least 5 times more frequent.
-The sequences with positive counts at the end of the process are considered
+The remaining sequences at the end of the process are considered
 canonical, and their associated count is the estimated cluster size. A barcode
@@
 \subsection{Benchmark options}
-The benchmark dataset is described in the Results section. Starcode was run with
-options \texttt{-t 20}. CD-HIT version 4.6.1 was run with options
-\texttt{-c 0.925 -T 20 -M 0}. USEARCH 32 bit version 7.0.1090 was run with options
-\texttt{-threads 20 -cluster\_fast -id 0.925}. The Bowtie2 indexer was run as
-\texttt{bowtie2-build --quiet}, and Bowtie2 version 2.1.0 was then run with
-options \texttt{-f -a -p 20 --no-hd --very-sensitive}. The BWA indexer was run
-as \texttt{bwa index -a is} and then BWA version 0.7.9a was run with options
-\texttt{mem -t 20 -a -k1 -B0}. The GEM indexer was first run as
-\texttt{gem-indexer -T 20}. GEM version 1.423 (beta) was then run with options
-\texttt{-e3 -s -T 20 --granularity 100000}. The sequential algorithm of 
-\cite{pmid23953119} was implemented in R and relied on the
-\texttt{stringdist}
-function of the \texttt{stringdist} package, used with options
-\texttt{method='lv', maxDist=2}.
+
+The benchmark dataset is described in the Results section. All software were
+set to run with 20 parallel threads on a 20-core Intel Xeon processor. The
+execution options for each software are summarized in
+Table~\ref{tab:soft-options}.
+
+\begin{table}[h]
+\centering
+\caption{Software execution parameters}
+\resizebox{\columnwidth}{!}{
+\begin{tabular}{|l|l|}
+    \hline
+    starcode & \texttt{-t 20}\\ \hline
+    CD-HIT 4.6.1 & \texttt{-c 0.925 -T 20 -M 0}\\ \hline
+    USEARCH 7.0.1090  & \texttt{-threads 20 -cluster\_fast -id 0.925}\\ \hline
+        Bowtie2 indexer & \texttt{bowtie2-build --quiet}\\ \hline
+    Bowtie2 2.1.0 & \texttt{-f -a -p 20 --no-hd --very-sensitive}\\ \hline
+        BWA indexer & \texttt{bwa index -a is}\\ \hline
+    BWA 0.7.9a & \texttt{mem -t 20 -a -k1 -B0} \\ \hline
+    GEM indexer & \texttt{gem-indexer -T 20} \\ \hline
+    GEM 1.423 & \texttt{-e3 -s -T 20 --granularity 100000}\\ \hline
+    \emph{sequential R} & \texttt{method='lv', maxDist=2}\\ \hline
+\end{tabular}
+}
+\label{tab:soft-options}
+\end{table}
+
+Denoted as \emph{sequential R} is the sequential algorithm of
+\cite{pmid23953119} which implemented in R and relied on the \texttt{stringdist}
+function of the \texttt{stringdist} package.
 
@@
 We generated 1 million random 40-mers that were duplicated 47 times. For each
-40-mer, we added 3 mutant sequences to the pool by sampling again 3 nucleotides
+40-mer, we inserted 3 addional mutant sequences to the pool by sampling again
+3 nucleotides
 at random positions. The probability that such a mutant sequence is within a
-distance 3 of another 40-mer is of the order of 10-12 and can be neglected.
+distance 3 of any other 40-mer is of the order of 10-12 and can be neglected.
 Each cluster thus consists of 50 sequences, of which 3 have a distance at most
@@
 We submitted the unique sequences of the dataset to CD-HIT and USEARCH. The
-results are summarized in Figure~\ref{fig:05}. The running times of CD-HIT and USEARCH
+results are summarized in Figure~\ref{fig:04}. The running times of CD-HIT and USEARCH
 were 59 and 64 minutes respectively, and they identified 2,362,352 and 2,739,954
-clusters respectively. Both algorithms rely on heuristics, which is why the
-number of clusters is off target. These results show that starcode outperforms
-other clustering algorithms in speed and accuracy on this particular task.
+clusters respectively. Both algorithms rely on heuristics, fact that explains why the
+number of clusters is off target. These results clearly demonstrate that starcode
+outperforms other clustering algorithms in speed and accuracy on this particular task.
 
 \begin{figure}[!tpb]%figure5
-\centerline{\includegraphics{Fig5.pdf}}
+\centerline{\includegraphics{Fig4.pdf}}
 \caption{Benchmark of starcode against sequence clustering algorithms and short
@@
 This is an underestimate of the true error rate.}
-\label{fig:05}
+\label{fig:04}
 \end{figure}
@@
 
-For this test, we used again the dataset described above, in which every
-sequence has at least 2 distinct matches at a distance 3. This means that we
+For this test, we used the same dataset described above, in which every
+sequence has at least 2 distinct matches at a distance 3 (mutant sequences
+from the same cluster may be at a distance up to 6). This means that we
 can use matches with a single hit (themselves) as a lower bound on the inaccuracy.
@@
 be the most accurate aligner with a lower bound of 19\% errors. This measure
-is an underestimate of the true error rate, but given that starcode makes no
+is indeed an underestimate of the true error rate, but given that starcode makes no
 mistake on this dataset, a more accurate measure is not necessary. In conclusion,
@@
 \cite{pmid23953119}. In the experiment labelled mPGKA, we identified
-24.1 million barcode-containing reads out of 26.6 million (91\%), consisting
-of approximately 223,000 and 220,000 unique barcode sequences for replicates
+24.1 million (91\%) barcode-containing reads out of 26.6 million, consisting
+of approximately 223,000 and 220,000 unique barcode sequences for PCR replicates
 1 and 2 respectively. Following the authors, we kept only the barcodes with
@@
 algorithm. The running time of starcode averaged over the replicates was
-2.90 seconds, instead of more than 3 minutes 40 seconds, which represents a
-75 fold speedup.
+2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which represents a
+75-fold speedup.
 
@@
 computational burden. When all barcodes were kept, the average running time of
-starcode was 10.27 seconds, versus $>$ 6 hours for the sequential algorithm.
-Using the Hamming distance as the authors originally suggested decreased the
-average running time of the sequential algorithm to 3 hours and 30 minutes (and
+starcode was 10.27 seconds, \textit{versus} $>$ 6 hours for the sequential
+algorithm. Using the Hamming distance, as the authors originally suggested,
+decreased the average running time of the sequential algorithm to about 3 hours
+and 30 minutes (and
 the output differed from that of starcode). Note that in all the cases
@@
 Sequence motifs are thought to play an important role in DNA metabolism. Key
-regulators such as transcription factors, nucleosomes and non coding RNAs have
+regulators, such as transcription factors, nucleosomes and non coding RNAs have
 sequence preferences targeting them to the sites where they act. Identifying
@@
 are involved in. However, the sequence motifs are not strictly identical
-between sites, which is why they are better identified by inexact matching.
+at different sites, hence they are better identified by inexact matching.
 This problem becomes computationally difficult for long motifs (above 12-13
-nucleotides) because of the combinatorial explosion. But as motifs become
+nucleotides) because of the combinatorial scaling. But as motifs become
 longer, the problem of identifying abundant inexact matches becomes similar
@@
 extracted the 12-mers from both orientations of  the 2.19 Mb genome, yielding
-4.39 million 12-mers, consisting of 2.77 million unique sequences. Using
-starcode to cluster the 12-mers within a Levenshtein distance of 2 took less
+4.39 million 12-mers, consisting of 2.77 million unique sequences. Clustering
+the 12-mers with starcode within a Levenshtein distance of 2 took less
 than 3 minutes with 20 cores. We identified the known DNA uptake sequence of
 Neisseria meningitidis (ATGCCGTCTGAA) as the most abundant 12-mer, comprising
-1,466 exact matches and 4,625 inexact matches. This shows that starcode can
-be used to identify biologically relevant motifs in bacterial genomes.
+1,466 exact matches and 4,625 inexact matches. This result testifies to the
+fact that starcode can be used to identify biologically relevant motifs in
+bacterial genomes.
 
 To test starcode on the scale of a metazoan genome, we used a Drosophila
-dataset from \cite{pmid20888037}. That study describes two
-signatures of chromatin proteins called Red and Yellow, that both correspond
+dataset from \cite{pmid20888037}. The study describes two
+signatures of chromatin proteins called Red and Yellow, both corresponding
 to transcribed regions of the genome. The distinction is based on epigenomic
 features only, which prompted us to identify sequence features that could
-discriminate Red versus Yellow regions. To this end, we used starcode to
+discriminate Red \textit{versus} Yellow regions. To this end, we used starcode to
 cluster 18-mers from both regions.
@@
 The total coverage of Red and Yellow chromatin in the Drosophila genome is
-10.7 and 21.3 Mb respectively, corresponding to 21.5 million and 42.5 million
-18-mers respectively. We used starcode to identify clusters of 18-mers within
+10.7 Mb and 21.3 Mb respectively, corresponding to 21.5 million and 42.5 million
+18-mers, respectively. We used starcode to identify clusters of 18-mers within
 a Levenshtein distance of 3 (the running time was approximately 35 minutes for
-Red 18-mers and 70 minutes for Yellow 18-mers using 20 cores). The frequencies
+Red 18-mers and 70 minutes for Yellow 18-mers using 20 threads). The frequencies
 of the most abundant 18-mers showed that Red chromatin has a lower sequence
 complexity than Yellow chromatin (not shown). The most discriminating 18-mer
-was (GA)$_9$, which showed a 4.8 fold enrichment in Red versus Yellow chromatin.
-We counted 832 exact matches and 6,749 inexact matches (over 3,880 distinct
-18-mers) in Red chromatin, showing that a larger portion of the signal comes
+was (GA)$_9$, which showed a 4.8 fold enrichment in Red \textit{versus} Yellow
+chromatin.
+We counted 832 exact matches and 6,749 inexact matches over 3,880 distinct
+18-mers in Red chromatin, showing that a larger portion of the signal comes
 from inexact matches. Taken together, these results show that starcode can be
@@
 multi-core platforms. Our benchmark shows that starcode is superior in speed
-and accuracy to other methods on the problem of clustering short random
-sequences. Part of the reason is that this task is stretching the tested
+and accuracy compared to other methods on the problem of clustering short random
+sequences. Part of the reason is that such task is stretching the tested
 software far from their initial design. Sequence clustering algorithms are
@@
 group of tries) and on the mean number of matches per sequence. Every match
-has to be stored until the clustering phase, which can represent a heavy
-load. As counterintuitive as it may seem, long queries will usually impose
-a lower memory footprint because the matches between sequences are less
+has to be stored until the clustering phase and may represent a heavy
+memory overhead. As counterintuitive as it may seem, long queries will usually
+impose a lower memory footprint because the matches between sequences are less
 frequent.
 
-The speed of starcode could benefit other applications, such as searching
+The speed of starcode could also benefit other applications, such as searching
 shared subsequences in genome assembly. However, starcode supports only global
-and not local alignment. A work around is to compare k-mers as we have done
+and not local alignment. A workaround is to compare k-mers as we have done
 in the examples above, but this approach is expected to be memory demanding.
-Extending the starcode algorithm to support clustering of protein sequences
-is straightforward but the speed is not expected to meet the level of the
-algorithm shown here. On the one hand, there are more letters in the protein
-alphabet, so the shared prefix between consecutive queries is expected to be
+Even though extending the starcode algorithm to support clustering of protein
+sequences is straightforward, it is not expected to meet the performance level
+of the presented algorithm. On the one hand, there are more letters in the protein
+alphabet, so the shared prefix between consecutive queries is inherently
 shorter on average, which defeats the poucet search. On the other hand, if

### 2014-06-19 doc/document.tex
Corrected some more typos in the manuscript.

@@
 sorting. The matching algorithm proceeds in two phases. During the first phase,
-an independent trie is created for each block and filled with its sequence using
-the seek and construct process described above. In the second
-phase, blocks are only queried against but not inserted into the tries built in
-the first phase. If the
+an independent trie is created and filled with the sequences
+of one block using the seek and construct process described above. In the second
+phase, blocks of sequences are queried against but not inserted into the tries
+built in the first phase. If the
 queries are segregated into $N$ blocks, the first phase consists of $N$ seek and
@@
 canonical, and their associated count is the estimated cluster size. A barcode
-is assigned to a cluster if all its read count are eventually given to the
+is assigned to a cluster if all its read counts are eventually given to the
 corresponding canonical barcode. The barcodes for which the read counts are
@@
 Denoted as \emph{sequential R} is the sequential algorithm of
-\cite{pmid23953119} which implemented in R and relied on the \texttt{stringdist}
+\cite{pmid23953119} which is implemented in R and relied on the \texttt{stringdist}
 function of the \texttt{stringdist} package.
@@
 at random positions. The probability that such a mutant sequence is within a
-distance 3 of any other 40-mer is of the order of 10-12 and can be neglected.
+distance 3 of any other 40-mer is of the order of $10^{-12}$ and can be neglected.
 Each cluster thus consists of 50 sequences, of which 3 have a distance at most
@@
 We submitted the unique sequences of the dataset to CD-HIT and USEARCH. The
-results are summarized in Figure~\ref{fig:04}. The running times of CD-HIT and USEARCH
+results are summarized in Figure~\ref{fig:04}a and \ref{fig:04}b. The running
+times of CD-HIT and USEARCH
 were 59 and 64 minutes respectively, and they identified 2,362,352 and 2,739,954
@@
 We used unique sequences to build an index and queried the index with the same
-sequences. Figure~\ref{fig:05} shows the running time of starcode along with the
-decomposition of the running time for the three mappers. Note that time to
+sequences. Figure~\ref{fig:04}c shows the running time of starcode along with the
+decomposition of the running time for the three mappers. Note that the time to
 extract unique sequences is always the same because we used the Linux
@@
 starcode is about the same as the time required by BWA (the fastest aligner in
-this test) to index and run the query proper. Figure~\ref{fig:05} shows the
+this test) to index and run the query proper. Figure~\ref{fig:04}d shows the
 lower bound on
@@
 \subsection{Clustering TRIP barcodes}
-In the course of setting up the TRIP technology in our lab \citep{pmid23953119}, we
+In the course of setting up the TRIP technology in our laboratory
+\citep{pmid23953119}, we
 realized the need to develop efficient algorithms to cluster similar sequences.
@@
 of approximately 223,000 and 220,000 unique barcode sequences for PCR replicates
-1 and 2 respectively. Following the authors, we kept only the barcodes with
+1 and 2, respectively. Following the authors, we kept only the barcodes with
 at least 5 counts and performed clustering as described: ``First we sorted
@@
 based on similar computer resources. In conclusion, the barcode clustering
-problem can be simplified by various heuristics, but starcode brings down the
+problem can be simplified by various tricks, but starcode brings down the
 running time to nearly instantaneous, and thereby obviates the need for such
@@
 than 3 minutes with 20 cores. We identified the known DNA uptake sequence of
-Neisseria meningitidis (ATGCCGTCTGAA) as the most abundant 12-mer, comprising
+\textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant 12-mer, comprising
 1,466 exact matches and 4,625 inexact matches. This result testifies to the
@@
 
-To test starcode on the scale of a metazoan genome, we used a Drosophila
+To test starcode on the scale of a metazoan genome, we used a \textit{Drosophila}
 dataset from \cite{pmid20888037}. The study describes two
@@
 
-The total coverage of Red and Yellow chromatin in the Drosophila genome is
+The total coverage of Red and Yellow chromatin in the \textit{Drosophila} genome is
 10.7 Mb and 21.3 Mb respectively, corresponding to 21.5 million and 42.5 million
@@
 a Levenshtein distance of 3 (the running time was approximately 35 minutes for
-Red 18-mers and 70 minutes for Yellow 18-mers using 20 threads). The frequencies
-of the most abundant 18-mers showed that Red chromatin has a lower sequence
-complexity than Yellow chromatin (not shown). The most discriminating 18-mer
-was (GA)$_9$, which showed a 4.8 fold enrichment in Red \textit{versus} Yellow
+Red and 70 minutes for Yellow chromatin using 20 threads). The most discriminating
+18-mer was (GA)$_9$, which showed a 4.8 fold enrichment in Red \textit{versus} Yellow
 chromatin.
@@
 software far from their initial design. Sequence clustering algorithms are
-meant to to cluster long biological sequences, while short read mappers are
+meant to cluster long biological sequences, while short read mappers are
 meant to map reads on potentially large genomes. In this respect, starcode
@@
 In the second, we identify a sequence feature that discriminates genomic
-regions found in two different chromatin types in Drosophila. GA repeats are
-known to be bound in vivo by the protein GAF \citep{pmid12601174}, which suggests
+regions found in two different chromatin types in \textit{Drosophila}. GA
+repeats are known to be bound \textit{in vivo} by the protein GAF
+\citep{pmid12601174}, which suggests
 that this transcription factor is instrumental in setting the distinction
@@
 group of tries) and on the mean number of matches per sequence. Every match
-has to be stored until the clustering phase and may represent a heavy
+has to be stored until the clustering phase, which may represent a large
 memory overhead. As counterintuitive as it may seem, long queries will usually
@@
 The speed of starcode could also benefit other applications, such as searching
-shared subsequences in genome assembly. However, starcode supports only global
+shared subsequences during genome assembly. However, starcode supports only global
 and not local alignment. A workaround is to compare k-mers as we have done

### 2014-06-19 doc/document.tex
Added more information about the NW algorithm in caption of Fig 1.

@@
 the NW alignment matrix. The margins (purple) are initialized and the cells of
-the matrix (yellow) are computed from left to right and from top to bottom by a
-dynamic programming algorithm. The Levenshtein distance distance between the two
-sequences is found in the bottom right cell. \textbf{b}
+the matrix (yellow) are computed from left to right and from top to bottom by
+the following
+dynamic programming algorithm. The value in each cell is computed from the values
+of the neighboring top, left and top-left cells. Denoting $a$, $b$ and $c$ these
+respective terms, the value in the cell of coordinates $(i,j)$ is computed as
+$\min(a+1, b+1, c)$ if the $i$-th letter from the first sequence is identical to
+the $j$-th letter from the second, or $\min(a+1, b+1, c+1)$ if the letters are
+different. The Levenshtein distance
+between the two sequences is found in the bottom right cell. \textbf{b}
 Lower complexity algorithm to

### 2014-06-28 test/testset.c
More extensive test for 'insert_string_wo_malloc()'.

@@ -233,7 +233,9 @@ test_base_1
    const char cache[17] = {8,7,6,5,4,3,2,1,0,1,2,3,4,5,6,7,8};
    for (char maxtau = 0 ; maxtau < 9 ; maxtau++) {
       node_t *node = new_trienode(maxtau);
+      // Check that creation succeeded.
       g_assert(node != NULL);
+      // Check initialization.
       g_assert(node->cache != NULL);
       g_assert_cmpint(node->path, ==, 0);
       for (int i = 0 ; i < 6 ; i++) {
@@ -242,6 +244,7 @@ test_base_1
       for (int i = 0 ; i < 2*maxtau + 1 ; i++) {
          g_assert(node->cache[i] == cache[i+(8-maxtau)]);
       }
+      // Check no memory corruption.
       free(node);
    }
    return;
@@ -263,9 +266,9 @@ test_base_2
       for (int j = 0 ; j < 6 ; j++) {
          g_assert(node->child[j] == NULL);
       }
-      const char cache[] = {3,2,1,0,1,2,3};
-      for (int j = 0 ; j < 7 ; j++) {
-         g_assert(node->cache[j] == cache[j]);
+      const char cache[17] = {8,7,6,5,4,3,2,1,0,1,2,3,4,5,6,7,8};
+      for (int i = 0 ; i < 2*maxtau + 1 ; i++) {
+         g_assert(node->cache[i] == cache[i+(8-maxtau)]);
       }
       g_assert_cmpint(node->path, ==, i);
       g_assert(root->child[i] == node);
@@ -298,9 +301,9 @@ test_base_3
       for (int j = 0 ; j < 6 ; j++) {
          g_assert(node->child[j] == NULL);
       }
-      const char cache[] = {3,2,1,0,1,2,3};
-      for (int j = 0 ; j < 7 ; j++) {
-         g_assert(node->cache[j] == cache[j]);
+      const char cache[17] = {8,7,6,5,4,3,2,1,0,1,2,3,4,5,6,7,8};
+      for (int i = 0 ; i < 2*maxtau + 1 ; i++) {
+         g_assert(node->cache[i] == cache[i+(8-maxtau)]);
       }
       g_assert_cmpint(node->path, ==, i);
       g_assert(root->child[i] == node);
@@ -397,7 +400,6 @@ test_base_5
[…]

### 2014-06-30 doc/pseudocode.tex
Updated pseudocode and Lookup table section.

@@
 
+\section{lookup search}
+The lookup table search algorithm is based on the fact that if a
+sequence is divided in $\tau+1$ non-overlapping subsequences and
+mutated with $\tau$ mismatches, at least one subsequence must remain
+unchanged for any mismatch combination. In other words,
+the distance between two sequences is lower bounded by the number of
+different subsequences between them. One can take advantage of this
+property to build a table containing all the $k$-mers that have been
+previously inserted in the trie. When a new sequence is to be queried,
+it is first divided in $\tau+1$ subsequences and all them are queried
+against the lookup table. If none of the subsequences are found, the
+minimum distance between the query sequence and all the sequences in
+the trie is at least $\tau+1$ and the trie-search can therefore be
+omitted. This technique is especially useful for long input sequences,
+in which the length of the subsequences ($k$) is big and the
+probability of finding a coinciding succession of $k$ nucleotides is
+only significant when the two sequences do actually belong to the same
+cluster. For the rest of the cases, the negative match is detected
+only at the computational cost of several table lookups compared to the
+enormous cost of measuring the Levenshtein distance between many long
+sequences. In our implementation, the length $k$ of the subsequences
+is computed in a way that $\tau$ nucleotides are left out. Hence, the
+original sequence is divided into $\tau+2$ subsequences, the last of
+them containing only $\tau$ nucleotides that will be later used to
+quickly detect up to $\tau$ insertions or deletions. The prospective
+length of the subsequences is
+\begin{equation}
+  k = \frac{L}{\tau+1}
+\end{equation}
+where $L$ is the total length of the sequence. The number of remaining
+nucleotides to fill the last subsequence is found as
+\begin{equation}
+  R = \tau - \big[ L\mod{(\tau+1)} \big]
+\end{equation}
+and finally the length of each subsequence is computed as
+\begin{equation}
+  k_i =
+  \left\{
+  \begin{array}{cl}
+    k-1 \qquad & i = 1\ldots R \\
+    k \qquad & i = R+1\ldots \tau+1
+    \end{array}
+    \right.
+\end{equation}
+The subsequences are inserted in the lookup table after splitting
+the original sequence in chunks of length $k_i$. Note that, since the
+coincidence with the original sequence must be exact both in position
+and content, the rate of false positives returned by the lookup method
+is greatly reduced if an independent table is used for each
+subsequence. In order to allow for insertions and deletions as candidate
+mismatches, the lookup search must also query for shifted versions of
+the $k$-mers. Assuming the worst case scenario in which only one of
+the subsequences is found and the mismatches are evenly distributed in
+the query sequence, the maximum number of cummulative
+insertions/deletions at the $i$-th subsequence is exactly $i-1$ ($i =
+1\ldots \tau+1$). Hence, the $i$-th subsequence has to be shifted and
+queried against the lookup table $2(i-1)+1$ times to cover all the
+range from $i-1$ insertions to $i-1$ deletions. The ($\tau+1$)-th
+subsequence will therefore be queried $2\tau+1$ times, corresponding
+to shifts from $\tau$ nucleotides on the right to $\tau$ nucleotides
+on the left, the latter requiring the $\tau$ extra nucleotides left
+during the computation of the subsequence lengths.
+
 \begin{algorithm}
@@
   \begin{algorithmic}[1]
-    \State Define: $\tau$
-    \State \textbf{Initialize variables:} $seed$, $start = 0$,
-    $height$, $seq$, $trie$
-    \State \textbf{Initialize containers:} $hits$, $pebbles$
+    \State \textbf{Define:} $\tau$
+    \State \textbf{Variables:} $seed$, $start = 0$,
+    $height$, $seq$, $trie$, $lastseq$, $k$
+    \State \textbf{Containers:} $hits$, $pebbles$
     \State \textsc{read} sequence file
@@
     \State \textsc{sort} sequences alphabetically
+    \State $k \gets$ \textsc{compute} lookup sequence length
+    \Comment{Eq. 3}
     \State $trie \gets$ \textsc{create} an empty trie of height $height$
@@
     \State $seq \gets$ \textsc{get} next sequence
+    \If{at least one $k$-mer of $seq$ is in the lookup table}
     \State $seed \gets$ \textsc{length} of shared prefix
-    between current and next $seq$
+    between current and next sequence
+    \State $start \gets$ \textsc{length} of shared prefix
+    between $seq$ and $lastseq$
     \State \textsc{clear} hits
@@
     \State $node \gets$ \textsc{get} next node from $pebbles$
-    \State \textbf{call} \textsc{poucet}($node$, $seed$, $hits$)
+    \State \textbf{call} \textsc{poucet}($seq$, $node$, $seed$, $hits$)
     \EndFor
+    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
+    \State $lastseq \gets seq$
+    \EndIf
     \State \textsc{insert} $seq$ path in $trie$
-    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
-    \State $start \gets seed$
+    \State \textsc{insert} $seq$ $k$-mers into the lookup table
     \EndFor
@@
   \begin{algorithmic}[1]
-    \Procedure{poucet}{$node$, $seed$, $hits$}:
-    \State \textsc{compute} $node$-specific column values following NW algorithm
+    \Procedure{poucet}{$query$, $node$, $seed$, $hits$}:
+    \State \textsc{compute} $node$-specific column values following NW
+    algorithm \Comment{Fig.1}
     \ForAll{$child$ nodes in $node$}
-    \State \textsc{compute} $child$-specific row values following NW algorithm
-    \State \textsc{compute} center value using row and column
-    values
+    \State \textsc{compute} $child$-specific row values following NW
+    algorithm \Comment{Fig.1}
+    \State \textsc{compute} center value using row and column values \Comment{Fig.1}
     \If{center value $>\tau$} \Comment{Mismatches exceeded.}

### 2014-06-30 starcode.c
Added useq_t comparison function.

@@ -1065,3 +1065,26 @@ const void *bp
    if (la == lb)  return strcmp(a,b);
    return (la < lb ? -1 : 1);
 }
+
+int
+AtoZ_useq
+(
+ const void *ap,
+const void *bp
+ )
+{
+   useq_t * a = (useq_t *) a;
+   useq_t * b = (useq_t *) b;
+   int la = strlen(a->seq);
+   int lb = strlen(b->seq);
+   if (la == lb) {
+      int cmp = strcmp(a->seq, b->seq);
+      if (cmp == 0) {
+         a->count += b->count;
+         free(b->seq);
+         free(b);
+      }
+      return cmp;
+   }
+   return (la < lb ? -1 : 1);
+}

### 2014-07-01 starcode.c
The function 'read_file()' returns stack of useq.

@@ -26,8 +26,9 @@ mtplan_t *plan_mt(int, int, int, int, gstack_t*, const int);
 void message_passing_clustering(gstack_t*, int);
 void sphere_clustering(gstack_t*, int);
 void * _mergesort(void *);
-int seqsort (void **, int, int (*)(const void*, const void*), int);
+int seqsort(void **, int, int (*)(const void*, const void*), int);
 long count_trie_nodes(useq_t **, int, int);
+int AtoZ_useq(const void *, const void *);
 
 FILE *OUTPUT = NULL;
 
@@ -51,18 +52,15 @@ starcode
    if (ntries % 2 == 0) abort();
    
    if (verbose) fprintf(stderr, "reading input files\n");
-   gstack_t *seqS = read_file(inputf);
+   gstack_t *useqS = read_file(inputf);
 
-   if (verbose) fprintf(stderr, "preprocessing\n");
-   // Count unique sequences.
-   gstack_t *useqS = seq2useq(seqS, maxthreads);
-   for (int i = 0 ; i < seqS->nitems ; i++) free(seqS->items[i]);
-   free(seqS);
+   seqsort(useqS->items, useqS->nitems, AtoZ_useq, maxthreads);
+   
    // Pad sequences. (And return the median size)
-   int median;
-   int height = pad_useq(useqS, &median);
+   int med;
+   int height = pad_useq(useqS, &med);
    // Make multithreading plan.
-   mtplan_t *mtplan = plan_mt(tau, height, median, ntries, useqS, clusters);
+   mtplan_t *mtplan = plan_mt(tau, height, med, ntries, useqS, clusters);
    // Run the query.
    run_plan(mtplan, verbose, maxthreads);
    if (verbose) fprintf(stderr, "starcode progress: 100.00%%\n");
@@ -610,8 +608,14 @@ _mergesort
       // Accumulate repeats
       sortargs->repeats = arg1.repeats + arg2.repeats;
 
+      int i = 0;
+      int j = 0;
+      int nulli = 0;
+      int nullj = 0;
+      int cmp = 0;
+      int repeats = 0;
+
       // Merge sets
-      int i = 0, j = 0, nulli = 0, nullj = 0, cmp = 0, repeats = 0;
       for (int k = 0, idx = 0, n = 0; k < sortargs->size; k++) {
          if (j == arg2.size) {
             // Insert pending nulls, if any.
@@ -678,14 +682,13 @@ read_file
[…]

### 2014-07-01 starcode.c
Updated seqsort: NULL pointers at the end of the array.

@@ -54,16 +54,22 @@ starcode
    if (verbose) fprintf(stderr, "reading input files\n");
    gstack_t *useqS = read_file(inputf);
 
-   seqsort(useqS->items, useqS->nitems, AtoZ_useq, maxthreads);
-   
+   // Sort and realloc.
+   int nuseq = seqsort(useqS->items, useqS->nitems, AtoZ_useq, maxthreads);
+   useqS = realloc(useqS, gstack_size(nuseq));
+   useqS->nitems = useqS->nslots = nuseq;
+
    // Pad sequences. (And return the median size)
    int med;
    int height = pad_useq(useqS, &med);
+   
    // Make multithreading plan.
    mtplan_t *mtplan = plan_mt(tau, height, med, ntries, useqS, clusters);
+
    // Run the query.
    run_plan(mtplan, verbose, maxthreads);
    if (verbose) fprintf(stderr, "starcode progress: 100.00%%\n");
+
    // Remove padding characters.
    unpad_useq(useqS);
 
@@ -564,7 +570,6 @@ seqsort
    free(buffer);
    
    return numels - args.repeats;
-
 }
 
 void *
@@ -610,61 +615,31 @@ _mergesort
 
       int i = 0;
       int j = 0;
-      int nulli = 0;
-      int nullj = 0;
+      int idx = 0;
       int cmp = 0;
       int repeats = 0;
 
       // Merge sets
-      for (int k = 0, idx = 0, n = 0; k < sortargs->size; k++) {
-         if (j == arg2.size) {
-            // Insert pending nulls, if any.
-            for (n = 0; n < nulli; n++)
-               buf[idx++] = NULL;
-            nulli = 0;
-            buf[idx++] = l[i++];
-         }
-         else if (i == arg1.size) {
-            // Insert pending nulls, if any.
-            for (n = 0; n < nullj; n++)
-               buf[idx++] = NULL;
-            nullj = 0;
-            buf[idx++] = r[j++];
-         }
-         else if (l[i] == NULL) {
-            nulli++;
-            i++;
-         }
-         else if (r[j] == NULL) {
-            nullj++;
-            j++;
-         }
+      while (i + j < sortargs->size) {
+         // Only NULLS at the end of the buffers.
+         if (l[i] == NULL && r[j] == NULL) break;
+         // Buffers exhausted.
+         if (j == arg2.size || r[j] == NULL)      buf[idx++] = l[i++];
+         else if (i == arg1.size || l[i] == NULL) buf[idx++] = r[j++];
[…]

### 2014-07-01 view.c
Graphical output files. Not functional at the moment.

@@ -1,0 +1,503 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <math.h>
+#include <time.h>
+#include <errno.h>
+#include <cairo.h>
+//#include <execinfo.h>
+//#include <signal.h>
+//#include <unistd.h>
+#include "view.h"
+
+//void SIGSEGV_handler(int sig) {
+//   void *array[10];
+//   size_t size;
+//
+//   // get void*'s for all entries on the stack
+//   size = backtrace(array, 10);
+//            
+//   // print out all the frames to stderr
+//   fprintf(stderr, "Error: signal %d:\n", sig);
+//   backtrace_symbols_fd(array, size, STDERR_FILENO);
+//   exit(1);
+//}
+
+int main(int argc, const char *argv[])
+{
+   //signal(SIGSEGV, SIGSEGV_handler);
+   // Create list of ball pointers.
+   //FILE * inputf = fopen("example_starcode_output.txt", "r");
+   //int n_balls;
+   //ball_t * ball_list = create_ball_list(inputf, &n_balls);
+
+   /* Begin test code */
+   const int n_balls = 8;
+   ball_t * ball_list[8]; // in the stack
+   ball_list[0] = malloc(BALL_SIZE(3));
+   ball_list[1] = malloc(BALL_SIZE(0));
+   ball_list[2] = malloc(BALL_SIZE(0));
+   ball_list[3] = malloc(BALL_SIZE(0));
+   ball_list[4] = malloc(BALL_SIZE(2));
+   ball_list[5] = malloc(BALL_SIZE(1));
+   ball_list[6] = malloc(BALL_SIZE(0));
+   ball_list[7] = malloc(BALL_SIZE(0));
+
+   // Cluster 1
+   ball_list[0]->size = 1000;
+   ball_list[0]->n_children = 3;
+   ball_list[0]->root = ball_list[0];
+   ball_list[0]->children[0] = ball_list[1];
+   ball_list[0]->children[1] = ball_list[2];
+   ball_list[0]->children[2] = ball_list[3];
+
+   ball_list[1]->size = 10;
+   ball_list[1]->n_children = 0;
+   ball_list[1]->root = ball_list[0];
+   
+   ball_list[2]->size = 5;
+   ball_list[2]->n_children = 0;
+   ball_list[2]->root = ball_list[0];
+
+   ball_list[3]->size = 30;
+   ball_list[3]->n_children = 0;
[…]

### 2014-07-01 view.h

@@ -1,0 +1,51 @@
+#define BALL_SIZE(elem) 2 * sizeof(int)   + \
+                        4 * sizeof(double) + \
+                  (elem+1)* sizeof(ball_t *)
+#define WIDTH 600.0
+#define RAND_FACTOR WIDTH / RAND_MAX
+#define ASPECT_TYPE 0
+#if ASPECT_TYPE == 0 
+   #define ASPECT 4 / 3
+#elif ASPECT_TYPE == 1
+   #define ASPECT 16 / 9
+#endif
+#define PI 3.141592653589793238462643383279502884L
+
+struct ball_t;
+struct cluster_t;
+
+typedef struct ball_t ball_t;
+typedef struct cluster_t cluster_t;
+
+struct ball_t {
+   int      size;        // # of barcodes
+   int      n_children;  // # of children
+   double   position[2]; // (x,y)
+   double   force[2];    // (x,y)
+   ball_t * root;        // address to root
+   ball_t * children[];  // list of direct children
+};
+
+struct cluster_t {
+   ball_t * root;          // address of root ball_t
+   double position[2];     // (x,y) initial position of the root
+   double displacement[2]; // (x,y) distance from position
+   double radius;          // distance to most distant ball + its radius
+};
+
+ball_t ** create_ball_list(char * filename);
+ball_t ** new_ball(int n_children);
+double norm(double coord_x, double coord_y);
+double electric(ball_t * ball1, ball_t * ball2, double dist);
+double elastic(double dist);
+void compute_force(ball_t * ball1, ball_t * ball2, int force_type);
+double move_ball(ball_t * ball, double kt);
+int compar(const void * elem1, const void * elem2);
+cluster_t ** create_cluster_list(int n_balls, ball_t ** ball_list);
+void spiralize_displacements(int n_clusters, cluster_t ** cluster_list);
+void move_clusters(int n_balls, int n_clusters,
+                   ball_t ** ball_list, cluster_t ** cluster_list);
+void measure_space(int n_balls, int * max_size, ball_t ** ball_list);
+void draw_cairo_env(cairo_t * cr, int n_balls, ball_t ** ball_list, int last);
+void draw_edges(cairo_t * cr, ball_t * ball, int last);
+void draw_circles(cairo_t * cr, ball_t * ball);

### 2014-07-01 doc/document.tex
Added lookup search section in the manuscript.

@@
 
+\subsection{Lookup search}
+If the Levenshtein distance between two sequences of length $L$ is
+less then or equal to $\tau$, then there must exist a perfect
+match of length at least $k = L / (\tau+1)$, assuming for the time being
+that $k$ is an integer.
+We take advantage of this property to build a lookup table containing
+all such $k$-mers from sequences inserted in the trie. When
+a new sequence is queried, subsequences are first extracted from it
+and queried against the lookup table. If none of the subsequences
+are found, the minimum distance between the query and all the
+sequences in the trie is at least $\tau+1$, and the trie-search can
+therefore be omitted. This lookup is particulary useful for long query
+sequences because in this case $k$ is large. The probability of an
+exact match between subsequences of length $k$ is small and the
+trie-search will be avoided for many such queries. The absence of match
+is detected at the computational cost of several table lookups, compared
+to the higher cost of measuring the Levenshtein distance between long
+sequences.
+
+In general $L / (\tau+1)$ is not an integer. To deal
+with this complication, we divide the query into $\tau+2$ subsequences.
+The first $k+1$ are called words and have target length
+$(L-\tau) / (\tau+1)$. The last subsequence is called the tail and has
+length $\tau$ for reasons explained below. The length of the words is
+computed by integer division, and the remainder $R$ is distributed
+equally among the last $R$ words. As a result, the words have a length
+that differs by at most 1, with a total length equal to $L-\tau$.
+
+A word at the beginning of the query is not a match for the same
+word at the end of a sequence in the trie. For this reason, we query
+the words against $k+1$ distinct lookup tables. In order to allow
+insertions and deletions, the lookup search must also query shifted
+versions of the words. In the worst case, only one word is found and the
+mismatches are evenly distributed in the query sequence. The maximum
+number of cumulative insertions/deletions at the $i$-th word
+is $i-1$. Hence, the $i$-th word has to be shifted and queried against
+the lookup table $2(i-1)+1$ times to cover all the possible cases.
+The last word will therefore be queried $2\tau+1$ times, including
+$\tau$ shifts on the right and $\tau$ shifts on the left. For the latter
+we use the $\tau$ nucleotides stored in the tail.
+
 \subsection{Seek and construct}
@@
     \State \textsc{sort} sequences alphabetically
+    \State $k \gets$ \textsc{compute} lookup word lengths
     \State $trie \gets$ \textsc{create} an empty trie of height $height$
@@
     \State $seq \gets$ \textsc{get} next sequence
+    \If{at least one $k$-mer of $seq$ is in the lookup table}
     \State $seed \gets$ \textsc{length} of shared prefix
-    between current and next $seq$
-    \State \textsc{clear} $hits$
+    between current and next sequence
+    \State $start \gets$ \textsc{length} of shared prefix
+    between $seq$ and $lastseq$
+    \State \textsc{clear} hits
     \State \textsc{clear} $pebbles$ at depth $>start$
@@
     \EndIf
-    \State \textbf{call} \textsc{poucet}($child$, $seed$, $hits$)
+    \State \textbf{call} poucet($child$, $seed$, $hits$)
     \EndFor

### 2014-07-01 test/testset.c
Removed useless functions and updated tests.

@@ -1062,14 +1062,16 @@ test_starcode_2
 (void)
 // Test 'seqsort()'.
 {
-   char *to_sort_1[] = {
+
+   // Case 1 (no repeat).
+   char *sequences_1[10] = {
       "IRrLv<'*3S?UU<JF4S<,", "tcKvz5JTm!h*X0mSTg",
       "tW:0K&Mvtax<PP/qY6er", "hcU+f!=`.Xs6[a,C7XpN",
       ":3ILp'w?)f]4(a;mf%A9", "RlEF',$6[}ouJQyWqqT#",
       "U Ct`3w8(#KAE+z;vh,",  "[S^jXvNS VP' cwg~_iq",
       ".*/@*Q/]]}32kNB#`qqv", "#`Hwp(&,z|bN~07CSID'",
    };
-   const char *sorted_1[] = {
+   const char *sorted_1[10] = {
       "tcKvz5JTm!h*X0mSTg",   "U Ct`3w8(#KAE+z;vh,",
       "#`Hwp(&,z|bN~07CSID'", ".*/@*Q/]]}32kNB#`qqv",
       ":3ILp'w?)f]4(a;mf%A9", "IRrLv<'*3S?UU<JF4S<,",
@@ -1077,21 +1079,40 @@ test_starcode_2
       "hcU+f!=`.Xs6[a,C7XpN", "tW:0K&Mvtax<PP/qY6er", 
    };
 
-   seqsort((void **) to_sort_1, 10, AtoZ, 1);
+   useq_t *to_sort_1[10];
    for (int i = 0 ; i < 10 ; i++) {
-      g_assert_cmpstr(to_sort_1[i], ==, sorted_1[i]);
+      to_sort_1[i] = new_useq(1, sequences_1[i]);
    }
 
-   char *to_sort_2[] = {
+   seqsort((void **) to_sort_1, 10, AtoZ, 1);
+   for (int i = 0 ; i < 10 ; i++) {
+      g_assert_cmpstr(to_sort_1[i]->seq, ==, sorted_1[i]);
+      g_assert_cmpint(to_sort_1[i]->count, ==, 1);
+      free(to_sort_1[i]);
+   }
+
+   // Case 2 (repeats).
+   char *sequences_2[6] = {
       "repeated", "repeated", "repeated", "repeated", "repeated", "xyz"
    };
-   char *sorted_2[] = {
-      "xyz", NULL, NULL, NULL, NULL, "repeated",
[…]

### 2014-07-02 doc/document.tex
Corrected a bug. Added plots for the manuscript.

@@
     BWA 0.7.9a & \texttt{mem -t 20 -a -k1 -B0} \\ \hline
-    GEM indexer & \texttt{gem-indexer -T 20} \\ \hline
-    GEM 1.423 & \texttt{-e3 -s -T 20 --granularity 100000}\\ \hline
+    GEM indexer & \texttt{gem-indexer -T 12} \\ \hline
+    GEM 1.423 & \texttt{-e3 -s4 -T12 --granularity 100000}\\ \hline
     \emph{sequential R} & \texttt{method='lv', maxDist=2}\\ \hline
@@
 \subsection{Benchmark against sequence clustering algorithms}
-Sequence clustering is routinely used to curate databases of non redundant
-nucleotide or protein sequences. We benchmarked starcode against the two popular
-sequence clustering algorithms CD-HIT \citep{pmid23060610} and USEARCH 
-\citep{pmid20709691}.
+Sequence clustering is routinely used to curate databases of non
+redundant nucleotide or protein sequences. We benchmarked starcode
+against the two popular sequence clustering algorithms CD-HIT
+\citep{pmid23060610} and USEARCH \citep{pmid20709691}.
 
-We generated 1 million random 40-mers that were duplicated 47 times. For each
-40-mer, we inserted 3 addional mutant sequences to the pool by sampling again
-3 nucleotides
-at random positions. The probability that such a mutant sequence is within a
-distance 3 of any other 40-mer is of the order of $10^{-12}$ and can be neglected.
-Each cluster thus consists of 50 sequences, of which 3 have a distance at most
-3 from the canonical representant.
+We generated 1 million random 40-mers that were duplicated 47 times.
+For each 40-mer, we inserted 3 addional mutant sequences to the pool
+by sampling again 3 nucleotides at random positions. The probability
+that such a mutant sequence is within a distance 3 of any other
+40-mer is of the order of $10^{-12}$ and can be neglected. Each
+cluster thus consists of 50 sequences, of which 3 have a Levenshtein
+distance 3 from the canonical representant.
 
-Using 20 cores, starcode clustered the shuffled 50 million sequences in 4 minutes
-without a single error (the output consisted of 1 million clusters of size 50).
-We submitted the unique sequences of the dataset to CD-HIT and USEARCH. The
-results are summarized in Figure~\ref{fig:04}a and \ref{fig:04}b. The running
-times of CD-HIT and USEARCH
-were 59 and 64 minutes respectively, and they identified 2,362,352 and 2,739,954
-clusters respectively. Both algorithms rely on heuristics, fact that explains why the
-number of clusters is off target. These results clearly demonstrate that starcode
-outperforms other clustering algorithms in speed and accuracy on this particular task.
+Using 12 cores, starcode clustered the shuffled 50 million sequences
+in 4 minutes and 32 seconds without a single error (the output
+consisted of 1 million clusters of size 50).  The running times
+of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
+identified 2,362,352 and 2,739,954 clusters respectively. The results
+are summarized in Figure~\ref{fig:04}. Both algorithms rely on
+heuristics, explaining why the number of clusters is off target. These
+results clearly demonstrate that starcode outperforms other clustering
+algorithms in speed and accuracy to cluster random sequences.
+
+We also compared starcode on the biological benchmark dataset proposed
+by \cite{pmid23060610}.
 
@@
 \centerline{\includegraphics{Fig4.pdf}}
-\caption{Benchmark of starcode against sequence clustering algorithms and short
-read mappers. \textbf{a} Running time compared to CD-HIT and UCLUST on the
-dataset described in the text. \textbf{b} Error rate compared to CD-HIT and UCLUST,
-expressed as the percentage of clusters failing to be merged. \textbf{c} Running
-time compared to Bowtie2, BWA and GEM on this dataset. The running time is
-decomposed in extraction of unique sequences (purple), Burrows-Wheeler indexing
-(yellow), and query proper (white). The extraction was performed with the Linux
-sort command, so the time is the same in every case. \textbf{d} Error rate compared
-to Bowtie2, BWA and GEM, expressed as the proportion of matches with a single hit.
-This is an underestimate of the true error rate.}
+\caption{Benchmark of starcode against sequence clustering algorithms
+and short read mappers. \textbf{a} Running time compared to CD-HIT and
+UCLUST on the dataset described in the text. \textbf{b} Error rate
+compared to CD-HIT and UCLUST, expressed as the percentage of clusters
+failing to be merged. \textbf{c} Running time compared to Bowtie2, BWA
+and GEM on this dataset. The running time is decomposed in extraction
+of unique sequences (purple), Burrows-Wheeler indexing (yellow), and
+query proper (white). The extraction was performed with the Linux sort
+command, so the time is the same in every case. \textbf{d} Error rate
+compared to Bowtie2, BWA and GEM, expressed as the proportion of matches
+with a single hit. This is an underestimate of the true error rate.}
 \label{fig:04}

### 2014-07-03 doc/document.tex
Updated the manuscript. Added a figure.

@@
 (NW) algorithm \citep{pmid5420325}. In the original algorithm
-(Figure~\ref{fig:01}a), the Levenshtein distance between two sequences is found
+(Figure~\ref{fig:NW}a), the Levenshtein distance between two sequences is found
 by applying a recurrence relation throughout a matrix of $mn$ terms, where $m$
@@
 
-\begin{figure}[!tpb]%figure1
-\centerline{\includegraphics{Fig1.pdf}}
+\begin{figure}[!tpb]
+\centerline{\includegraphics{NW.pdf}}
 \caption{Needleman-Wunsch (NW) alignment. \textbf{a}
@@
 nevertheless identical. The values of the white cells are never computed, which
-contributes to making the algorithm less complex.}\label{fig:01}
+contributes to making the algorithm less complex.}\label{fig:NW}
 \end{figure}
@@
 of initializing the margins of the matrix and computing all the
-terms, the matrix is initialized as shown on Figure~\ref{fig:01}b and only the
+terms, the matrix is initialized as shown on Figure~\ref{fig:NW}b and only the
 terms around
@@
 matrix are updated row-wise as a depth-first search traverses the trie from the
-root, as illustrated in Figure~\ref{fig:02}. Every time a node is visited, the row
+root, as illustrated in Figure~\ref{fig:trie}. Every time a node is visited, the row
 corresponding to its depth is recomputed. If the threshold value $d$ is exceeded
@@
 
-\begin{figure}[!tpb]%figure2
-\centerline{\includegraphics{Fig2.pdf}}
+\begin{figure}[!tpb]
+\centerline{\includegraphics{trie.pdf}}
 \caption{Trie query with NW alignment. Each sequence of the index is a path in
 the trie. The query GTTGCA is written at the top of a NW matrix, which is
-initialized as shown on Figure~\ref{fig:01}b. The trie is traversed by a
+initialized as shown on Figure~\ref{fig:NW}b. The trie is traversed by a
 depth-first search (red path) from the root. At each depth, the node added to
@@
 in red). Even if this node has children, they are not visited (red crosses).}
-\label{fig:02}
+\label{fig:trie}
 \end{figure}
@@
 form an angle shape, looking
-like a horizontally flipped L character as shown on Figure~\ref{fig:03}. This way, the
+like a horizontally flipped L character as shown on Figure~\ref{fig:poucet}. This way, the
 computation intermediates stored in a node of depth $k$ depend only on the previous
@@
 
-\begin{figure}[!tpb]%figure3
-\centerline{\includegraphics{Fig3.pdf}}
+\begin{figure}[!tpb]
+\centerline{\includegraphics{poucet.pdf}}
 \caption{Poucet search algorithm. The algorithm proceeds with the same principles
-as shown on Figure~\ref{fig:02} with the difference that the NW matrix is not
+as shown on Figure~\ref{fig:trie} with the difference that the NW matrix is not
 updated row-wise, but along horizontally flipped L shapes. As the depth-first
@@
 $k$ with the current query.}
-\label{fig:03}
+\label{fig:poucet}
 \end{figure}
@@
 are found, the minimum distance between the query and all the
-sequences in the trie is at least $\tau+1$, and the trie-search can
+sequences in the trie is at least $\tau+1$, and the trie search can
 therefore be omitted. This lookup is particulary useful for long query
-sequences because in this case $k$ is large. The probability of an
-exact match between subsequences of length $k$ is small and the
-trie-search will be avoided for many such queries. The absence of match
-is detected at the computational cost of several table lookups, compared
-to the higher cost of measuring the Levenshtein distance between long
-sequences.
+sequences (see \ref{scalability}). In this case $k$ is large, so the
+probability of an exact match between subsequences of length $k$ is small.
+The absence of match is detected at the computational cost of several
+table lookups, compared to the higher cost of measuring the Levenshtein
+distance between long sequences.
 
@@
 trie as ``thin'' as possible, thereby reducing the search time.
-%The whole matching process is summarized in the pseudocode shown Figure~\ref{fig:04}
 The whole matching process is summarized in the pseudocode shown in
@@
 \section{Results}
+\subsection{Scalability}
+\label{scalability}
+Write some crap here.
+
+\begin{figure}[!tpb]%figurex
+\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
+\caption{Scalability. Top left: the running time of starcode is
+approximlately $O(n^{1.7})$, where $n$ is the number of sequences to
+cluster. Top right: the running time is very sensitive to the maximum
+Levenshtein distance used in custering.  Bottom left: the running time
+as a function of the length of the sequences has a complex behavior. It
+increases initially but then decreases for long sequences because the
+lookup search becomes more effective. Bottom right: the scaling of the
+relative performance is sub-linear as the number of threads increases.
+The algorithm is fully parallel, but the memory bandwidth becomes
+limiting on the tested hardware.
+}\label{fig:X}
+\end{figure}
+
 \subsection{Benchmark against sequence clustering algorithms}
@@
 Using 12 cores, starcode clustered the shuffled 50 million sequences
-in 4 minutes and 32 seconds without a single error (the output
+in 5 minutes and 17 seconds without a single error (the output
 consisted of 1 million clusters of size 50).  The running times
@@
 identified 2,362,352 and 2,739,954 clusters respectively. The results
-are summarized in Figure~\ref{fig:04}. Both algorithms rely on
+are summarized in Figure~\ref{fig:benchmark}. Both algorithms rely on
 heuristics, explaining why the number of clusters is off target. These
@@
 
-\begin{figure}[!tpb]%figure5
-\centerline{\includegraphics{Fig4.pdf}}
+\begin{figure}[!tpb]
+\centerline{\includegraphics{benchmark.pdf}}
 \caption{Benchmark of starcode against sequence clustering algorithms
@@
 with a single hit. This is an underestimate of the true error rate.}
-\label{fig:04}
+\label{fig:benchmark}
 \end{figure}
@@
 sequence has at least 2 distinct matches at a distance 3 (mutant sequences
-from the same cluster may be at a distance up to 6). This means that we
+from the same cluster may be at a distance up to 6 from each other). This
+means that we
 can use matches with a single hit (themselves) as a lower bound on the inaccuracy.
 We used unique sequences to build an index and queried the index with the same
-sequences. Figure~\ref{fig:04}c shows the running time of starcode along with the
+sequences. Figure~\ref{fig:benchmark}c shows the running time of starcode along with the
 decomposition of the running time for the three mappers. Note that the time to
 extract unique sequences is always the same because we used the Linux
-\texttt{sort -u} command in every case. It appears that the running time of
-starcode is about the same as the time required by BWA (the fastest aligner in
-this test) to index and run the query proper. Figure~\ref{fig:04}d shows the
-lower bound on
-the error rate, as estimated by the proportion of single hits. BWA was found to
-be the most accurate aligner with a lower bound of 19\% errors. This measure
-is indeed an underestimate of the true error rate, but given that starcode makes no
-mistake on this dataset, a more accurate measure is not necessary. In conclusion,
-starcode runs slightly faster than short read mappers when all steps are taken
-into account, and clearly outperforms them in precision on this particular problem.
+\texttt{sort -u} command in every case. The running time of
+starcode is slightly lower than short read mappers when all steps ares
+taken into account.  Figure~\ref{fig:benchmark}d shows the lower bound on
+the error rate, as estimated by the proportion of single hits. BWA was
+found to be the most accurate aligner with a lower bound of 31\% sequences.
+with a single match. This measure is an underestimate of the true error
+rate, but given that starcode is an exact algorithm, a more accurate
+measure is not necessary. In conclusion, starcode can compete with the
+short read aligners on the matching problem, and it provides a precision
+that none of the other tools can offer.
 
@@
 the 12-mers with starcode within a Levenshtein distance of 2 took less
-than 3 minutes with 20 cores. We identified the known DNA uptake sequence of
-\textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant 12-mer, comprising
-1,466 exact matches and 4,625 inexact matches. This result testifies to the
-fact that starcode can be used to identify biologically relevant motifs in
-bacterial genomes.
+than 45 seconds with 12 cores. We identified the known DNA uptake sequence of
+\textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant 12-mer.
+This result testifies to the fact that starcode can be used to identify biologically
+relevant motifs in bacterial genomes.
 
-To test starcode on the scale of a metazoan genome, we used a \textit{Drosophila}
-dataset from \cite{pmid20888037}. The study describes two
-signatures of chromatin proteins called Red and Yellow, both corresponding
-to transcribed regions of the genome. The distinction is based on epigenomic
-features only, which prompted us to identify sequence features that could
-discriminate Red \textit{versus} Yellow regions. To this end, we used starcode to
-cluster 18-mers from both regions.
-
-The total coverage of Red and Yellow chromatin in the \textit{Drosophila} genome is
-10.7 Mb and 21.3 Mb respectively, corresponding to 21.5 million and 42.5 million
-18-mers, respectively. We used starcode to identify clusters of 18-mers within
-a Levenshtein distance of 3 (the running time was approximately 35 minutes for
-Red and 70 minutes for Yellow chromatin using 20 threads). The most discriminating
-18-mer was (GA)$_9$, which showed a 4.8 fold enrichment in Red \textit{versus} Yellow
-chromatin.
-We counted 832 exact matches and 6,749 inexact matches over 3,880 distinct
-18-mers in Red chromatin, showing that a larger portion of the signal comes
-from inexact matches. Taken together, these results show that starcode can be
-used to cluster short k-mers from metazoan-scale datasets.
+To test starcode on another application, we used the RNA-protein
+interaction data produced by RNAcompete \citep{pmid19561594}. The mammalian
+splicing factor SRSF1 is known to bind RNA GA-rich motifs, but there is
+some disagreement about the motif that it recognizes \citep{pmid23562324}.
+For each replicate, we replaced the RNAcompete signals by their rank and
+extracted the 10-mers from the microarray probes. The 10-mers were given
+a score equal to the rank of the probe they belong, and enriched motifs
+were found using the sphere clustering of starcode with max Levenshtein
+distance 2. The score of the most enriched 10-mer is thus the sum of the
+ranks of all 10-mers within this distance. Clustering the 6.3 million
+extracted 10-mers with 12 cores took about 20 seconds for each replicate.
+The most enriched 10-mers were AGGACACGGA, AGGACACGGA, AGGACGGAGG,
+AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except for the last replicate, the
+motifs consist of AGGAC and GGA, with a spacer of variable length. This
+suggests that the binding of SRSF1 to RNA is flexible, which explains the
+poor agreement between the motifs derived from 6-mers or 7-mers.
 
@@
 recover a known enriched 12-mer in the genome of \textit{Neisseria meningitidis}.
-In the second, we identify a sequence feature that discriminates genomic
-regions found in two different chromatin types in \textit{Drosophila}. GA
-repeats are known to be bound \textit{in vivo} by the protein GAF
-\citep{pmid12601174}, which suggests
-that this transcription factor is instrumental in setting the distinction
-between Red and Yellow chromatin.
 

### 2014-07-03 doc/plot-scalability.R

@@ -1,0 +1,21 @@
+# Data
+X1 = c(0,1,2,3,4,5)
+Y1 = c(0.465, 0.631, 0.776, 19.072, 2*60+20.767, 9*60+13.506)
+X2 = 10000*c(1,2,4,8,16,32,64)
+Y2 =c(0.294, 0.966, 3.374, 11.849, 39.074, 60+55.348, 4*60+55.366)
+X3 = c(1,2,4,6,8,12)
+Y3 = c(40*60+9, 20*60+20, 10*60+41, 7*60+43, 6*60+2, 4*60+32)
+X4 = 10*c(1,2,3,4,5,6,7,8)
+Y4 = c(22.072, 25.714, 26.361, 19.865, 5.114, 4.959, 5.591, 6.257)
+# Plot
+pdf("/mnt/shared/src/C/starcode/doc/scalability.pdf")
+par(mfrow=c(2,2))
+plot(log(X2),log(Y2), type='b', xlab="Log number of sequences", ylab="Log Running time (seconds)", cex.lab=1.3)
+abline(lm(log(Y2)~log(X2)), col=2)
+text(x=13, y=5.65, "-16.7+1.7x", adj=1)
+plot(X1,Y1, type='b', xlab="Levenshtein distance", ylab="Running time (seconds)", cex.lab=1.3)
+plot(X4,Y4, type='b', ylim=c(0,30), xlab="Sequence length", ylab="Running time (seconds)", cex.lab=1.3)
+plot(X3,Y3[1]/Y3, type='b', ylim=c(1,12), xlim=c(1,12), xlab="Number of threads", ylab="Relative performance", cex.lab=1.3)
+abline(0,1, col=2)
+dev.off()
+

### 2014-07-03 doc/document.tex
Added counts for N. meningitidis.
Written 'scalability' section. Moved 'Benchmark options' to Results section. Corrected some typos.

@@
 performing a pairwise comparison of all the barcodes, which is unfeasible for
-excessive computationally complexity.
+excessive computational complexity.
 
@@
 \subsection{Lookup search}
-If the Levenshtein distance between two sequences of length $L$ is
-less then or equal to $\tau$, then there must exist a perfect
-match of length at least $k = L / (\tau+1)$, assuming for the time being
-that $k$ is an integer.
-We take advantage of this property to build a lookup table containing
-all such $k$-mers from sequences inserted in the trie. When
-a new sequence is queried, subsequences are first extracted from it
-and queried against the lookup table. If none of the subsequences
-are found, the minimum distance between the query and all the
-sequences in the trie is at least $\tau+1$, and the trie search can
-therefore be omitted. This lookup is particulary useful for long query
-sequences (see \ref{scalability}). In this case $k$ is large, so the
-probability of an exact match between subsequences of length $k$ is small.
-The absence of match is detected at the computational cost of several
-table lookups, compared to the higher cost of measuring the Levenshtein
-distance between long sequences.
+If the Levenshtein distance of two sequences of length $L$ is
+less than or equal to $\tau$, there must exist a perfect match between
+them of at least $k =\lfloor L / (\tau+1)\rfloor$ nucleotides. We take
+advantage of this property to build a lookup table containing all such
+$k$-mers of the sequences already present in the trie. When a new
+sequence is queried, all its subsequences are extracted and queried
+against the lookup table. If none of the subsequences are found, the
+minimum distance between the query and all the sequences in the trie
+is necessarily greater than $\tau$, and the trie search can therefore
+be omitted. This lookup is particulary useful for long query sequences
+(see \ref{scalability}), where $k$ is large and the probability
+of an exact match between subsequences is small unless they are
+actually related, i.e. they belong to the same cluster. With this
+method, the absence of match is detected at the computational cost of
+several table lookups, compared to the higher cost of measuring the
+Levenshtein distance between long sequences.
 
-In general $L / (\tau+1)$ is not an integer. To deal
+In general, however, $L / (\tau+1)$ is not an integer. To deal
 with this complication, we divide the query into $\tau+2$ subsequences.
-The first $k+1$ are called words and have target length
-$(L-\tau) / (\tau+1)$. The last subsequence is called the tail and has
-length $\tau$ for reasons explained below. The length of the words is
-computed by integer division, and the remainder $R$ is distributed
-equally among the last $R$ words. As a result, the words have a length
-that differs by at most 1, with a total length equal to $L-\tau$.
+The first $k+1$ are called \emph{words} and have a target length
+$(L-\tau) / (\tau+1)$. The last subsequence is called \emph{tail} and has
+length $\tau$. The length of the words is computed by integer
+division, and the remainder $R$ is distributed equally among the last
+$R$ words. As a result, the words have a length that differs by at
+most 1, with an aggregated length equal to $L-\tau$.
 
-A word at the beginning of the query is not a match for the same
-word at the end of a sequence in the trie. For this reason, we query
-the words against $k+1$ distinct lookup tables. In order to allow
-insertions and deletions, the lookup search must also query shifted
-versions of the words. In the worst case, only one word is found and the
-mismatches are evenly distributed in the query sequence. The maximum
-number of cumulative insertions/deletions at the $i$-th word
-is $i-1$. Hence, the $i$-th word has to be shifted and queried against
-the lookup table $2(i-1)+1$ times to cover all the possible cases.
-The last word will therefore be queried $2\tau+1$ times, including
-$\tau$ shifts on the right and $\tau$ shifts on the left. For the latter
-we use the $\tau$ nucleotides stored in the tail.
+Since two exact words at different positions in the sequence shall not
+represent a lookup match, we store and query the words against $k+1$
+distinct lookup tables. In order to allow insertions and deletions, 
+the lookup search must also query shifted versions of each word. In
+the worst matching case, only one word is found and the mismatches are
+evenly distributed in the query sequence. In such case, the maximum 
+number of cumulative insertions/deletions at the $i$-th word is
+$i-1$. Hence, the $i$-th word has to be shifted and queried against
+the lookup table $2(i-1)+1$ times to cover all the possible
+cases. Hence, the last word will be queried $2\tau+1$ times, including
+$\tau$ shifts on the right and $\tau$ shifts on the left. For the
+latter shift we make use of the $\tau$ nucleotides stored in the
+tail. The total number of lookups per sequence for a given $\tau$ is
+$(\tau+1)^2$.
 
@@
     \State $node \gets$ \textsc{get} next node from $pebbles$
-    \State \textbf{call} \textsc{poucet}($seq$, $node$, $seed$, $hits$)
+    \State \textbf{call} \textsc{poucet}($seq$, $node$, $seed$,
+    $hits$, $pebbles$)
     \EndFor
@@
   \begin{algorithmic}[1]
-    \Procedure{poucet}{$query$, $node$, $seed$, $hits$}:
-    \State \textsc{compute} $node$-specific column values following NW
-    algorithm \Comment{Fig.1}
+    \Procedure{poucet}{$query$, $node$, $seed$, $hits$, $pebbles$}:
+    \State \textsc{compute} $node$-specific column following NW
+    \Comment{Fig.1}
     \ForAll{$child$ nodes in $node$}
-    \State \textsc{compute} $child$-specific row values following NW
-    algorithm \Comment{Fig.1}
-    \State \textsc{compute} center value using row and column values \Comment{Fig.1}
+    \State \textsc{compute} $child$-specific row following NW
+    \Comment{Fig.1}
+    \State \textsc{compute} center value using row and column \Comment{Fig.1}
     \If{center value $>\tau$} \Comment{Mismatches exceeded.}
@@
     \If{$node$ depth $\leq seed$}
-    \State \textsc{save} $node$ in pebbles at current depth
+    \State \textsc{save} $node$ in $pebbles$ at current depth
     \EndIf
-    \State \textbf{call} poucet($child$, $seed$, $hits$)
+    \State \textbf{call} poucet($query$, $child$, $seed$, $hits$, $pebbles$)
     \EndFor
@@
 To parallelize the search, queries are separated into contiguous blocks after
-sorting. The matching algorithm proceeds in two phases. During the first phase,
-an independent trie is created and filled with the sequences
-of one block using the seek and construct process described above. In the second
-phase, blocks of sequences are queried against but not inserted into the tries
-built in the first phase. If the
-queries are segregated into $N$ blocks, the first phase consists of $N$ seek and
-construct jobs, and the second consists of $N(N-1)/2$ query jobs. Since the jobs
-show little dependence on each other, the matching algorithm can be efficiently
-parallelized provided $N$ is larger than the number of independent threads.
-
+sorting. The matching algorithm proceeds in two phases. During the
+first phase, an independent trie is created and filled with the
+sequences of its associated block using the seek and construct process
+described above. In the second phase, all the other blocks of
+sequences are queried against but not inserted into each trie built in
+the first phase. If the queries are segregated into $N$ blocks, the
+first phase consists of $N$ seek and construct jobs, whereas the second
+consists of $N(N-1)/2$ query jobs. Since the jobs show little
+dependence on each other, the matching algorithm can be efficiently
+parallelized provided $N$ is larger than the number of independent
+threads.
 \subsection{Clustering}
 Starcode implements a multi-purpose clustering algorithm called ``sphere
-clustering'', and a message passing algorithm \citep{mackay} tailored for the
-task of clustering random barcodes. In sphere clustering, barcodes are sorted
-by frequency of occurrence and each barcode, starting from the most frequent,
-can claim its $d$-neighbors that were not already claimed. This is the method
-used by \cite{pmid23953119}.
+clustering'' \citep{pmid23953119}, and a message passing algorithm
+\citep{mackay} tailored for the task of clustering random barcodes. In
+sphere clustering, barcodes are sorted by frequency of occurrence and
+each barcode, starting from the most frequent, can claim its
+$d$-neighbors that were not already claimed.
 
-In message passing clustering, read counts are distributed equally among the
-closest neighbors of each barcode if they are at least 5 times more frequent.
-The remaining sequences at the end of the process are considered
-canonical, and their associated count is the estimated cluster size. A barcode
-is assigned to a cluster if all its read counts are eventually given to the
-corresponding canonical barcode. The barcodes for which the read counts are
-split between different canonical barcodes are not assigned to any cluster. The
-reason for imposing a factor 5 or larger in order to transfer the read counts
-is that barcodes with similar frequencies are not likely derived from each other
-through sequencing errors. More likely they are either unrelated, or they both
-derive from another more abundant barcode.
+In message passing clustering, read counts are distributed equally
+among the closest neighbors of each barcode only if they are at least
+5 times more frequent. The remaining sequences at the end of the
+process are considered canonical, and their associated count is the
+estimated cluster size. A barcode is assigned to a cluster if all its
+read counts are eventually given to the corresponding canonical
+barcode. The barcodes for which the read counts are split between
+different canonical barcodes are not assigned to any cluster. The
+reason for imposing a factor 5 or larger in order to transfer the read
+counts is that barcodes with similar frequencies are not likely
+derived from each other through sequencing errors. More likely they
+are either unrelated, or they both derive from another more abundant
+barcode.
 
-\subsection{Benchmark options}
+\end{methods}
 
-The benchmark dataset is described in the Results section. All software were
-set to run with 20 parallel threads on a 20-core Intel Xeon processor. The
-execution options for each software are summarized in
-Table~\ref{tab:soft-options}.
+\section{Results}
+
+\subsection{Scalability}
+\label{scalability}
+In Figure~\ref{fig:X} we present an evaluation of the performance and
+scalability of starcode as a function of many critical parameters. The
+scalability tests were performed using an **artificial set of random
+sequences**. The sequence length and the number of input sequences
+varies depending on the test. The standard configuration consists of a
+set of **X** sequences of length **Y** running on **Z** threads and a
+cluster distance of **D**. To test the scalability of starcode as a
+function of a single parameter, only the selected value is modified
+whereas the others are kept constant.
+
+In Figure~\ref{fig:X}a we show the running time as a function of the
+selected Levenshtein distance for the clusters. The evolution with the
+increasing distance shows the common exponential trend. Note, however,
+that for distances below 3 the running time remains constant. This is
+due to the fact that the average sequence distance of the input
+sequences lies above this threshold. Consequently, the constant part
+is clearly dependent on the diversity of the input set and when the
+average distance between sequences is reached, the running time scales
+exponentially.
+
+In Figure~\ref{fig:X}b we measure the running time of starcode as a
+function of the number of input sequences $n$. Both axes are in
+logarithmic scale, drawing an evident linear evolution. Based on this
+result, we conclude that the running time of starcode is approximately
+$O(n^{1.7})$.
+
+To prove the effectiveness of the parallel processing, we ran starcode
+over the same dataset using different number of threads. As shown in
+Figure~\ref{fig:X}c, the algorithm is fully parallel as the relative
+performance increases linearly up to 4 threads. The curve bending
+produced thereafter is an artifact of having insufficient memory
+bandwidth to satisfy the increased demand of memory accesses and it is
+due to hardware limitations.
+
+Finally, in Figure~\ref{fig:X}d we investigate the execution time
+with varying sequence lengths. It increases initially due to the
+ineffectiveness of the lookup table and the limited diversity of short
+sequences, i.e. short sequences are less diverse, yielding a bigger
+clustering rate. At a certain length, the lookup search algorithm
+starts to be effective and boosts the performance since most of the
+queries are resolved without the need for trie-search.
+
+\begin{figure}[!tpb]%figurex
+\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
+\caption{Scalability. a. Running time as a function of the clustering
+  distance. b. Logarithm of the running time versus the logarithm of
+  the number of sequences to be clustered. c. Relative performance increase for
+  different number of parallel threads. d. Running time versus length
+  of the input sequences.
+}\label{fig:X}
+\end{figure}
+
+\subsection{Benchmark against sequence clustering algorithms}
+Sequence clustering is routinely used to curate databases of non
+redundant nucleotide or protein sequences. We benchmarked starcode
+against the two popular sequence clustering algorithms CD-HIT
+\citep{pmid23060610} and USEARCH \citep{pmid20709691}.
+
+We generated 1 million random 40-mers that were duplicated 47 times.
+For each 40-mer, we inserted 3 addional mutant sequences to the pool
+by sampling again 3 nucleotides at random positions. The probability
+that such a mutant sequence is within a distance 3 of any other
+40-mer is of the order of $10^{-12}$ and can be neglected. Each
+cluster thus consists of 50 sequences, of which 3 have a Levenshtein
+distance 3 from the canonical representant.
+
+All software were set to run on a 6-core single-processor Intel Xeon
+E5-2620 system with **64~GB** of DDR3-RAM at 1333~Mhz. The execution
+options for each software are summarized in
+Table~\ref{tab:soft-options}. Denoted as \emph{sequential R} is the
+sequential algorithm of \cite{pmid23953119} which is implemented in R
+and relies on the \texttt{stringdist} function of the
+\texttt{stringdist} package.
 
@@
 
-Denoted as \emph{sequential R} is the sequential algorithm of
-\cite{pmid23953119} which is implemented in R and relied on the \texttt{stringdist}
-function of the \texttt{stringdist} package.
-
-\end{methods}
-
-\section{Results}
-\subsection{Scalability}
-\label{scalability}
-Write some crap here.
-
-\begin{figure}[!tpb]%figurex
-\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
-\caption{Scalability. Top left: the running time of starcode is
-approximlately $O(n^{1.7})$, where $n$ is the number of sequences to
-cluster. Top right: the running time is very sensitive to the maximum
-Levenshtein distance used in custering.  Bottom left: the running time
-as a function of the length of the sequences has a complex behavior. It
-increases initially but then decreases for long sequences because the
-lookup search becomes more effective. Bottom right: the scaling of the
-relative performance is sub-linear as the number of threads increases.
-The algorithm is fully parallel, but the memory bandwidth becomes
-limiting on the tested hardware.
-}\label{fig:X}
-\end{figure}
-
-\subsection{Benchmark against sequence clustering algorithms}
-Sequence clustering is routinely used to curate databases of non
-redundant nucleotide or protein sequences. We benchmarked starcode
-against the two popular sequence clustering algorithms CD-HIT
-\citep{pmid23060610} and USEARCH \citep{pmid20709691}.
-
-We generated 1 million random 40-mers that were duplicated 47 times.
-For each 40-mer, we inserted 3 addional mutant sequences to the pool
-by sampling again 3 nucleotides at random positions. The probability
-that such a mutant sequence is within a distance 3 of any other
-40-mer is of the order of $10^{-12}$ and can be neglected. Each
-cluster thus consists of 50 sequences, of which 3 have a Levenshtein
-distance 3 from the canonical representant.
-
-Using 12 cores, starcode clustered the shuffled 50 million sequences
+Using **12** cores, starcode clustered the shuffled 50 million sequences
 in 5 minutes and 17 seconds without a single error (the output
@@
 results clearly demonstrate that starcode outperforms other clustering
-algorithms in speed and accuracy to cluster random sequences.
+algorithms both in speed and accuracy with random input sequences.
 

### 2014-07-03 doc/document.tex
Edited the scalability in the manuscript.

@@
 be omitted. This lookup is particulary useful for long query sequences
-(see \ref{scalability}), where $k$ is large and the probability
+(see \ref{performance}), where $k$ is large and the probability
 of an exact match between subsequences is small unless they are
@@
 
-\subsection{Scalability}
-\label{scalability}
-In Figure~\ref{fig:X} we present an evaluation of the performance and
-scalability of starcode as a function of many critical parameters. The
-scalability tests were performed using an **artificial set of random
-sequences**. The sequence length and the number of input sequences
-varies depending on the test. The standard configuration consists of a
-set of **X** sequences of length **Y** running on **Z** threads and a
-cluster distance of **D**. To test the scalability of starcode as a
-function of a single parameter, only the selected value is modified
-whereas the others are kept constant.
+\subsection{Performance}
+\label{performance}
+We measured basic performance and scalability metrics of starcode on a
+dataset of pseudo random sequences (Figure~\ref{fig:perf}). The standard
+configuration consists of a set of 1,000,000 sequences of length 40 running
+on 1 thread and with a maximum Levenshtein distance of 3. To test the
+scalability as a function of a single parameter, only the parameter under
+study was modified whereas the others were kept constant.
 
-In Figure~\ref{fig:X}a we show the running time as a function of the
-selected Levenshtein distance for the clusters. The evolution with the
-increasing distance shows the common exponential trend. Note, however,
-that for distances below 3 the running time remains constant. This is
-due to the fact that the average sequence distance of the input
-sequences lies above this threshold. Consequently, the constant part
-is clearly dependent on the diversity of the input set and when the
-average distance between sequences is reached, the running time scales
-exponentially.
-
-In Figure~\ref{fig:X}b we measure the running time of starcode as a
-function of the number of input sequences $n$. Both axes are in
-logarithmic scale, drawing an evident linear evolution. Based on this
-result, we conclude that the running time of starcode is approximately
-$O(n^{1.7})$.
-
-To prove the effectiveness of the parallel processing, we ran starcode
-over the same dataset using different number of threads. As shown in
-Figure~\ref{fig:X}c, the algorithm is fully parallel as the relative
-performance increases linearly up to 4 threads. The curve bending
-produced thereafter is an artifact of having insufficient memory
-bandwidth to satisfy the increased demand of memory accesses and it is
-due to hardware limitations.
-
-Finally, in Figure~\ref{fig:X}d we investigate the execution time
-with varying sequence lengths. It increases initially due to the
-ineffectiveness of the lookup table and the limited diversity of short
-sequences, i.e. short sequences are less diverse, yielding a bigger
-clustering rate. At a certain length, the lookup search algorithm
-starts to be effective and boosts the performance since most of the
-queries are resolved without the need for trie-search.
+Figure~\ref{fig:perf}a shows the running time of starcode as a
+function of the number of input sequences $n$. In double logarithmic scale
+the trend is a straight line with slope 1.7, which suggests that the
+running time complexity of starcode is approximately $O(n^{1.7})$.
+Figure~\ref{fig:perf}b shows that the running time gros exponentially
+as a function of the maximum Levenshtein distance used for clustering.
+This is a common feature of exact algorithms based on trie search. As
+this parameter increases, the poucet search bails out at a deeper
+average depth in a trie that fans out exponentially.
+The running time first increases as a function of the sequence length,
+but then plummets and stays low (Figure~\ref{fig:perf}c).
+Beyond a certain length, the lookup search algorithm starts to be
+efficient, and most of the queries are resolved without the need for
+trie-search.
+Finally, in we show the scalability of starcode with increasing
+number of treahds in Figure~\ref{fig:perf}d. The search algorithm is fully
+parallel and the relative performance increases linearly up to 8 threads.
+The bending observed thereafter has two sources. The first is 
+that the input reading and clustering steps are brief but not parallel,
+the second is that there is insufficient memory bandwidth to satisfy
+the increased demand of memory accesses and  due to hardware limitations.
 
@@
 \centerline{\includegraphics[scale=0.47]{scalability.pdf}}
-\caption{Scalability. a. Running time as a function of the clustering
-  distance. b. Logarithm of the running time versus the logarithm of
-  the number of sequences to be clustered. c. Relative performance increase for
-  different number of parallel threads. d. Running time versus length
-  of the input sequences.
-}\label{fig:X}
+\caption{Scalability. a. Logarithm of the running time versus the
+logarithm of the number of sequences to be clustered. b. Running time as
+a function of the clustering distance. c. Running time versus length of
+the input sequences. d. Relative performance increase for different number
+of parallel threads. 
+}\label{fig:perf}
 \end{figure}
@@
 
-We generated 1 million random 40-mers that were duplicated 47 times.
-For each 40-mer, we inserted 3 addional mutant sequences to the pool
+We generated 1 million random 40-mers duplicated 47 times. For each
+40-mer, we also inserted 3 addional mutant sequences to the pool
 by sampling again 3 nucleotides at random positions. The probability
@@
 cluster thus consists of 50 sequences, of which 3 have a Levenshtein
-distance 3 from the canonical representant.
+distance 3 from the centroid of the cluster (the canonical
+representant).
 
 All software were set to run on a 6-core single-processor Intel Xeon
-E5-2620 system with **64~GB** of DDR3-RAM at 1333~Mhz. The execution
+E5-2620 system with 50~GB of DDR3-RAM at 1333~Mhz. The execution
 options for each software are summarized in
-Table~\ref{tab:soft-options}. Denoted as \emph{sequential R} is the
-sequential algorithm of \cite{pmid23953119} which is implemented in R
-and relies on the \texttt{stringdist} function of the
-\texttt{stringdist} package.
+Table~\ref{tab:soft-options}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%             WHERE TO PUT THIS NOW??                               %
+%Denoted as \emph{sequential R} is the
+%sequential algorithm of \cite{pmid23953119} which is implemented in R
+%and relies on the \texttt{stringdist} function of the
+%\texttt{stringdist} package.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Using **12** cores, starcode clustered the shuffled 50 million sequences
-in 5 minutes and 17 seconds without a single error (the output
-consisted of 1 million clusters of size 50).  The running times
+Using 12 cores, starcode clustered the shuffled 50 million sequences
+in 5 minutes and 3 seconds without a single error (the output
+consisted of 1 million clusters of size 50). There is no obvious
+way to compare exact algorithms with heuristic algorithms such as
+CD-HIT and USEARCH. We decided to allocate the heuristics 6 minutes
+on the same dataset ($\approx$ 20\% additional time) and use the
+parameters giving the best accurary under this constraint.
+However, we could not find a combination of
+parameters such that either of them could terminate in less than
+6 minutes. To simplify the task, we extracted the unique sequences
+from the input file, bringing it down to 4,000,000 and followed
+the same rationale. We could not find any set of parameters such
+that USEARCH could process this input in less than 6 minutes,
+but CD-HIT could perform the task when run as shown in
+Table~\ref{tab:soft-options}.
+
+we decided to 
+allocating them 6 minutes with the same computer resources 
+USEARCH could not be
+run on the same input because
+The running times
 of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
@@
 and short read mappers. \textbf{a} Running time compared to CD-HIT and
-UCLUST on the dataset described in the text. \textbf{b} Error rate
-compared to CD-HIT and UCLUST, expressed as the percentage of clusters
+USEARCH on the dataset described in the text. \textbf{b} Error rate
+compared to CD-HIT and USEARCH, expressed as the percentage of clusters
 failing to be merged. \textbf{c} Running time compared to Bowtie2, BWA
@@
 longer, the problem of identifying abundant inexact matches becomes similar
-to barcode clustering.
+to barcode clustering. We reasoned that starcode could also be used for the
+task of identifying biologically meaningful sequence motifs.
 
 We set up a test based on the meningitis-causing agent
-\textit{Neisseria meningitidis}. The genome of this bacterium is interspersed
-with a frequent 12 bp sequence known as DNA uptake sequence \citep{pmid10673000}. We
-extracted the 12-mers from both orientations of  the 2.19 Mb genome, yielding
-4.39 million 12-mers, consisting of 2.77 million unique sequences. Clustering
-the 12-mers with starcode within a Levenshtein distance of 2 took less
-than 45 seconds with 12 cores. We identified the known DNA uptake sequence of
-\textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant 12-mer.
-This result testifies to the fact that starcode can be used to identify biologically
+\textit{Neisseria meningitidis}. The genome of this bacterium is
+interspersed with a frequent 12 bp sequence known as DNA uptake
+sequence \citep{pmid10673000}. We extracted the 12-mers from both
+orientations of  the 2.19 Mb genome, yielding 4.39 million 12-mers,
+consisting of 2.77 million unique sequences. Clustering the 12-mers
+with starcode within a Levenshtein distance of 2 took less than 45
+seconds with 12 cores. We identified the known DNA uptake sequence
+of \textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant
+12-mer, with 1466 exact and 2096 inexact hits. This result testifies
+to the fact that starcode can be used to identify biologically
 relevant motifs in bacterial genomes.
@@
 some disagreement about the motif that it recognizes \citep{pmid23562324}.
-For each replicate, we replaced the RNAcompete signals by their rank and
+For each SRSF1 replicate of the RNAcompete dataset,
+we replaced the RNAcompete signals by their rank and
 extracted the 10-mers from the microarray probes. The 10-mers were given
 a score equal to the rank of the probe they belong, and enriched motifs
-were found using the sphere clustering of starcode with max Levenshtein
-distance 2. The score of the most enriched 10-mer is thus the sum of the
-ranks of all 10-mers within this distance. Clustering the 6.3 million
-extracted 10-mers with 12 cores took about 20 seconds for each replicate.
-The most enriched 10-mers were AGGACACGGA, AGGACACGGA, AGGACGGAGG,
-AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except for the last replicate, the
-motifs consist of AGGAC and GGA, with a spacer of variable length. This
-suggests that the binding of SRSF1 to RNA is flexible, which explains the
-poor agreement between the motifs derived from 6-mers or 7-mers.
+were found using the sphere clustering of starcode with maximum
+Levenshtein distance 2. The score of the most enriched 10-mer is thus the
+sum of the ranks of all 10-mers within this distance. Clustering the 6.3
+million extracted 10-mers with 12 cores took about 20 seconds for each
+replicate.  The most enriched 10-mers were AGGACACGGA, AGGACACGGA,
+AGGACGGAGG, AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except for the last
+replicate, the motifs consist of AGGAC and GGA, with a spacer of variable
+length. This suggests that the binding of SRSF1 to RNA may be flexible,
+which would explain the disagreement between the motifs derived from 6-mers
+or 7-mers.
 

### 2014-07-03 doc/plot-scalability.R

@@ -3,19 +3,23 @@ X1 = c(0,1,2,3,4,5)
 Y1 = c(0.465, 0.631, 0.776, 19.072, 2*60+20.767, 9*60+13.506)
 X2 = 10000*c(1,2,4,8,16,32,64)
 Y2 =c(0.294, 0.966, 3.374, 11.849, 39.074, 60+55.348, 4*60+55.366)
-X3 = c(1,2,4,6,8,12)
-Y3 = c(40*60+9, 20*60+20, 10*60+41, 7*60+43, 6*60+2, 4*60+32)
+X3 = c(1,2,4,6,8,10,12)
+Y3 = c(8*60+52, 4*60+20, 2*60+14, 1*60+29, 1*60+8, 58, 51)
 X4 = 10*c(1,2,3,4,5,6,7,8)
 Y4 = c(22.072, 25.714, 26.361, 19.865, 5.114, 4.959, 5.591, 6.257)
 # Plot
 pdf("/mnt/shared/src/C/starcode/doc/scalability.pdf")
 par(mfrow=c(2,2))
-plot(log(X2),log(Y2), type='b', xlab="Log number of sequences", ylab="Log Running time (seconds)", cex.lab=1.3)
+plot(log(X2),log(Y2), type='b', xlab="Ln number of sequences", ylab="Ln Running time (seconds)", cex.lab=1.3)
 abline(lm(log(Y2)~log(X2)), col=2)
 text(x=13, y=5.65, "-16.7+1.7x", adj=1)
+mtext("a", adj=-.05, cex=1.5, line=1)
 plot(X1,Y1, type='b', xlab="Levenshtein distance", ylab="Running time (seconds)", cex.lab=1.3)
+mtext("b", adj=-.05, cex=1.5, line=1)
 plot(X4,Y4, type='b', ylim=c(0,30), xlab="Sequence length", ylab="Running time (seconds)", cex.lab=1.3)
+mtext("c", adj=-.05, cex=1.5, line=1)
 plot(X3,Y3[1]/Y3, type='b', ylim=c(1,12), xlim=c(1,12), xlab="Number of threads", ylab="Relative performance", cex.lab=1.3)
 abline(0,1, col=2)
+mtext("d", adj=-.05, cex=1.5, line=1)
 dev.off()
 

### 2014-07-04 doc/document.tex
Corrected some spelling typos.

@@
 running time complexity of starcode is approximately $O(n^{1.7})$.
-Figure~\ref{fig:perf}b shows that the running time gros exponentially
+Figure~\ref{fig:perf}b shows that the running time grows exponentially
 as a function of the maximum Levenshtein distance used for clustering.
@@
 trie-search.
-Finally, in we show the scalability of starcode with increasing
+Finally, we show the scalability of starcode with increasing
 number of treahds in Figure~\ref{fig:perf}d. The search algorithm is fully
@@
 that the input reading and clustering steps are brief but not parallel,
-the second is that there is insufficient memory bandwidth to satisfy
-the increased demand of memory accesses and  due to hardware limitations.
+the second is due to hardware limitations, i.e. there is insufficient
+memory bandwidth to satisfy the increased demand of memory accesses.
 

### 2014-07-05 doc/document.tex
Manuscript ready for submission.

@@
 
-\title[Exact sequence clustering with starcode]{Starcode: an exact algorithm for
-sequence clustering}
+\title[Exact sequence clustering with starcode]{Starcode: an exact
+algorithm for sequence clustering}
 \author[Valera Zorita \textit{et~al}]{Eduard Valera Zorita\,$^{1,2}$, Pol
-Cusc\'o\,$^{1,2}$ and Guillaume Filion\,$^{1,2}$\footnote{to whom correspondence
-should be addressed}}
-\address{$^{1}$Genome Architecture, Gene Regulation, Stem Cells and Cancer Programme,
-Centre for Genomic Regulation (CRG), Dr. Aiguader 88, 08003 Barcelona, Spain.\\
+Cusc\'o\,$^{1,2}$ and Guillaume Filion\,$^{1,2}$\footnote{to whom
+correspondence should be addressed}}
+\address{$^{1}$Genome Architecture, Gene Regulation, Stem Cells and
+Cancer Programme, Centre for Genomic Regulation (CRG), Dr. Aiguader 88,
+08003 Barcelona, Spain.\\
 $^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain.}
@@
 \section{Motivation:}
-The increasing throughput of sequencing technologies offers new applications
-and challenges for computational biology. One such application is the use of
-random barcodes to trace and quantify transcripts or lineages in experimentals
-setups. The high error rate of modern sequencers calls for additional
-post-processing techniques capable of detecting and reverting the misreads.
-However, in the absence of a reference population, the problem amounts to
-performing a pairwise comparison of all the barcodes, which is unfeasible for
-excessive computational complexity.
+The increasing throughput of sequencing technologies offers new
+applications and challenges for computational biology. One such
+application is the use of random barcodes to trace and quantify
+transcripts or lineages in experimental setups. The high error
+rate of modern sequencers calls for additional post-processing
+techniques capable of detecting and reverting the misreads.
+However, in the absence of a reference population, the problem
+amounts to performing a pairwise comparison of all the barcodes,
+which is unfeasible for excessive computationally complexity.
 
 \section{Results:}
-Here we address this problem and describe an exact algorithm to determine which
-pairs of sequences lie within a given Levenshtein distance. The matched pairs
-are merged into clusters represented by a canonical sequence. The effiency of
-starcode is attributable to the poucet search, a novel  implementation of the
-Needleman-Wunsch algorithm performed on the nodes of a trie. Parallelization
-achieves linear scaling of performance on multi-core machines. On the task of
-clustering random barcodes, starcode outperforms short read mappers and sequence
-clustering algorithms in both speed and precision.
+Here we address this problem and describe an exact algorithm to
+determine which pairs of sequences lie within a given Levenshtein
+distance. The matched pairs are merged into clusters represented
+by a canonical sequence. The effiency of starcode is attributable
+to the poucet search, a novel  implementation of the Needleman-Wunsch
+algorithm performed on the nodes of a trie. On the task of clustering
+random barcodes, starcode outperforms sequence clustering algorithms
+in both speed and precision. We further show that starcode can also
+be used to identify enriched motifs in DNA and RNA sequences.
 
@@
 \citep{pmid18809713, pmid23953119}. Sequencing errors will create erroneous
-barcodes that have to be reverted to the original sequence. Hands on experience
-with real datasets show that this step becomes limiting when
+barcodes that have to be reverted to the original sequence. Hands on
+experience with real datasets shows that this step becomes limiting when
 the number of unique sequences is high. In search for a solution to this problem, we
@@
 
-Here we describe the starcode algorithm and we benchmark it against existing
-software that perform related tasks. We show that starcode is both faster and
-more precise than the alternatives, achieving perfect clustering on ideal datasets.
-We further show that starcode is a good candidate to cluster k-mers from large
-datasets from bacterial or even metazoan genomes.
+Here we describe the starcode algorithm and we benchmark it against
+existing software that perform related tasks. We show that starcode is
+both faster and more precise than the alternatives, achieving perfect
+clustering on ideal datasets. We further show that starcode can be used
+to identify enriched motifs in large datasets such as bacterial genomes
+or protein-DNA/protein-RNA interaction experiments.
 
@@
 \caption{Needleman-Wunsch (NW) alignment. \textbf{a}
-  Alignment of GTTGCA and GATCCA using
-the NW alignment matrix. The margins (purple) are initialized and the cells of
-the matrix (yellow) are computed from left to right and from top to bottom by
-the following
-dynamic programming algorithm. The value in each cell is computed from the values
-of the neighboring top, left and top-left cells. Denoting $a$, $b$ and $c$ these
-respective terms, the value in the cell of coordinates $(i,j)$ is computed as
-$\min(a+1, b+1, c)$ if the $i$-th letter from the first sequence is identical to
-the $j$-th letter from the second, or $\min(a+1, b+1, c+1)$ if the letters are
-different. The Levenshtein distance
-between the two sequences is found in the bottom right cell. \textbf{b}
-Lower complexity algorithm to
-determine whether GTTGCA and GATCCA are 2-neighbors. The values in purple cells
-are set during initialization. The dynamic programming algorithm proceeds as above,
-with the difference that it is interrupted if the value of a diagonal cell (bold
-borders) is larger than 2. The values in purple cells are not always the same as
-their equivalent in a (purple arrow), but the values in the yellow cells are
-nevertheless identical. The values of the white cells are never computed, which
-contributes to making the algorithm less complex.}\label{fig:NW}
+Alignment of GTTGCA and GATCCA using the NW alignment matrix. The
+margins (purple) are initialized and the cells of the matrix (yellow)
+are computed from left to right and from top to bottom by the following
+dynamic programming algorithm. The value in each cell is computed from
+the values of the neighboring top, left and top-left cells. Denoting
+$a$, $b$ and $c$ these respective terms, the value in the cell of
+coordinates $(i,j)$ is computed as $\min(a+1, b+1, c)$ if the $i$-th
+letter from the first sequence is identical to the $j$-th letter from
+the second, or $\min(a+1, b+1, c+1)$ if the letters are different. The
+Levenshtein distance between the two sequences is found in the bottom
+right cell. \textbf{b} Lower complexity algorithm to determine whether
+GTTGCA and GATCCA are 2-neighbors. The values in purple cells are set
+during initialization. The dynamic programming algorithm proceeds as
+above, with the difference that it is interrupted if the value of a
+diagonal cell (bold borders) is larger than 2. The values in purple
+cells are not always the same as their equivalent in \textbf{a} (purple
+arrow), but the values in the yellow cells are nevertheless identical.
+The values of the white cells are never computed, which contributes to
+making the algorithm less complex.}\label{fig:NW}
 \end{figure}
 
-In many instances, the only information of interest is to find out whether the
-sequences are $d$-neighbors (separated by a distance less than or equal to a
-threshold $d$).  In that case, the complexity then reduces to
-$O(d \min(m,n))$ as described below \citep{ukkonen}. Instead
-of initializing the margins of the matrix and computing all the
-terms, the matrix is initialized as shown on Figure~\ref{fig:NW}b and only the
-terms around
-the diagonal are computed. If the computation of a diagonal term yields a value
-greater than $d$, the distance is known to be greater than $d$ and the process
-is halted. Otherwise, the bottom-right term indicates the actual distance
-between the sequences, as in the NW algorithm.
+In many instances, the only information of interest is to find out
+whether the sequences are $d$-neighbors (separated by a distance less
+than or equal to a threshold $d$).  In that case, the complexity then
+reduces to $O(d \min(m,n))$ as described below \citep{ukkonen}. Instead
+of initializing the margins of the matrix and computing all the terms,
+the matrix is initialized as shown on Figure~\ref{fig:NW}b and only the
+terms around the diagonal are computed. If the computation of a
+diagonal term yields a value greater than $d$, the distance is known to
+be greater than $d$ and the process is halted. Otherwise, the
+bottom-right term indicates the actual distance between the sequences,
+as in the NW algorithm.
 
-This method can be used for inexact matching of sequences against a reference
-set indexed as prefix tree, also known as a trie \citep{ukkonen}. The terms of the
-matrix are updated row-wise as a depth-first search traverses the trie from the
-root, as illustrated in Figure~\ref{fig:trie}. Every time a node is visited, the row
-corresponding to its depth is recomputed. If the threshold value $d$ is exceeded
-for a diagonal term, the Levenshtein distance for all the downstream sequences is
-also necessarily greater than $d$. Therefore, no more hits are to be discovered
-in this path and the depth-first search backtracks to the parent node. When the
-process halts, every tail node
-(corresponding to a sequence of the database) on the path of this search is a
-$d$-neighbor of the query. This method is efficient because it eliminates large
-areas of the search space, and because the NW alignment of the query with each
-prefix of the database is computed only once.
+This method can be used for inexact matching of sequences against a
+reference set indexed as a prefix tree, also known as a trie
+\citep{ukkonen}. The terms of the matrix are updated row-wise as a
+depth-first search traverses the trie from the root, as illustrated
+in Figure~\ref{fig:trie}. Every time a node is visited, the row
+corresponding to its depth is recomputed. If the threshold value $d$
+is exceeded for a diagonal term, the Levenshtein distance for all the
+downstream sequences is also necessarily greater than $d$. Therefore,
+no more hits are to be discovered in this path and the depth-first
+search backtracks to the parent node. When the process halts, every
+tail node (corresponding to a sequence of the database) on the path
+of this search is a $d$-neighbor of the query. This method is
+efficient because it eliminates large areas of the search space, and
+because the NW alignment of the query with each prefix of the database
+is computed only once.
 
@@
 \centerline{\includegraphics{trie.pdf}}
-\caption{Trie query with NW alignment. Each sequence of the index is a path in
-the trie. The query GTTGCA is written at the top of a NW matrix, which is
-initialized as shown on Figure~\ref{fig:NW}b. The trie is traversed by a
-depth-first search (red path) from the root. At each depth, the node added to
-the path is written on the left of the NW matrix and the row is computed.
-Checkpoints from 1 to 4 (circled red numbers) show the state of the NW matrix
-as the search proceeds. The node labelled 3 is a leaf and thus corresponds to a
-2-neighbor of the query. The search path then backtracks to the node labelled 2
-and the last rows of the NW are erased. The search path then goes to node labelled
-4, in which case the newly computed diagonal cell exceeds the threshold (circled
-in red). Even if this node has children, they are not visited (red crosses).}
+\caption{Trie query with NW alignment. Each sequence of the index is a
+path in the trie. The query GTTGCA is written at the top of a NW matrix,
+which is initialized as shown on Figure~\ref{fig:NW}b. The trie is
+traversed by a depth-first search (red path) from the root. At each
+depth, the node added to the path is written on the left of the NW
+matrix and the row is computed. Checkpoints from 1 to 4 (circled red
+numbers) show the state of the NW matrix as the search proceeds. The
+node labeled 3 is a leaf and thus corresponds to a 2-neighbor of the
+query. The search path then backtracks to the node labeled 2
+and the last rows of the NW are erased. The search path then goes to
+the node labeled 4, in which case the newly computed diagonal cell
+exceeds the threshold (circled in red). Even if this node has children,
+they are not visited (red crosses).}
 \label{fig:trie}
@@
 the lookup table $2(i-1)+1$ times to cover all the possible
-cases. Hence, the last word will be queried $2\tau+1$ times, including
+cases. The last word will be queried $2\tau+1$ times, including
 $\tau$ shifts on the right and $\tau$ shifts on the left. For the
 latter shift we make use of the $\tau$ nucleotides stored in the
-tail. The total number of lookups per sequence for a given $\tau$ is
-$(\tau+1)^2$.
+tail. The total number of lookups per sequence is $(\tau+1)^2$.
 
@@
 average depth in a trie that fans out exponentially.
-The running time first increases as a function of the sequence length,
+As a function of the sequence length, the running time first increases 
 but then plummets and stays low (Figure~\ref{fig:perf}c).
@@
 Finally, we show the scalability of starcode with increasing
-number of treahds in Figure~\ref{fig:perf}d. The search algorithm is fully
+number of threads in Figure~\ref{fig:perf}d. The search algorithm is fully
 parallel and the relative performance increases linearly up to 8 threads.
@@
 that the input reading and clustering steps are brief but not parallel,
-the second is due to hardware limitations, i.e. there is insufficient
-memory bandwidth to satisfy the increased demand of memory accesses.
+the second is that there is insufficient memory bandwidth to satisfy
+the increased demand of memory accesses due to hardware limitations.
 
@@
 We generated 1 million random 40-mers duplicated 47 times. For each
-40-mer, we also inserted 3 addional mutant sequences to the pool
+40-mer, we also inserted 3 additional mutant sequences to the pool
 by sampling again 3 nucleotides at random positions. The probability
@@
 cluster thus consists of 50 sequences, of which 3 have a Levenshtein
-distance 3 from the centroid of the cluster (the canonical
-representant).
+distance 3 from the canonical representant (referred to as the centroid
+in clustering terms).
 
 All software were set to run on a 6-core single-processor Intel Xeon
-E5-2620 system with 50~GB of DDR3-RAM at 1333~Mhz. The execution
-options for each software are summarized in
-Table~\ref{tab:soft-options}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%             WHERE TO PUT THIS NOW??                               %
-%Denoted as \emph{sequential R} is the
-%sequential algorithm of \cite{pmid23953119} which is implemented in R
-%and relies on the \texttt{stringdist} function of the
-%\texttt{stringdist} package.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}[h]
-\centering
-\caption{Software execution parameters}
-\resizebox{\columnwidth}{!}{
-\begin{tabular}{|l|l|}
-    \hline
-    starcode & \texttt{-t 20}\\ \hline
-    CD-HIT 4.6.1 & \texttt{-c 0.925 -T 20 -M 0}\\ \hline
-    USEARCH 7.0.1090  & \texttt{-threads 20 -cluster\_fast -id 0.925}\\ \hline
-        Bowtie2 indexer & \texttt{bowtie2-build --quiet}\\ \hline
-    Bowtie2 2.1.0 & \texttt{-f -a -p 20 --no-hd --very-sensitive}\\ \hline
-        BWA indexer & \texttt{bwa index -a is}\\ \hline
-    BWA 0.7.9a & \texttt{mem -t 20 -a -k1 -B0} \\ \hline
-    GEM indexer & \texttt{gem-indexer -T 12} \\ \hline
-    GEM 1.423 & \texttt{-e3 -s4 -T12 --granularity 100000}\\ \hline
-    \emph{sequential R} & \texttt{method='lv', maxDist=2}\\ \hline
-\end{tabular}
-}
-\label{tab:soft-options}
-\end{table}
-
-Using 12 cores, starcode clustered the shuffled 50 million sequences
+E5-2620 system with 48~GB of DDR3-RAM at 1333~Mhz.
+Using 12 threads, starcode clustered the shuffled 50 million sequences
 in 5 minutes and 3 seconds without a single error (the output
@@
 CD-HIT and USEARCH. We decided to allocate the heuristics 6 minutes
-on the same dataset ($\approx$ 20\% additional time) and use the
+on the same dataset (20\% additional time) and use the
 parameters giving the best accurary under this constraint.
@@
 6 minutes. To simplify the task, we extracted the unique sequences
-from the input file, bringing it down to 4,000,000 and followed
+from the input file, bringing the number down to 4,000,000 and followed
 the same rationale. We could not find any set of parameters such
-that USEARCH could process this input in less than 6 minutes,
-but CD-HIT could perform the task when run as shown in
-Table~\ref{tab:soft-options}.
+that USEARCH could process this input in less than 6 minutes
+(because of poor multi-threading performance), but CD-HIT could
+perform the task when run as shown in Table~\ref{tab:params1}.
+In such conditions, it identified 2,798,656 clusters, indicating that
+the clustering was not nearly complete.
 
-we decided to 
-allocating them 6 minutes with the same computer resources 
-USEARCH could not be
-run on the same input because
-The running times
-of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
-identified 2,362,352 and 2,739,954 clusters respectively. The results
-are summarized in Figure~\ref{fig:benchmark}. Both algorithms rely on
-heuristics, explaining why the number of clusters is off target. These
-results clearly demonstrate that starcode outperforms other clustering
-algorithms both in speed and accuracy with random input sequences.
+\begin{table}[h]
+\centering
+\caption{Software execution parameters (benchmark 1)}
+\resizebox{\columnwidth}{!}{
+\begin{tabular}{|l|l|}
+    \hline
+    starcode & \texttt{-d3 -t12 -s}\\ \hline
+    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 7 -c .925 -T 12 -M 0 -r 0 -A 40}\\ \hline
+\end{tabular}
+}
+\label{tab:params1}
+\end{table}
 
-We also compared starcode on the biological benchmark dataset proposed
-by \cite{pmid23060610}.
+We next benchmarked starcode on the biological dataset used by
+\cite{pmid23060610}. Because USEARCH imposes a memory limit on the
+size of the input file, we clustered the forward reads on 95\%
+identity from one lane only (reference ERR011089), adding up to
+4,189,859 reads, each 75 nucleotides long. Using 12 threads, starcode
+identified 4,111,111 clusters in 6 minute 19 seconds. Allocating up to
+7 minutes 30 seconds to CD-HIT and USEARCH, we selected the parameters
+giving highest accuracy. In such conditions (see Table~\ref{tab:params2})
+CD-HIT identified 4,126,352 clusters while USEARCH could not be run in
+the allocated time. These results indicate that starcode achieves good
+performance on clustering high throughput sequencing reads.
+
+\begin{table}[h]
+\centering
+\caption{Software execution parameters (benchmark 2)}
+\resizebox{\columnwidth}{!}{
+\begin{tabular}{|l|l|}
+    \hline
+    starcode & \texttt{-d4 -t12 -s}\\ \hline
+    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 8 -c .95 -T 12 -M 0 -r 0 -A 75}\\ \hline
+\end{tabular}
+}
+\label{tab:params2}
+\end{table}
 
@@
 \centerline{\includegraphics{benchmark.pdf}}
-\caption{Benchmark of starcode against sequence clustering algorithms
-and short read mappers. \textbf{a} Running time compared to CD-HIT and
-USEARCH on the dataset described in the text. \textbf{b} Error rate
-compared to CD-HIT and USEARCH, expressed as the percentage of clusters
-failing to be merged. \textbf{c} Running time compared to Bowtie2, BWA
-and GEM on this dataset. The running time is decomposed in extraction
-of unique sequences (purple), Burrows-Wheeler indexing (yellow), and
-query proper (white). The extraction was performed with the Linux sort
-command, so the time is the same in every case. \textbf{d} Error rate
-compared to Bowtie2, BWA and GEM, expressed as the proportion of matches
-with a single hit. This is an underestimate of the true error rate.}
+\caption{Benchmark of starcode against short read mappers.
+\textbf{a} Running time compared to Bowtie2, BWA and GEM on the
+dataset described in the text. The running time is decomposed as
+extraction of unique sequences (purple), Burrows-Wheeler indexing
+(yellow) and query proper (white). The extraction was performed with
+the Linux \texttt{sort} command, so the time is the same in every
+case. \textbf{d} Error rate compared to Bowtie2, BWA and GEM,
+expressed as the proportion of matches with a single hit. This is an
+underestimate of the true error rate.}
 \label{fig:benchmark}
@@
 \subsection{Benchmark against short read mappers}
-A potential strategy to cluster short sequences is to use last generation short
-read mappers to match a set of sequences against itself. Once matches are
-available, several community detection algorithms can be used to identify the
-clusters. We benchmarked starcode against the short read mappers Bowtie2
-\citep{pmid22388286}, BWA \citep{pmid19451168} and GEM \citep{pmid23103880}.
-Because short read
-mappers do not perform clustering, we only evaluated their performance on the
-matching problem.
+A potential strategy to cluster short sequences is to use
+Burrows-Wheeler transform-based short read mappers to match a set of
+sequences against itself. Once matches are available, several community
+detection algorithms can be used to identify the clusters. We 
+benchmarked starcode against the short read mappers Bowtie2
+\citep{pmid22388286}, BWA \citep{pmid19451168} and GEM
+\citep{pmid23103880}. Short read mappers do not perform clustering, so
+we only evaluated their performance on the matching problem.
 
-For this test, we used the same dataset described above, in which every
-sequence has at least 2 distinct matches at a distance 3 (mutant sequences
-from the same cluster may be at a distance up to 6 from each other). This
-means that we
-can use matches with a single hit (themselves) as a lower bound on the inaccuracy.
-We used unique sequences to build an index and queried the index with the same
-sequences. Figure~\ref{fig:benchmark}c shows the running time of starcode along with the
-decomposition of the running time for the three mappers. Note that the time to
-extract unique sequences is always the same because we used the Linux
-\texttt{sort -u} command in every case. The running time of
-starcode is slightly lower than short read mappers when all steps ares
-taken into account.  Figure~\ref{fig:benchmark}d shows the lower bound on
-the error rate, as estimated by the proportion of single hits. BWA was
-found to be the most accurate aligner with a lower bound of 31\% sequences.
-with a single match. This measure is an underestimate of the true error
-rate, but given that starcode is an exact algorithm, a more accurate
-measure is not necessary. In conclusion, starcode can compete with the
+For this test, we used the first dataset described above, in which
+every sequence has at least 2 distinct matches at a distance 3 (mutant
+sequences from the same cluster may be at a distance up to 6 from each
+other, but their distance to the centroid is always 3). This means that
+we can use matches with a single hit (every read trivially matches
+itself) as a lower bound on the inaccuracy. We used the 4,000,000 unique
+40-mers to build an index that was later queried with the same 40-mers
+using parameters shown in Table~\ref{tab:srm}.
+
+\begin{table}[h]
+\centering
+\caption{Software execution parameters (benchmark 3)}
+\resizebox{\columnwidth}{!}{
+\begin{tabular}{|l|l|}
+    \hline
+    Bowtie2 indexer & \texttt{bowtie2-build --quiet}\\ \hline
+    Bowtie2 2.1.0 & \texttt{-f -a -p 12 --no-hd --very-sensitive}\\ \hline
+    BWA indexer & \texttt{bwa index -a is}\\ \hline
+    BWA 0.7.9a & \texttt{mem -t 12 -a -k1 -B0} \\ \hline
+    GEM indexer & \texttt{gem-indexer -T 12} \\ \hline
+    GEM 1.423 & \texttt{-e3 -s4 -T12 --granularity 100000}\\ \hline
+\end{tabular}
+}
+\label{tab:srm}
+\end{table}
+
+Figure~\ref{fig:benchmark}a shows the running time of starcode along
+with the decomposition of the running time for the three mappers. Note
+that the time to extract unique sequences is always the same because
+we used the Linux \texttt{sort -u} command in every case. The running
+time of starcode is slightly lower than short read mappers when all
+steps ares taken into account. Figure~\ref{fig:benchmark}b shows the
+lower bound on the error rate, as estimated by the proportion of single
+hits. BWA was found to be the most accurate aligner with a lower bound
+of 31\% sequences with a single match. This measure is an underestimate
+of the true error rate, but given that starcode is an exact algorithm,
+a more accurate measure is not necessary: starcode can compete with the
 short read aligners on the matching problem, and it provides a precision
@@
 In the course of setting up the TRIP technology in our laboratory
-\citep{pmid23953119}, we
-realized the need to develop efficient algorithms to cluster similar sequences.
-Briefly, the principle of TRIP (Thousands of Reporters Integrated in Parallel)
-is to tag reporter transcripts with random barcodes and measure the abundance
-of barcodes in the RNA as a proxy for gene expression. There is no reference to
-match aberrant barcodes against, because the tagging sequences are unknown.
-Instead, barcodes are matched against each other and clustered by similarity to
-infer canonical sequences.
+\citep{pmid23953119}, we realized the need to develop efficient
+algorithms to cluster similar sequences. Briefly, the principle of TRIP
+(Thousands of Reporters Integrated in Parallel) is to tag reporter
+transcripts with random barcodes and measure the abundance of barcodes
+in the RNA as a proxy for gene expression. There is no reference to
+match aberrant barcodes against, because the tagging sequences are
+unknown. Instead, barcodes are matched against each other and clustered
+by similarity to infer canonical sequences.
 
 We tested the efficiency of starcode on the TRIP dataset from
-\cite{pmid23953119}. In the experiment labelled mPGKA, we identified
-24.1 million (91\%) barcode-containing reads out of 26.6 million, consisting
-of approximately 223,000 and 220,000 unique barcode sequences for PCR replicates
-1 and 2, respectively. Following the authors, we kept only the barcodes with
-at least 5 counts and performed clustering as described: ``First we sorted
-barcodes according to their counts. Then, for each barcode (starting from the
-most frequent one), we identified and removed all its mutant versions, defined
-as barcodes within a Hamming distance of 2.'' This method is here on referred
-to as the ``sequential algorithm''. When replacing the Hamming distance by the
-Levenshtein distance, starcode produced exactly the same output as the sequential
+\cite{pmid23953119}. In the experiment labeled mPGKA, we identified
+24.1 million (91\%) barcode-containing reads out of 26.6 million,
+consisting of approximately 223,000 and 220,000 unique barcode sequences
+for PCR replicates 1 and 2, respectively. Following the authors, we kept
+only the barcodes with at least 5 counts and performed clustering as
+described: ``First we sorted barcodes according to their counts. Then,
+for each barcode (starting from the most frequent one), we identified
+and removed all its mutant versions, defined as barcodes within a Hamming
+distance of 2.'' We implemented this method, here on referred to as the
+``sequential algorithm'' with the function call
+\texttt{stringDist(method='lv', maxDist=2)} from R the package
+\texttt{stringDist} \citep{R}.  When replacing the Hamming distance by
+the Levenshtein
+distance, starcode produced exactly the same output as the sequential
 algorithm. The running time of starcode averaged over the replicates was
-2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which represents a
-75-fold speedup.
+2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which
+represents a 75-fold speedup.
 
-The performance of the sequential algorithm relies on the arbitrary exclusion
-of reads with less than 5 counts, the main purpose of which is to reduce the
-computational burden. When all barcodes were kept, the average running time of
-starcode was 10.27 seconds, \textit{versus} $>$ 6 hours for the sequential
-algorithm. Using the Hamming distance, as the authors originally suggested,
-decreased the average running time of the sequential algorithm to about 3 hours
-and 30 minutes (and
+The performance of the sequential algorithm relies on the arbitrary
+exclusion of reads with less than 5 counts, the main purpose of which
+is to reduce the computational burden. When all barcodes were kept,
+the average running time of starcode was 10.27 seconds, \textit{versus}
+$>$ 6 hours for the sequential algorithm. Using the Hamming distance,
+as the authors originally suggested, decreased the average running time
+of the sequential algorithm to about 3 hours and 30 minutes (and
 the output differed from that of starcode). Note that in all the cases
-mentioned above, starcode was run with a single core to compare the algorithms
-based on similar computer resources. In conclusion, the barcode clustering
-problem can be simplified by various tricks, but starcode brings down the
-running time to nearly instantaneous, and thereby obviates the need for such
-arbitrary heuristics.
+mentioned above, starcode was run with a single core to compare the
+algorithms based on similar computer resources. In conclusion, the
+barcode clustering problem can be simplified by various tricks, but
+starcode brings down the running time to nearly instantaneous, and
+thereby obviates the need for such arbitrary heuristics.
 
 \subsection{Identifying enriched sequence motifs}
-Sequence motifs are thought to play an important role in DNA metabolism. Key
-regulators, such as transcription factors, nucleosomes and non coding RNAs have
-sequence preferences targeting them to the sites where they act. Identifying
-those sequences is a way to pinpoint the regulators and the mechanisms they
-are involved in. However, the sequence motifs are not strictly identical
-at different sites, hence they are better identified by inexact matching.
-This problem becomes computationally difficult for long motifs (above 12-13
-nucleotides) because of the combinatorial scaling. But as motifs become
-longer, the problem of identifying abundant inexact matches becomes similar
-to barcode clustering. We reasoned that starcode could also be used for the
-task of identifying biologically meaningful sequence motifs.
+Sequence motifs are thought to play an important role in DNA metabolism.
+Key regulators, such as transcription factors, nucleosomes and non
+coding RNAs have sequence preferences targeting them to the sites where
+they act. Identifying those sequences is a way to pinpoint the
+regulators and the mechanisms they are involved in. However, the
+sequence motifs are not strictly identical at different sites, hence
+they are better identified by inexact matching. This problem becomes
+computationally difficult for long motifs (above 12-13 nucleotides)
+because of the combinatorial scaling. But as motifs become longer, the
+problem of identifying abundant inexact matches becomes similar to
+barcode clustering. We reasoned that starcode could also be used for
+the task of identifying biologically meaningful sequence motifs.
 
@@
 with starcode within a Levenshtein distance of 2 took less than 45
-seconds with 12 cores. We identified the known DNA uptake sequence
+seconds with 12 threads. We identified the known DNA uptake sequence
 of \textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant
@@
 To test starcode on another application, we used the RNA-protein
-interaction data produced by RNAcompete \citep{pmid19561594}. The mammalian
-splicing factor SRSF1 is known to bind RNA GA-rich motifs, but there is
-some disagreement about the motif that it recognizes \citep{pmid23562324}.
-For each SRSF1 replicate of the RNAcompete dataset,
-we replaced the RNAcompete signals by their rank and
-extracted the 10-mers from the microarray probes. The 10-mers were given
-a score equal to the rank of the probe they belong, and enriched motifs
-were found using the sphere clustering of starcode with maximum
-Levenshtein distance 2. The score of the most enriched 10-mer is thus the
-sum of the ranks of all 10-mers within this distance. Clustering the 6.3
-million extracted 10-mers with 12 cores took about 20 seconds for each
-replicate.  The most enriched 10-mers were AGGACACGGA, AGGACACGGA,
-AGGACGGAGG, AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except for the last
-replicate, the motifs consist of AGGAC and GGA, with a spacer of variable
-length. This suggests that the binding of SRSF1 to RNA may be flexible,
-which would explain the disagreement between the motifs derived from 6-mers
-or 7-mers.
+interaction data produced by RNAcompete \citep{pmid19561594}. The
+mammalian splicing factor SRSF1 is known to bind RNA GA-rich motifs,
+but there is some disagreement about the motif that it recognizes
+\citep{pmid23562324}. For each replicate of the human SRSF1 in the
+RNAcompete dataset, we replaced the microarray signals by their rank
+and extracted the 10-mers from the microarray probes. The 10-mers were
+given a score equal to the rank of the probe they belong, and enriched
+motifs were found using the sphere clustering of starcode with maximum
+Levenshtein distance 2. The score of the most enriched 10-mer is thus
+the sum of the ranks of all 10-mers within this distance. Clustering
+the 6.3 million extracted 10-mers with 12 threads took about 20 seconds
+for each replicate.  The most enriched 10-mers were AGGACACGGA,
+AGGACACGGA, AGGACGGAGG, AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except
+for the last replicate, the motifs consist of AGGAC and GGA, with a
+spacer of variable length. This suggests that the binding of SRSF1 to
+RNA may involve a spacer sequence, which would explain the disagreement
+between the motifs derived from 6-mers or 7-mers.
 
 \section{Discussion and conclusion}
-Through the parallel poucet search algorithm, starcode implements an exact
-sequence clustering algorithm that is even faster than widely used heuristics.
-By design, starcode is tailored to process high throughput sequencing data on
-multi-core platforms. Our benchmark shows that starcode is superior in speed
-and accuracy compared to other methods on the problem of clustering short random
-sequences. Part of the reason is that such task is stretching the tested
-software far from their initial design. Sequence clustering algorithms are
-meant to cluster long biological sequences, while short read mappers are
-meant to map reads on potentially large genomes. In this respect, starcode
-fills a need arising from the development of barcoding technologies.
+Through the parallel poucet search algorithm, starcode implements an
+exact sequence clustering algorithm that can be faster than popular
+heuristics. By design, starcode is tailored to process high throughput
+sequencing data on multi-core platforms. Our benchmark shows that
+starcode achieves perfect clustering on short random sequences in less
+time than what is considered acceptable for heuristic searches. We also
+show that starcode outperforms next generation read mappers in this
+context. The software tools used for this benchmark are usually not
+used for barcode or random sequence clustering. In this respect,
+starcode fills a need arising from the development of barcoding
+technologies.
 
-The speed of starcode also makes it proficient at more general clustering
-problems, such as identifying enriched k-mers in genomes and in experimental
-data. Here we have given two examples of such applications. In the first, we
-recover a known enriched 12-mer in the genome of \textit{Neisseria meningitidis}.
+The speed of starcode also makes it useful for other clustering
+tasks, such as identifying enriched motifs in microbial genomes and
+in experimental data. Here we have given two examples of such
+applications. In the first, we recover a known enriched 12-mer in the
+genome of \textit{Neisseria meningitidis}. In the second, we recover
+the motif of the human RNA binding protein SRSF1 and notice that it
+seems to consist of two halves separated by a linker. This hypothesis
+is consistent with the fact that SRSF1 binds RNA through two
+consecutive RNA Recognition Motifs (RRM) that are known to bind
+3-4 nucleotides in a row\citep{pmid23253355}. The Levenshtein distance,
+which incorporates insertions and deletions is more likely to capture
+bi-partite binding motifs than position weight matrix representations.
+The use of a clustering method to tackle this problem is unusual, but it 
+it illustrates the potential advantages of distance-based approaches.
 
-The current version of starcode has been mildly optimized for memory
-consumption. The memory footprint depends on the number of sequences to
-cluster (because the sequences of the input set are loaded in memory as a
-group of tries) and on the mean number of matches per sequence. Every match
-has to be stored until the clustering phase, which may represent a large
-memory overhead. As counterintuitive as it may seem, long queries will usually
-impose a lower memory footprint because the matches between sequences are less
-frequent.
+The current version of starcode has been primarily optimized for speed.
+The memory footprint depends on the number of sequences to cluster
+(because the sequences of the input set are loaded in memory as a set of
+of tries) and on the mean number of matches per sequence. Every match is
+stored in memory until the clustering phase, which may represent a large
+overhead if the dataset is dense. As counterintuitive as it may seem,
+long queries will usually impose a lower memory footprint because the
+matches between the sequences are more sparse. We have shown that the
+running time will also be shorter thanks to the lookup table
+search (Figure~\ref{fig:perf}c).
 
-The speed of starcode could also benefit other applications, such as searching
-shared subsequences during genome assembly. However, starcode supports only global
-and not local alignment. A workaround is to compare k-mers as we have done
-in the examples above, but this approach is expected to be memory demanding.
-Even though extending the starcode algorithm to support clustering of protein
-sequences is straightforward, it is not expected to meet the performance level
-of the presented algorithm. On the one hand, there are more letters in the protein
-alphabet, so the shared prefix between consecutive queries is inherently
-shorter on average, which defeats the poucet search. On the other hand, if
-a threshold distance of the order of 1 per 10 nucleotides is sufficient to
-identify similar DNA sequences, a distance of 1 per 10 residues is
-unrealistically low for protein sequences, and this parameter critically
-determines the speed of starcode. More generally, storing computational
-intermediates for shared prefixes could find some applications in other
-algorithms such as short read mappers. The idea of the poucet search seems
-simple in retrospect, but it is a powerful way to tap into the data
-structuration provided by string sorting.
+Perhaps the most surprising element of this study is that an exact
+algorithm can compete with extremely fast heuristics. This will no
+longer be true when clustering divergent sequences because the Levenshtein
+distance will have to be increased, leading to exponentially longer
+running times (Figure~\ref{fig:perf}b). However, for the imporant
+practical case that the divergence is driven by sequencing errors,
+starcode illustrates that there is still room for algorithmic innovations
+that can outperform heuristics. The idea of the poucet search seems
+simple in retrospect, yet it is a powerful way to tap into the data
+structuration provided by string sorting. This principle could find some
+applications in algorithms used in various fields.
 
@@
 \section*{Acknowledgement}
-We would like to thank...
+We would like to thank Maria Chatzou for her precious feedback on the
+preliminary version of this manuscript.
 

### 2014-07-08 main-starcode.c
Added LICENSE and README files.

@@ -1,3 +1,28 @@
+/*
+** Copyright 2014 Guillaume Filion, Eduard Valera Zorita and Pol Cusco.
+**
+** File authors:
+**  Guillaume Filion     (guillaume.filion@gmail.com)
+**  Eduard Valera Zorita (ezorita@mit.edu)
+**
+** Last modified: July 8, 2014
+**
+** License: 
+**  This program is free software: you can redistribute it and/or modify
+**  it under the terms of the GNU General Public License as published by
+**  the Free Software Foundation, either version 3 of the License, or
+**  (at your option) any later version.
+**
+**  This program is distributed in the hope that it will be useful,
+**  but WITHOUT ANY WARRANTY; without even the implied warranty of
+**  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+**  GNU General Public License for more details.
+**
+**  You should have received a copy of the GNU General Public License
+**  along with this program.  If not, see <http://www.gnu.org/licenses/>.
+**
+*/
+
 #include "starcode.h"
 
 char *USAGE = "Usage:\n"

### 2014-07-08 starcode.c

@@ -1,3 +1,28 @@
+/*
+** Copyright 2014 Guillaume Filion, Eduard Valera Zorita and Pol Cusco.
+**
+** File authors:
+**  Guillaume Filion     (guillaume.filion@gmail.com)
+**  Eduard Valera Zorita (ezorita@mit.edu)
+**
+** Last modified: July 8, 2014
+**
+** License: 
+**  This program is free software: you can redistribute it and/or modify
+**  it under the terms of the GNU General Public License as published by
+**  the Free Software Foundation, either version 3 of the License, or
+**  (at your option) any later version.
+**
+**  This program is distributed in the hope that it will be useful,
+**  but WITHOUT ANY WARRANTY; without even the implied warranty of
+**  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+**  GNU General Public License for more details.
+**
+**  You should have received a copy of the GNU General Public License
+**  along with this program.  If not, see <http://www.gnu.org/licenses/>.
+**
+*/
+
 #include "starcode.h"
 #include "trie.h"
 
@@ -55,9 +80,6 @@ starcode
       ntries = 1;
       maxthreads = 1;
    }
-   // XXX The number of tries must be odd otherwise the     XXX
-   // XXX scheduler will make mistakes. So, just in case... XXX
-   if (ntries % 2 == 0) abort();
  
    // Pad sequences. (And return the median size)
    int med = -1;
@@ -122,7 +144,10 @@ run_plan
             mtjob_t *job = mttrie->jobs + mttrie->currentjob++;
             pthread_t thread;
             // Start job and detach thread.
-            if (pthread_create(&thread, NULL, do_query, job)) abort();
+            if (pthread_create(&thread, NULL, do_query, job)) {
+               fprintf(stderr, "error creating thread (run_plan): %s\n", strerror(errno));
+               abort();
+            }
             pthread_detach(thread);
             if (verbose) {
                fprintf(stderr, "starcode progress: %.2f%% \r",
@@ -163,7 +188,7 @@ do_query
    // Create local hit stack.
    gstack_t **hits = new_tower(tau+1);
    if (hits == NULL) {
-      fprintf(stderr, "error: cannot create hit stack (%d)\n",
+      fprintf(stderr, "error: cannot create hit stack (%d).\n",
[…]

### 2014-07-08 starcode.h

@@ -1,3 +1,28 @@
+/*
+** Copyright 2014 Guillaume Filion, Eduard Valera Zorita and Pol Cusco.
+**
+** File authors:
+**  Guillaume Filion     (guillaume.filion@gmail.com)
+**  Eduard Valera Zorita (ezorita@mit.edu)
+**
+** Last modified: July 8, 2014
+**
+** License: 
+**  This program is free software: you can redistribute it and/or modify
+**  it under the terms of the GNU General Public License as published by
+**  the Free Software Foundation, either version 3 of the License, or
+**  (at your option) any later version.
+**
+**  This program is distributed in the hope that it will be useful,
+**  but WITHOUT ANY WARRANTY; without even the implied warranty of
+**  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+**  GNU General Public License for more details.
+**
+**  You should have received a copy of the GNU General Public License
+**  along with this program.  If not, see <http://www.gnu.org/licenses/>.
+**
+*/
+
 #define _GNU_SOURCE
 #include <execinfo.h>
 #include <signal.h>

### 2014-07-08 trie.c

@@ -1,3 +1,28 @@
+/*
+** Copyright 2014 Guillaume Filion, Eduard Valera Zorita and Pol Cusco.
+**
+** File authors:
+**  Guillaume Filion     (guillaume.filion@gmail.com)
+**  Eduard Valera Zorita (ezorita@mit.edu)
+**
+** Last modified: July 8, 2014
+**
+** License: 
+**  This program is free software: you can redistribute it and/or modify
+**  it under the terms of the GNU General Public License as published by
+**  the Free Software Foundation, either version 3 of the License, or
+**  (at your option) any later version.
+**
+**  This program is distributed in the hope that it will be useful,
+**  but WITHOUT ANY WARRANTY; without even the implied warranty of
+**  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+**  GNU General Public License for more details.
+**
+**  You should have received a copy of the GNU General Public License
+**  along with this program.  If not, see <http://www.gnu.org/licenses/>.
+**
+*/
+
 #include "trie.h"
 
 #define min(a,b) (((a) < (b)) ? (a) : (b))

### 2014-07-08 trie.h

@@ -1,3 +1,28 @@
+/*
+** Copyright 2014 Guillaume Filion, Eduard Valera Zorita and Pol Cusco.
+**
+** File authors:
+**  Guillaume Filion     (guillaume.filion@gmail.com)
+**  Eduard Valera Zorita (ezorita@mit.edu)
+**
+** Last modified: July 8, 2014
+**
+** License: 
+**  This program is free software: you can redistribute it and/or modify
+**  it under the terms of the GNU General Public License as published by
+**  the Free Software Foundation, either version 3 of the License, or
+**  (at your option) any later version.
+**
+**  This program is distributed in the hope that it will be useful,
+**  but WITHOUT ANY WARRANTY; without even the implied warranty of
+**  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+**  GNU General Public License for more details.
+**
+**  You should have received a copy of the GNU General Public License
+**  along with this program.  If not, see <http://www.gnu.org/licenses/>.
+**
+*/
+
 #include <stdlib.h>
 #include <string.h>
 #include <stdio.h>

### 2014-07-08 view.c

@@ -1,3 +1,27 @@
+/*
+** Copyright 2014 Guillaume Filion, Eduard Valera Zorita and Pol Cusco.
+**
+** File authors:
+**  Pol Cusco (pcusco@gmail.com)
+**
+** Last modified: July 8, 2014
+**
+** License: 
+**  This program is free software: you can redistribute it and/or modify
+**  it under the terms of the GNU General Public License as published by
+**  the Free Software Foundation, either version 3 of the License, or
+**  (at your option) any later version.
+**
+**  This program is distributed in the hope that it will be useful,
+**  but WITHOUT ANY WARRANTY; without even the implied warranty of
+**  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+**  GNU General Public License for more details.
+**
+**  You should have received a copy of the GNU General Public License
+**  along with this program.  If not, see <http://www.gnu.org/licenses/>.
+**
+*/
+
 #include <stdio.h>
 #include <stdlib.h>
 #include <math.h>

### 2014-07-08 src/main-starcode.c
New dir struct. Removed old files. Substituted error messages by strerror.

@@ -66,8 +66,8 @@ main(
    int threads_flag = -1;
    int cluster_flag = -1;
    // Unset options (value 'UNSET').
-   char * const UNSET;
-   char *format_option = UNSET;
+   char * const UNSET = "unset";
+//   char *format_option = UNSET;
    char *input = UNSET;
    char *output = UNSET;
 

### 2014-07-08 src/starcode.c

@@ -24,34 +24,6 @@
 */
 
 #include "starcode.h"
-#include "trie.h"
-
-#define str(a) (char *)(a)
-#define min(a,b) (((a) < (b)) ? (a) : (b))
-#define max(a,b) (((a) > (b)) ? (a) : (b))
-
-lookup_t *new_lookup(int, int, int, int);
-int lut_search(lookup_t *, useq_t *);
-void lut_insert(lookup_t *, useq_t *);
-int seq2id(char *, int);
-useq_t *new_useq(int, char*);
-gstack_t *read_file(FILE*);
-void destroy_useq(useq_t*);
-void addmatch(useq_t*, useq_t*, int, int);
-void transfer_counts_and_update_canonicals(useq_t*);
-void unpad_useq (gstack_t*);
-int pad_useq(gstack_t*, int*);
-int canonical_order(const void*, const void*);
-int count_order(const void *a, const void *b);
-void *do_query(void*);
-void run_plan(mtplan_t*, int, int);
-mtplan_t *plan_mt(int, int, int, int, gstack_t*, const int);
-void message_passing_clustering(gstack_t*, int);
-void sphere_clustering(gstack_t*, int);
-void * _mergesort(void *);
-int seqsort(void **, int, int (*)(const void*, const void*), int);
-long count_trie_nodes(useq_t **, int, int);
-int AtoZ(const void *, const void *);
 
 FILE *OUTPUT = NULL;
 
@@ -188,8 +160,7 @@ do_query
    // Create local hit stack.
    gstack_t **hits = new_tower(tau+1);
    if (hits == NULL) {
-      fprintf(stderr, "error: cannot create hit stack (%d).\n",
-            check_trie_error_and_reset());
+      fprintf(stderr, "error creating hit stack (do_query): %s\n", strerror(errno));
       abort();
    }
 
@@ -207,8 +178,7 @@ do_query
          lut_insert(lut, query);
          data = insert_string_wo_malloc(trie, query->seq, &node_pos);
          if (data == NULL || *data != NULL) {
-            fprintf(stderr, "error: cannot build trie (%d).\n",
-                  check_trie_error_and_reset());
+            fprintf(stderr, "error building trie (do_query): %s.\n", strerror(errno));
             abort();
          }
       }
@@ -235,7 +205,7 @@ do_query
          for (int j = 0 ; hits[j] != TOWER_TOP ; j++) hits[j]->nitems = 0;
[…]

### 2014-07-08 src/starcode.h

@@ -52,6 +52,10 @@
 #define STRATEGY_EQUAL  1
 #define STRATEGY_PREFIX  99
 
+#define str(a) (char *)(a)
+#define min(a,b) (((a) < (b)) ? (a) : (b))
+#define max(a,b) (((a) > (b)) ? (a) : (b))
+
 #if !defined( __GNUC__) || defined(__APPLE__)
    ssize_t getline(char **lineptr, size_t *n, FILE *stream);
 #endif
@@ -136,5 +140,28 @@ int tquery(FILE*, FILE*, FILE*, const int, const int, const int, const int);
 void * tquery_thread(void*);
 int bisection(int,int,char*,useq_t**,int,int);
 match_t * new_match(useq_t *,int);
+
+lookup_t *new_lookup(int, int, int, int);
+int lut_search(lookup_t *, useq_t *);
+void lut_insert(lookup_t *, useq_t *);
+int seq2id(char *, int);
+useq_t *new_useq(int, char*);
+gstack_t *read_file(FILE*);
+void destroy_useq(useq_t*);
+void addmatch(useq_t*, useq_t*, int, int);
+void transfer_counts_and_update_canonicals(useq_t*);
+void unpad_useq (gstack_t*);
+int pad_useq(gstack_t*, int*);
+int canonical_order(const void*, const void*);
+int count_order(const void *a, const void *b);
+void *do_query(void*);
+void run_plan(mtplan_t*, int, int);
+mtplan_t *plan_mt(int, int, int, int, gstack_t*, const int);
+void message_passing_clustering(gstack_t*, int);
+void sphere_clustering(gstack_t*, int);
+void * _mergesort(void *);
+int seqsort(void **, int, int (*)(const void*, const void*), int);
+long count_trie_nodes(useq_t **, int, int);
+int AtoZ(const void *, const void *);
 
 #endif

### 2014-10-31 doc/document.tex
Documents as uploaded to Bioinformatics.
Updating the manuscript for revisions (in progress).

@@
 
-\title[Exact sequence clustering with starcode]{Starcode: an exact
-algorithm for sequence clustering}
-\author[Valera Zorita \textit{et~al}]{Eduard Valera Zorita\,$^{1,2}$, Pol
+\title[Sequence clustering by all pairs-search]{Starcode: sequence
+clustering based on all-pairs search }
+\author[Valera Zorita \textit{et~al}.]{Eduard Valera Zorita\,$^{1,2}$, Pol
 Cusc\'o\,$^{1,2}$ and Guillaume Filion\,$^{1,2}$\footnote{to whom
@@
 The increasing throughput of sequencing technologies offers new
-applications and challenges for computational biology. One such
-application is the use of random barcodes to trace and quantify
-transcripts or lineages in experimental setups. The high error
-rate of modern sequencers calls for additional post-processing
-techniques capable of detecting and reverting the misreads.
-However, in the absence of a reference population, the problem
-amounts to performing a pairwise comparison of all the barcodes,
+applications and challenges for computational biology. In many
+of those applications, sequencing errors need to be corrected.
+This is particularly important when using random DNA barcodes to
+trace and quantify transcripts or lineages, since spurious barcodes
+introduce inexistent entities. The reference population of random
+barcodes is unknown and in this case, error correction amounts
+to performing a pairwise comparison of all the barcodes,
 which is unfeasible for excessive computationally complexity.
@@
 determine which pairs of sequences lie within a given Levenshtein
-distance. The matched pairs are merged into clusters represented
-by a canonical sequence. The effiency of starcode is attributable
-to the poucet search, a novel  implementation of the Needleman-Wunsch
-algorithm performed on the nodes of a trie. On the task of clustering
+distance. For error correction or redundancy reduction purposes,
+matched pairs are then merged into clusters represented by a
+canonical sequence. The effiency of starcode is attributable to
+the poucet search, a novel implementation of the Needleman-Wunsch
+algorithm performed on the nodes of a trie. On the task of matching
 random barcodes, starcode outperforms sequence clustering algorithms
-in both speed and precision. We further show that starcode can also
-be used to identify enriched motifs in DNA and RNA sequences.
+in both speed and precision.
 
@@
 
-Sequence clustering is the process of grouping similar biological sequences.
-It has been traditionally applied to identify related protein families and to
-reduce sequence redundancy in databases. Recently, the advent of high throughput
-sequencing has created additional needs for efficient clustering algorithms, in
-particular because of the high error rate of such technologies. For instance,
-the Illumina platform \citep{pmid16056220} shows a 1-2\% error rate consisting of
- substitutions near the 3' end of the read \citep{pmid18660515, pmid21576222}. The
-PacBio platform shows a 15\% error rate that mostly corresponds to insertions
-and deletions \citep{pmid19023044}. As a consequence, the same sequence is often
-decoded in different ways, which artificially increases the diversity of the
-output.
+Sequencing technologies never achieve perfect precision. For instance,
+the Illumina platform \citep{pmid16056220} has a 1-2\% error rate
+consisting of substitutions \citep{pmid18660515, pmid21576222}
+and the PacBio platform has a 15\% error rate consisting of insertions
+and deletions \citep{pmid19023044}. The throughput of such technologies
+has recently created additional needs to develop efficient error
+correction algorithms.
 
-Sequencing errors can be discovered by mapping the reads onto a reference, if it
-is available. When the sequences are random or drawn from an unknown
-reference, clustering is the best option to tell real from spurious reads. One
-such case is the use of random barcodes to track cells or transcripts
-\citep{pmid18809713, pmid23953119}. Sequencing errors will create erroneous
-barcodes that have to be reverted to the original sequence. Hands on
-experience with real datasets shows that this step becomes limiting when
-the number of unique sequences is high. In search for a solution to this problem, we
-realized that heuristic approaches rely on assumptions that may not hold
-as technologies evolve. We therefore set out to find an exact algorithm.
+Sequencing errors can be discovered by comparing the reads to a
+reference genome. However, such a reference is not always available.
+When the sequences are random or taken from an unknown source,
+clustering is the main strategy to correct the errors.
+For instance, this situation arises when using random barcodes to
+track cells or transcripts \citep{pmid18809713, pmid23953119}.
+Sequencing errors will create erroneous (nonexistent) barcodes
+that have to be removed.
 
-The first task of clustering is a matching phase where closely related barcodes are
-paired, similarly to linked nodes on a graph. The second task is the clustering proper,
-where communities in this graph are merged. We called our algorithm ``starcode'', in
-reference to the star shape of the graph formed by the barcodes in the same
-cluster. The first version of the algorithm had the same performance regardless
-of the order in which the input sequences were processed, because it did not
-exploit data structuring of any kind. Exploiting the prefix redundancy of
-alphabetically sorted sequences allowed us to avoid unnecessary recomputations
-and gain speed. This is the rationale behind the poucet search algorithm at the
-heart of the matching step.
+Sequence clustering can be viewed as a community detection problem
+on graphs, where nodes represent sequences and edges represent matches
+between related sequences. The process consists of a matching phase
+(the most computationally intensive), where the graph is constructed,
+and a clustering phase where communities are identified.
 
-Here we describe the starcode algorithm and we benchmark it against
-existing software that perform related tasks. We show that starcode is
-both faster and more precise than the alternatives, achieving perfect
-clustering on ideal datasets. We further show that starcode can be used
-to identify enriched motifs in large datasets such as bacterial genomes
-or protein-DNA/protein-RNA interaction experiments.
+Here we describe a sequence clustering algorithm focused on error
+correction. The algorithm is called ``starcode'' in reference to 
+clusters of random barcodes, which typically have a star shape.
+Starcode is based on all-pairs search, \textit{i.e.} all the pairs
+of sequences below a given Levenshtein distance are identified during
+the graph construction phase. Matching is carried out by lossless
+filtration, followed by a trie search.  The novelty of the algorithm
+is the poucet strategy, which uses the redundancy of alphabetically
+sorted sequences to avoid unnecessary recomputations and gain speed.
+
+In this article we present and benchmark starcode. We show that on
+real biological datasets, starcode is faster than existing sequence
+clustering software. Even though starcode was designed for error
+correction, we show that it can be used for other problems. As an
+illustration, we use it to identify enriched motifs in a bacterial
+genome and in protein-RNA interaction experiments.
 
@@
 \subsection{Inexact string matching using tries}
-The matching method of starcode is based on a variation of the Needleman-Wunsch
-(NW) algorithm \citep{pmid5420325}. In the original algorithm
-(Figure~\ref{fig:NW}a), the Levenshtein distance between two sequences is found
-by applying a recurrence relation throughout a matrix of $mn$ terms, where $m$
-and $n$ are the respective sequence lengths. The complexity of this dynamic
-programming approach is $O(mn)$.
+The matching method of starcode is based on a variation of the
+Needleman-Wunsch (NW) algorithm \citep{pmid5420325}. In the
+original algorithm (Figure~\ref{fig:NW}a), the Levenshtein
+distance between two sequences is found by applying a recurrence
+relation throughout a matrix of $mn$ terms (the edit matrix),
+where $m$ and $n$ are the respective sequence lengths. The complexity
+of this dynamic programming approach is $O(mn)$.
 
@@
 \centerline{\includegraphics{NW.pdf}}
-\caption{Needleman-Wunsch (NW) alignment. \textbf{a}
-Alignment of GTTGCA and GATCCA using the NW alignment matrix. The
-margins (purple) are initialized and the cells of the matrix (yellow)
-are computed from left to right and from top to bottom by the following
-dynamic programming algorithm. The value in each cell is computed from
-the values of the neighboring top, left and top-left cells. Denoting
-$a$, $b$ and $c$ these respective terms, the value in the cell of
-coordinates $(i,j)$ is computed as $\min(a+1, b+1, c)$ if the $i$-th
-letter from the first sequence is identical to the $j$-th letter from
-the second, or $\min(a+1, b+1, c+1)$ if the letters are different. The
-Levenshtein distance between the two sequences is found in the bottom
-right cell. \textbf{b} Lower complexity algorithm to determine whether
-GTTGCA and GATCCA are 2-neighbors. The values in purple cells are set
+\caption{Needleman-Wunsch (NW) sequence comparison. \textbf{a}
+Comparison of GTTGCA and GATCCA. The margins of the edit matrix
+(purple) are initialized and the cells (yellow)
+are computed from left to right and from top to bottom by the NW
+dynamic programming algorithm. $E[i,j]$, the term of
+coordinates $(i,j)$ is computed as
+$\min(E[i-1,j]+1, E[i,j-1]+1, E[i-1,j-1]+\Delta(i,j))$, where
+$\Delta(i,j) = 0$ if the $i$-th symbol from the first sequence is the
+same as the $j$-th symbol from the second, and $\Delta(i,j) = 1$
+otherwise. The Levenshtein distance between the two sequences is the
+value of the bottom right cell. \textbf{b} Lower complexity algorithm
+to determine whether GTTGCA and GATCCA are 2-matches.
+The values in the purple cells are set
 during initialization. The dynamic programming algorithm proceeds as
-above, with the difference that it is interrupted if the value of a
-diagonal cell (bold borders) is larger than 2. The values in purple
-cells are not always the same as their equivalent in \textbf{a} (purple
-arrow), but the values in the yellow cells are nevertheless identical.
+above, with the difference that it is aborted if the value of a
+diagonal cell (bold borders) is larger than 2. The values in the
+purple cells may differ from the original NW scheme (purple arrow),
+but the values in the yellow cells are nevertheless identical.
 The values of the white cells are never computed, which contributes to
-making the algorithm less complex.}\label{fig:NW}
+reducing the complexity.}\label{fig:NW}
 \end{figure}
 
-In many instances, the only information of interest is to find out
-whether the sequences are $d$-neighbors (separated by a distance less
-than or equal to a threshold $d$).  In that case, the complexity then
-reduces to $O(d \min(m,n))$ as described below \citep{ukkonen}. Instead
-of initializing the margins of the matrix and computing all the terms,
-the matrix is initialized as shown on Figure~\ref{fig:NW}b and only the
-terms around the diagonal are computed. If the computation of a
-diagonal term yields a value greater than $d$, the distance is known to
-be greater than $d$ and the process is halted. Otherwise, the
-bottom-right term indicates the actual distance between the sequences,
-as in the NW algorithm.
+In many instances, the information of interest is to find out
+whether the sequences are $\tau$-matches (\textit{i.e.} their distance
+is less then or equal to a fixed threshold $\tau$).
+In that case, the complexity can be reduced to $O(\tau \min(m,n))$.
+Instead of computing all the terms of the edit matrix, it is initialized
+as shown on Figure~\ref{fig:NW}b and only the terms around the diagonal
+are computed. If a diagonal term has a value greater than $\tau$,
+the process is halted because the sequences are not $\tau$-matches.
 
-This method can be used for inexact matching of sequences against a
-reference set indexed as a prefix tree, also known as a trie
-\citep{ukkonen}. The terms of the matrix are updated row-wise as a
-depth-first search traverses the trie from the root, as illustrated
-in Figure~\ref{fig:trie}. Every time a node is visited, the row
-corresponding to its depth is recomputed. If the threshold value $d$
+This method can be used to match sequences against a prefix tree,
+also known as a trie \citep{ukkonen}. The terms of the edit matrix
+are updated row-wise while a depth-first search traverses the trie
+(Figure~\ref{fig:trie}). Every time a node is visited,
+a row is computed, and every time the search backtracks, a row is
+erased. If the threshold value $\tau$
 is exceeded for a diagonal term, the Levenshtein distance for all the
-downstream sequences is also necessarily greater than $d$. Therefore,
+downstream sequences is also necessarily greater than $\tau$. Therefore,
 no more hits are to be discovered in this path and the depth-first
@@
 tail node (corresponding to a sequence of the database) on the path
-of this search is a $d$-neighbor of the query. This method is
+of this search is a $\tau$-match of the query. This method is
 efficient because it eliminates large areas of the search space, and
-because the NW alignment of the query with each prefix of the database
+because the NW comparison of the query with each prefix of the database
 is computed only once.
@@
 \centerline{\includegraphics{trie.pdf}}
-\caption{Trie query with NW alignment. Each sequence of the index is a
-path in the trie. The query GTTGCA is written at the top of a NW matrix,
+\caption{NW algorithm on tries. Each sequence of the index is a
+path in the trie. The query GTTGCA is written at the top of the matrix,
 which is initialized as shown on Figure~\ref{fig:NW}b. The trie is
-traversed by a depth-first search (red path) from the root. At each
-depth, the node added to the path is written on the left of the NW
+traversed by a depth-first search (red path). At each
+depth, the node added to the path is written on the left of the edit
 matrix and the row is computed. Checkpoints from 1 to 4 (circled red
-numbers) show the state of the NW matrix as the search proceeds. The
-node labeled 3 is a leaf and thus corresponds to a 2-neighbor of the
-query. The search path then backtracks to the node labeled 2
-and the last rows of the NW are erased. The search path then goes to
-the node labeled 4, in which case the newly computed diagonal cell
-exceeds the threshold (circled in red). Even if this node has children,
-they are not visited (red crosses).}
+numbers) show the state of the edit matrix as the search proceeds. The
+node labeled 3 is a leaf and thus corresponds to a 2-match of the
+query. After discovering the hit, the search path backtracks to the
+node labeled 2 and the last rows of the edit matrix are erased.
+The search path then goes to the node labeled 4, in which case the
+newly computed diagonal cell exceeds the threshold (circled in red).
+Even if this node has children, they are not visited (red crosses)
+because there is no 2-match to discover.}
 \label{fig:trie}
@@
 \subsection{The poucet search algorithm}
-This strategy can be
-further improved. Notice that if two consecutive queries share a prefix of length
-$k$, the succession of computations up to the $k$-th row of the NW matrix will be exactly
-the same in both queries. To take advantage of this property, the input sequences are
-sorted alphabetically in order to maximize the prefix sharing. The main idea of the
-algorithm is to store the computational intermediates in the nodes of the trie and
-use them later to resume the computation in the next query. This way, the search can
-start at depth $k$ whenever a query shares a prefix of such length with the previous
-one.
+The search strategy can be further improved. If two consecutive
+queries share a prefix of length $k$, the succession of computations
+up to the $k$-th row of the edit matrix will be exactly the same
+for both queries. Therefore, computation intermediates can be
+stored in the nodes of the trie, so that the next trie search can start
+at depth $k$. However, storing the rows of the edit matrix in the nodes
+meets some difficulty. Indeed, on the $k$-th row, the terms on the
+right side of the diagonal depend on characters that are not shared
+between the two queries. This issue is solved by storing in each
+node a combination of row and column terms that form an angle shape,
+looking like a horizontally flipped L (Figure~\ref{fig:poucet}).
+Using this structure, the computation intermediates stored in a
+node at depth $k$ depend only on the first $k$ characters of the
+query.
 
-However, storing the rows of the NW matrix in the nodes, as suggested by the
-approach described in the previous section is not optimal. At depth equal to the
-length of the shared prefix, say $k$, the terms on the right of the diagonal were
-computed using characters that belong to the previous query. Since they also depend
-on the terms computed at lower depth, the search cannot restart at depth $k$. This
-issue is solved by storing in each node a combination of row and column terms that
-form an angle shape, looking
-like a horizontally flipped L character as shown on Figure~\ref{fig:poucet}. This way, the
-computation intermediates stored in a node of depth $k$ depend only on the previous
-characters of the query. Using this technique, the search can effectively restart
-at a depth equal to the length of the shared prefix with the previous query, thereby
-averting the need to recompute the first terms of the NW alignments.
-
-In the fairy tale ``Le Petit Poucet'', the hero seeds white pebbles for his older
-brothers to find their way home, which is reminiscent of the way queries pave
- the way for the next in this algorithm. We therefore called this search
+To take full advantage of this property, the input sequences are
+sorted alphabetically, which maximize prefix sharing between
+consecutive queries. In the fairy tale ``Le Petit Poucet'', the
+hero seeds white pebbles for his older brothers to find their way
+home, which is reminiscent of the way a smaller query (in alphabetical
+order) paves the way for the next. We therefore called this search
 algorithm ``poucet''.
@@
 \centerline{\includegraphics{poucet.pdf}}
-\caption{Poucet search algorithm. The algorithm proceeds with the same principles
-as shown on Figure~\ref{fig:trie} with the difference that the NW matrix is not
-updated row-wise, but along horizontally flipped L shapes. As the depth-first
-search proceeds, these values are stored in the nodes of the trie. Only the nodes
-at the top contain initialized values; for the other nodes, the values at the
-border are implicitly known to be 3. Since the values in the vertical part of
-the flipped L are the same for every child of the same node, they are computed
-only once (purple arrow). The values in the grey cells will be computed as the
-search path (red) visits the node. Storing the intermediates in the nodes allows
-the next query to restart at depth $k$ if it shares a common prefix of length
+\caption{Poucet search algorithm. The algorithm proceeds with the
+same principles as shown on Figure~\ref{fig:trie} with the difference
+that the edit matrix is not updated row-wise, but along a horizontally
+flipped L. As the depth-first search proceeds, these values
+are stored in the nodes of the trie. Since the values in the vertical
+part of the flipped L are the same for every child of a node,
+they are computed only once (purple arrow). The values in the grey
+cells will be computed as the search path (red) visits the node.
+Storing the intermediates in the nodes allows the next query to
+restart at depth $k$ if it shares a common prefix of length
 $k$ with the current query.}
@@
 
-\subsection{Lookup search}
-If the Levenshtein distance of two sequences of length $L$ is
-less than or equal to $\tau$, there must exist a perfect match between
-them of at least $k =\lfloor L / (\tau+1)\rfloor$ nucleotides. We take
-advantage of this property to build a lookup table containing all such
-$k$-mers of the sequences already present in the trie. When a new
-sequence is queried, all its subsequences are extracted and queried
-against the lookup table. If none of the subsequences are found, the
-minimum distance between the query and all the sequences in the trie
-is necessarily greater than $\tau$, and the trie search can therefore
-be omitted. This lookup is particulary useful for long query sequences
-(see \ref{performance}), where $k$ is large and the probability
-of an exact match between subsequences is small unless they are
-actually related, i.e. they belong to the same cluster. With this
-method, the absence of match is detected at the computational cost of
-several table lookups, compared to the higher cost of measuring the
-Levenshtein distance between long sequences.
+\subsection{Lossless filtration}
+When a query has no match, it is advantageous to omit the trie
+search. To this end, starcode uses a partition
+approach similar to that described by \cite{WuManber1992}. If the
+query is partitioned in $\tau+1$ segments, every $\tau$-match
+contains at least one of those (assuming that all the segments
+have length at least $\tau$). Indeed, there are at most $\tau$
+differences between the query and the match, so at least one segment
+is unmodified. In the $\tau$-match, the position of the shared
+segment is at most $\tau$ nucleotides on the left or on the right
+of its position in the query.
 
-In general, however, $L / (\tau+1)$ is not an integer. To deal
-with this complication, we divide the query into $\tau+2$ subsequences.
-The first $k+1$ are called \emph{words} and have a target length
-$(L-\tau) / (\tau+1)$. The last subsequence is called \emph{tail} and has
-length $\tau$. The length of the words is computed by integer
-division, and the remainder $R$ is distributed equally among the last
-$R$ words. As a result, the words have a length that differs by at
-most 1, with an aggregated length equal to $L-\tau$.
+These observations are the basis of a filtration method that proceeds
+as follows. Every time a sequence is added to the trie, its segments
+are added to $\tau+1$ different indexes. More specifically, the
+first $\tau$ nucleotides of the sequence are removed, and the rest
+of the sequence is partitioned in $\tau+1$ segments of sizes
+differing by at most 1 (the longer segments always in 3' for
+consistency). The first fragments are added to the first index,
+the second fragments to the second index \textit{etc}.
 
-Since two exact words at different positions in the sequence shall not
-represent a lookup match, we store and query the words against $k+1$
-distinct lookup tables. In order to allow insertions and deletions, 
-the lookup search must also query shifted versions of each word. In
-the worst matching case, only one word is found and the mismatches are
-evenly distributed in the query sequence. In such case, the maximum 
-number of cumulative insertions/deletions at the $i$-th word is
-$i-1$. Hence, the $i$-th word has to be shifted and queried against
-the lookup table $2(i-1)+1$ times to cover all the possible
-cases. The last word will be queried $2\tau+1$ times, including
-$\tau$ shifts on the right and $\tau$ shifts on the left. For the
-latter shift we make use of the $\tau$ nucleotides stored in the
-tail. The total number of lookups per sequence is $(\tau+1)^2$.
+Before the search, segments of the query are looked up in
+the indexes, and in case no match is found, the trie search is
+omitted because this query has no $\tau$-match. If at least one
+segment is found, the trie search is performed. As mentioned above,
+segments shared between the query and a $\tau$-match can be shifted
+up to $\tau$ nucleotides. For this reason, shifted segments are
+looked up in the indexes according to the following scheme ensuring
+that no match can be missed.
+(Figure~\ref{fig:filtration}). The rightmost segment is looked up
+in the $\tau$+1-th index, the second rightmost segment and the
+contiguous segments shifted by 1 nucleotides are are looked up in the
+$\tau$-th index and so on until the first segment and its
+contiguous segments shifted by up to $\tau$ nucleotides are
+looked up in the first index.
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics{filtration.pdf}}
+\caption{Lossless filtration illustrated by an example sequence
+of length 20 with $\tau=3$. The last $\tau$ nucleotides of the query
+are removed, and the rest is divided into 4 series of contiguous
+segments.  Each series is queried against a different index numbered
+I to IV.  For instance, the only segment queried against index I
+is GTTG, while those queried against index II are GCAA, CAAT and AATA.
+If any of the segments is found in the appropriate index, the trie
+search is performed, otherwise it is omited as there can be no
+$\tau$-match. Regardless of the result, segments labelled I to IV
+are then added to the corresponding respective index (\textit{i.e.}
+only one segment is added to each index).}
+\label{fig:filtration}
+\end{figure}
 
 \subsection{Seek and construct}
-To reduce the size of the search space, we use a dynamic ``seek and construct''
-approach whereby queries are processed meanwhile the trie is built. In other
-words, each sequence is matched against the trie before it is inserted. To
-illustrate why the trie does not need to contain all the sequences upon query,
-assume that two sequences A and B are $d$-neighbors. A is processed first. Since
-B is not yet inserted, A yields no hit. It is then inserted in the trie. At the
-time B is processed, A is a hit for the query and the match A-B is discovered.
-This approach guarantees that every hit is discovered, while maintaining the
-trie as ``thin'' as possible, thereby reducing the search time.
-The whole matching process is summarized in the pseudocode shown in
-Algorithms~\ref{alg:starcode} and \ref{alg:poucet}.
+To reduce the size of the search space, starcode uses a dynamic
+``seek and construct'' approach whereby queries are processed
+meanwhile the trie is built. In other words, each sequence is
+matched against the trie before it is inserted. If A and B are
+mutual $\tau$-matches, either A will be queried when B is in
+in the trie, or the converse. Either way, the match A-B is
+discovered. This guarantees that every $\tau$-match is discovered,
+while maintaining the trie as ``thin'' as possible, thereby reducing
+the search time.  The whole matching process is summarized in
+the pseudocode shown in Algorithms~\ref{alg:starcode} and
+\ref{alg:poucet}.
 
@@
         \Comment{Fig.1}
-    \State \textsc{compute} center value using row and column \Comment{Fig.1}
+    \State \textsc{compute} center value using row and column
+        \Comment{Fig.1}
     \If{center value $>\tau$} \Comment{Mismatches exceeded.}
@@
 \subsection{Parallelization}
-To parallelize the search, queries are separated into contiguous blocks after
-sorting. The matching algorithm proceeds in two phases. During the
-first phase, an independent trie is created and filled with the
-sequences of its associated block using the seek and construct process
-described above. In the second phase, all the other blocks of
-sequences are queried against but not inserted into each trie built in
-the first phase. If the queries are segregated into $N$ blocks, the
-first phase consists of $N$ seek and construct jobs, whereas the second
-consists of $N(N-1)/2$ query jobs. Since the jobs show little
-dependence on each other, the matching algorithm can be efficiently
-parallelized provided $N$ is larger than the number of independent
-threads.
+Queries are sorted and partitioned in contiguous blocks.
+The matching step then proceeds in two phases. In the first, a
+distinct trie is built from the sequences of each block according
+to the algorithm described above. In the second, all the sequence
+blocks are queried against the tries. If the queries are partitioned
+in $N$ blocks, the first phase consists of $N$ seek and construct
+jobs, wile the second consists of $N(N-1)/2$ query jobs. In each
+phase, the jobs show little dependency on each other, so the
+matching algorithm can be efficiently parallelized provided $N$
+is larger than the number of independent threads.
+
 \subsection{Clustering}
-Starcode implements a multi-purpose clustering algorithm called ``sphere
-clustering'' \citep{pmid23953119}, and a message passing algorithm
-\citep{mackay} tailored for the task of clustering random barcodes. In
-sphere clustering, barcodes are sorted by frequency of occurrence and
-each barcode, starting from the most frequent, can claim its
-$d$-neighbors that were not already claimed.
+The default clustering algorithm of starcode is designed to correct
+sequencing error. This method uses message passing \citep{mackay}
+to identfy and count ``canonical'' sequences (also referred to as
+centroids in the clustering terminology). Each sequence transfers its
+read count to its closest $\tau$-match provided the latter has at least 5
+times more counts. If the condition is not met, the transfer does
+not take place. If the sequence has several equally close
+$\tau$-matches, the counts are split equally among them. The
+process is repeated recursively, starting from sequences with
+lowest read count. The sequences with a positive read count at
+the end of the process are considered canonical. Clusters consist
+of all the sequences transferring their read counts to the same
+canonical sequence (sequence transferring their read counts to
+different canonicals are discarded).
 
-In message passing clustering, read counts are distributed equally
-among the closest neighbors of each barcode only if they are at least
-5 times more frequent. The remaining sequences at the end of the
-process are considered canonical, and their associated count is the
-estimated cluster size. A barcode is assigned to a cluster if all its
-read counts are eventually given to the corresponding canonical
-barcode. The barcodes for which the read counts are split between
-different canonical barcodes are not assigned to any cluster. The
-reason for imposing a factor 5 or larger in order to transfer the read
-counts is that barcodes with similar frequencies are not likely
-derived from each other through sequencing errors. More likely they
-are either unrelated, or they both derive from another more abundant
-barcode.
+Since no sequencing technology has an error rate higher than 20\%,
+it is expected that sequences appearing from sequencing errors will
+always have 5 times or lower read count than the canonical sequence.
+Otherwise, sequences are more likely unrelated, or both derived from
+the same canonical sequence.
+
+For other sequence clustering problems, starcode implements a
+multi-purpose algorithm called ``sphere clustering''
+\citep{pmid23953119}. In sphere clustering, sequences are sorted by
+frequency of occurrence. Starting from the most
+frequent, each sequence becomes canonical and claims all its
+$\tau$-matches, which forms a cluster of radius $\tau$ (hence the
+name). Claimed sequences are immediately removed, so that they can
+belong to only one cluster.
+
+\subsection{Benchmark conditions}
+All the tests were performed on a 16-core dual-processor
+Intel Xeon E5-2687W v2 system with 256~GB of DDR3-RAM at 1866~Mhz
 
@@
 
-\subsection{Performance}
+\subsection{Basic performance}
 \label{performance}
@@
 function of the number of input sequences $n$. In double logarithmic scale
-the trend is a straight line with slope 1.7, which suggests that the
-running time complexity of starcode is approximately $O(n^{1.7})$.
+the trend is a straight line with slope 1.5, suggesting that the
+running time complexity of starcode is lower than quadratic (unlike
+the naive implementation of all-pairs search).
 Figure~\ref{fig:perf}b shows that the running time grows exponentially
 as a function of the maximum Levenshtein distance used for clustering.
-This is a common feature of exact algorithms based on trie search. As
-this parameter increases, the poucet search bails out at a deeper
-average depth in a trie that fans out exponentially.
+The reason is that the trie fans out exponentially and the search bails
+out at a greater depth as the maximum distance increases.
 As a function of the sequence length, the running time first increases 
-but then plummets and stays low (Figure~\ref{fig:perf}c).
-Beyond a certain length, the lookup search algorithm starts to be
-efficient, and most of the queries are resolved without the need for
-trie-search.
+but then plummets and stays low (Figure~\ref{fig:perf}c). Beyond a
+threshold length, the filtering algorithm starts to be efficient, and
+most of the queries are resolved without searching the trie.
 Finally, we show the scalability of starcode with increasing
 number of threads in Figure~\ref{fig:perf}d. The search algorithm is fully
-parallel and the relative performance increases linearly up to 8 threads.
+parallel and the relative performance increases linearly up to 12 threads.
 The bending observed thereafter has two sources. The first is 
@@
 
-\begin{figure}[!tpb]%figurex
+\begin{figure}[!tpb]
+% Basic performance figure.
 \centerline{\includegraphics[scale=0.47]{scalability.pdf}}
-\caption{Scalability. a. Logarithm of the running time versus the
-logarithm of the number of sequences to be clustered. b. Running time as
-a function of the clustering distance. c. Running time versus length of
-the input sequences. d. Relative performance increase for different number
-of parallel threads. 
+\caption{Scalability. \textbf{a} Logarithm of the running time versus the
+logarithm of the number of sequences to be clustered. \textbf{b}
+Running time as a function of the clustering distance. \textbf{c}
+Running time versus length of the input sequences. \textbf{d} Relative
+performance increase for different number of parallel threads. 
 }\label{fig:perf}
@@
 
-\subsection{Benchmark against sequence clustering algorithms}
-Sequence clustering is routinely used to curate databases of non
-redundant nucleotide or protein sequences. We benchmarked starcode
-against the two popular sequence clustering algorithms CD-HIT
-\citep{pmid23060610} and USEARCH \citep{pmid20709691}.
+\subsection{Benchmark}
+We benchmarked starcode against the sequence clustering algorithms
+slidesort \citep{pmid21148542}, seed \citep{pmid21810899},
+rainbow \citep{pmid22942077} and cd-hit \citep{pmid23060610}. Even though
+slidesort is an all-pairs search algorithm, it was included in the
+benchmark because sequence comparison is the most computationally
+intensive step of the sequence clustering problem. Rainbow runs
+exclusively on paired-end reads, while the other tools run on single
+reads.
 
-We generated 1 million random 40-mers duplicated 47 times. For each
-40-mer, we also inserted 3 additional mutant sequences to the pool
-by sampling again 3 nucleotides at random positions. The probability
-that such a mutant sequence is within a distance 3 of any other
-40-mer is of the order of $10^{-12}$ and can be neglected. Each
-cluster thus consists of 50 sequences, of which 3 have a Levenshtein
-distance 3 from the canonical representant (referred to as the centroid
-in clustering terms).
+%% Synthetic data %%
+The performance of sequence clustering algorithms can be sensitive
+to the size of the clusters in the dataset, which in many applications
+is not known \textit{a priori}. We therefore set up a benchmark on
+artificial datasets to test the accuracy and the scaling of the
+tools on a known cluster structure. We generated 4 datasets of 1 million
+50-mers arranged in 1 to 1,000 clusters. Each cluster consisted
+of 100 repeats of the same centroid sequence, plus satellites
+derived from the centroid by incorporating 3 errors including at
+most 1 indel. The number of satellites per cluster ranged from
+999900 to 900. Evaluating the exactness of slidesort on these datasets
+was problematic because the number of 3-matches in each dataset is
+not known (pairs of statellites in the same cluster may be 3-matches
+or not). For this reason we only compared the number of pairs found
+by starcode with the number of pairs found by slidesort. The 
+outcome of the test is summarized in
+Figure~\ref{fig:benchmark}.
 
-All software were set to run on a 6-core single-processor Intel Xeon
-E5-2620 system with 48~GB of DDR3-RAM at 1333~Mhz.
-Using 12 threads, starcode clustered the shuffled 50 million sequences
-in 5 minutes and 3 seconds without a single error (the output
-consisted of 1 million clusters of size 50). There is no obvious
-way to compare exact algorithms with heuristic algorithms such as
-CD-HIT and USEARCH. We decided to allocate the heuristics 6 minutes
-on the same dataset (20\% additional time) and use the
-parameters giving the best accurary under this constraint.
-However, we could not find a combination of
-parameters such that either of them could terminate in less than
-6 minutes. To simplify the task, we extracted the unique sequences
-from the input file, bringing the number down to 4,000,000 and followed
-the same rationale. We could not find any set of parameters such
-that USEARCH could process this input in less than 6 minutes
-(because of poor multi-threading performance), but CD-HIT could
-perform the task when run as shown in Table~\ref{tab:params1}.
-In such conditions, it identified 2,798,656 clusters, indicating that
-the clustering was not nearly complete.
+\begin{figure}[!tpb]
+  \centerline{\includegraphics[scale=0.47]{benchmark.pdf}}
+\caption{Accuracy of the sequence clustering and all-pairs search
+algorithms. Starcode, seed and cd-hit were bencmarked on 4 datasets
+with the same number of sequences, but different number of clusters
+(see main text). \textbf{a} Starcode identifies the correct number
+of clusters, while seed and cd-hit identify about 40 false positives
+per true positive. \textbf{b} Slidesort identifies 5-10\% less
+pairs than starcode. In both panels, the red line is first
+bisector.}
+\label{fig:benchmark}
+\end{figure}
+
+While starcode achieves perfect clustering on all 4 datasets, the
+clustering achieved by seed and cd-hit is incomplete. Both tools
+identify approximately 40 false clusters per true cluster on all
+the datasets. We also observed that slidesort found 5-10\% less
+3-matches than starcode on all the datasets. We were surprised by
+this result because slidesort is claimed to be an exact algorithm.
+However, this was clearly not the case when we ran additional tests
+on smaller datasets where naive pairwise comparisons is feasible
+(the starcode repository contains such a dataset along with the
+instructions to reproduce our observations, see
+http://github.com/gui11aume/starcode/misc). In conclusion, starcode
+was the only tool to achieve perfect precision.
+
+%\begin{table}[h]
+%\centering
+%\caption{Software execution parameters (benchmark 1)}
+%\resizebox{\columnwidth}{!}{
+%\begin{tabular}{|l|l|}
+%    \hline
+%    starcode & \texttt{-d3 -t12 -s}\\ \hline
+%    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 7 -c .925 -T 12 -M 0 -r 0 -A 40}\\ \hline
+%\end{tabular}
+%}
+%\label{tab:params1}
+%\end{table}
+
+%% Biological datasets %%
+
+We benchmarked sequence clustering algorithms on the problem of
+clustering TRIP barcodes \citep{pmid23953119}. Briefly, the principle
+of TRIP (Thousands of Reporters Integrated in Parallel) is to tag
+reporter transcripts with random barcodes and measure the abundance
+of barcodes in the RNA as a proxy for gene expression. There is no
+reference to match aberrant barcodes against, because the tagging
+sequences are unknown.
+
+The basic properties of the datasets used for benchmarking are
+summarized in Table~\ref{tab:biobench_properties}. Datasets 1 and 2
+consist of Illumina single reads; they differ by the total read count
+and by the empirical cluster sizes. According to the output of starcode,
+the largest clusters of dataset 1 contain approximately 2,000 sequences,
+while dataset 2 contains 4 clusters with more than 1 million sequences.
+Dataset 3 consist of Illumina paired-end reads.
 
@@
 \centering
-\caption{Software execution parameters (benchmark 1)}
-\resizebox{\columnwidth}{!}{
-\begin{tabular}{|l|l|}
+\caption{Summary of the biological datasets used for benchmarking.
+All the datasets are Illumina reads.}
+\begin{tabular}{|l|l|l|l|}
     \hline
-    starcode & \texttt{-d3 -t12 -s}\\ \hline
-    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 7 -c .925 -T 12 -M 0 -r 0 -A 40}\\ \hline
+    \textit{dataset}   & \textit{read count} &
+    \textit{read length} & \textit{type} \\ \hline
+    dataset 1 & 16,457,527   & 50  & single      \\ \hline
+    dataset 2 & 127,675,537  & 50  & single      \\ \hline
+    dataset 3 & 2,460,226    & 100  & paired-end \\ \hline
 \end{tabular}
-}
-\label{tab:params1}
+\label{tab:biobench_properties}
 \end{table}
 
-We next benchmarked starcode on the biological dataset used by
-\cite{pmid23060610}. Because USEARCH imposes a memory limit on the
-size of the input file, we clustered the forward reads on 95\%
-identity from one lane only (reference ERR011089), adding up to
-4,189,859 reads, each 75 nucleotides long. Using 12 threads, starcode
-identified 4,111,111 clusters in 6 minute 19 seconds. Allocating up to
-7 minutes 30 seconds to CD-HIT and USEARCH, we selected the parameters
-giving highest accuracy. In such conditions (see Table~\ref{tab:params2})
-CD-HIT identified 4,126,352 clusters while USEARCH could not be run in
-the allocated time. These results indicate that starcode achieves good
-performance on clustering high throughput sequencing reads.
+The running times of starcode, seed, slidesort, rainbow and cd-hit-est
+are summarized in Table~\ref{tab:biobench_runtime}. Starcode was faster
+than the other tools on all the datasets. Seed came in second position
+on datasets 1 and 2, with a running time approximately 800 and 20 times
+greater, respectively. Rainbow came in second position on dataset 3,
+but it is the only other tool able cluster paired-end reads. We did not
+record the exact running times past 10 days since this is several
+orders of magnitude higher than the running time of starcode.
 
@@
 \centering
-\caption{Software execution parameters (benchmark 2)}
-\resizebox{\columnwidth}{!}{
-\begin{tabular}{|l|l|}
+\caption{Running time (in seconds) of the software on three biological
+datasets. Runs longer than 10 days were interrupted. A dash indicates
+that the software cannot be used for this dataset.}
+\begin{tabular}{|l|l|l|l|}
     \hline
-    starcode & \texttt{-d4 -t12 -s}\\ \hline
-    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 8 -c .95 -T 12 -M 0 -r 0 -A 75}\\ \hline
+    \textit{software}   & \textit{dataset 1} &
+      \textit{dataset 2} & \textit{dataset 3} \\ \hline
+    starcode   & 171         & 2,898       & 44    \\ \hline
+    seed       & 135,959     & 60,374      & -     \\ \hline
+    sidesort   & 632,168     & $>$ 10 days & -     \\ \hline
+    rainbow    & -           & -           & 306   \\ \hline
+    cd-hit-est & $>$ 10 days & 512,591     & -     \\ \hline
 \end{tabular}
-}
-\label{tab:params2}
+\label{tab:biobench_runtime}
 \end{table}
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics{benchmark.pdf}}
-\caption{Benchmark of starcode against short read mappers.
-\textbf{a} Running time compared to Bowtie2, BWA and GEM on the
-dataset described in the text. The running time is decomposed as
-extraction of unique sequences (purple), Burrows-Wheeler indexing
-(yellow) and query proper (white). The extraction was performed with
-the Linux \texttt{sort} command, so the time is the same in every
-case. \textbf{d} Error rate compared to Bowtie2, BWA and GEM,
-expressed as the proportion of matches with a single hit. This is an
-underestimate of the true error rate.}
-\label{fig:benchmark}
-\end{figure}
-
-\subsection{Benchmark against short read mappers}
-A potential strategy to cluster short sequences is to use
-Burrows-Wheeler transform-based short read mappers to match a set of
-sequences against itself. Once matches are available, several community
-detection algorithms can be used to identify the clusters. We 
-benchmarked starcode against the short read mappers Bowtie2
-\citep{pmid22388286}, BWA \citep{pmid19451168} and GEM
-\citep{pmid23103880}. Short read mappers do not perform clustering, so
-we only evaluated their performance on the matching problem.
-
-For this test, we used the first dataset described above, in which
-every sequence has at least 2 distinct matches at a distance 3 (mutant
-sequences from the same cluster may be at a distance up to 6 from each
-other, but their distance to the centroid is always 3). This means that
-we can use matches with a single hit (every read trivially matches
-itself) as a lower bound on the inaccuracy. We used the 4,000,000 unique
-40-mers to build an index that was later queried with the same 40-mers
-using parameters shown in Table~\ref{tab:srm}.
+The memory footprint of the different tools on the same datasets is
+shown in Table~\ref{tab:biobench_memory}. The values represent the
+peak memory usage throughout the run on the datasets described above.
+On datasets 1 and 3, starcode had a significantly larger memory usage
+than the best tool, while on dataset 2 it used significantly less
+memory than seed and approximately as much as cd-hit. A large portion
+of the memory usage of starcode is dedicated to the trie structures
+used in the poucet search, which explains why starcode is in general
+more memory-demanding than alternative tools.
 
@@
 \centering
-\caption{Software execution parameters (benchmark 3)}
-\resizebox{\columnwidth}{!}{
-\begin{tabular}{|l|l|}
+\caption{Memory usage (in Gb).}
+\begin{tabular}{|l|l|l|l|}
     \hline
-    Bowtie2 indexer & \texttt{bowtie2-build --quiet}\\ \hline
-    Bowtie2 2.1.0 & \texttt{-f -a -p 12 --no-hd --very-sensitive}\\ \hline
-    BWA indexer & \texttt{bwa index -a is}\\ \hline
-    BWA 0.7.9a & \texttt{mem -t 12 -a -k1 -B0} \\ \hline
-    GEM indexer & \texttt{gem-indexer -T 12} \\ \hline
-    GEM 1.423 & \texttt{-e3 -s4 -T12 --granularity 100000}\\ \hline
+    \textit{software}   & \textit{dataset 1} &
+      \textit{dataset 2} & \textit{dataset 3} \\ \hline
+    starcode   & 9.4 & 30.9 & 5.2 \\ \hline
+    seed       & 4.1 & 53.9 & -   \\ \hline
+    sidesort   & 1.9 & 13.9 & -   \\ \hline
+    rainbow    & -   & -  & 0.5   \\ \hline
+    cd-hit-est & 3.8 & 28.5 & -   \\ \hline
 \end{tabular}
-}
-\label{tab:srm}
+\label{tab:biobench_memory}
 \end{table}
 
-Figure~\ref{fig:benchmark}a shows the running time of starcode along
-with the decomposition of the running time for the three mappers. Note
-that the time to extract unique sequences is always the same because
-we used the Linux \texttt{sort -u} command in every case. The running
-time of starcode is slightly lower than short read mappers when all
-steps ares taken into account. Figure~\ref{fig:benchmark}b shows the
-lower bound on the error rate, as estimated by the proportion of single
-hits. BWA was found to be the most accurate aligner with a lower bound
-of 31\% sequences with a single match. This measure is an underestimate
-of the true error rate, but given that starcode is an exact algorithm,
-a more accurate measure is not necessary: starcode can compete with the
-short read aligners on the matching problem, and it provides a precision
-that none of the other tools can offer.
-
-\subsection{Clustering TRIP barcodes}
-In the course of setting up the TRIP technology in our laboratory
-\citep{pmid23953119}, we realized the need to develop efficient
-algorithms to cluster similar sequences. Briefly, the principle of TRIP
-(Thousands of Reporters Integrated in Parallel) is to tag reporter
-transcripts with random barcodes and measure the abundance of barcodes
-in the RNA as a proxy for gene expression. There is no reference to
-match aberrant barcodes against, because the tagging sequences are
-unknown. Instead, barcodes are matched against each other and clustered
-by similarity to infer canonical sequences.
-
-We tested the efficiency of starcode on the TRIP dataset from
-\cite{pmid23953119}. In the experiment labeled mPGKA, we identified
-24.1 million (91\%) barcode-containing reads out of 26.6 million,
-consisting of approximately 223,000 and 220,000 unique barcode sequences
-for PCR replicates 1 and 2, respectively. Following the authors, we kept
-only the barcodes with at least 5 counts and performed clustering as
-described: ``First we sorted barcodes according to their counts. Then,
-for each barcode (starting from the most frequent one), we identified
-and removed all its mutant versions, defined as barcodes within a Hamming
-distance of 2.'' We implemented this method, here on referred to as the
-``sequential algorithm'' with the function call
-\texttt{stringDist(method='lv', maxDist=2)} from R the package
-\texttt{stringDist} \citep{R}.  When replacing the Hamming distance by
-the Levenshtein
-distance, starcode produced exactly the same output as the sequential
-algorithm. The running time of starcode averaged over the replicates was
-2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which
-represents a 75-fold speedup.
-
-The performance of the sequential algorithm relies on the arbitrary
-exclusion of reads with less than 5 counts, the main purpose of which
-is to reduce the computational burden. When all barcodes were kept,
-the average running time of starcode was 10.27 seconds, \textit{versus}
-$>$ 6 hours for the sequential algorithm. Using the Hamming distance,
-as the authors originally suggested, decreased the average running time
-of the sequential algorithm to about 3 hours and 30 minutes (and
-the output differed from that of starcode). Note that in all the cases
-mentioned above, starcode was run with a single core to compare the
-algorithms based on similar computer resources. In conclusion, the
-barcode clustering problem can be simplified by various tricks, but
-starcode brings down the running time to nearly instantaneous, and
-thereby obviates the need for such arbitrary heuristics.
+%\subsection{Clustering TRIP barcodes}
+%In the course of setting up the TRIP technology in our laboratory
+%\citep{pmid23953119}, we realized the need to develop efficient
+%algorithms to cluster similar sequences. Briefly, the principle of TRIP
+%(Thousands of Reporters Integrated in Parallel) is to tag reporter
+%transcripts with random barcodes and measure the abundance of barcodes
+%in the RNA as a proxy for gene expression. There is no reference to
+%match aberrant barcodes against, because the tagging sequences are
+%unknown. Instead, barcodes are matched against each other and clustered
+%by similarity to infer canonical sequences.
+%
+%We tested the efficiency of starcode on the TRIP dataset from
+%\cite{pmid23953119}. In the experiment labeled mPGKA, we identified
+%24.1 million (91\%) barcode-containing reads out of 26.6 million,
+%consisting of approximately 223,000 and 220,000 unique barcode sequences
+%for PCR replicates 1 and 2, respectively. Following the authors, we kept
+%only the barcodes with at least 5 counts and performed clustering as
+%described: ``First we sorted barcodes according to their counts. Then,
+%for each barcode (starting from the most frequent one), we identified
+%and removed all its mutant versions, defined as barcodes within a Hamming
+%distance of 2.'' We implemented this method, here on referred to as the
+%``sequential algorithm'' with the function call
+%\texttt{stringDist(method='lv', maxDist=2)} from R the package
+%\texttt{stringDist} \citep{R}.  When replacing the Hamming distance by
+%the Levenshtein
+%distance, starcode produced exactly the same output as the sequential
+%algorithm. The running time of starcode averaged over the replicates was
+%2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which
+%represents a 75-fold speedup.
+%
+%The performance of the sequential algorithm relies on the arbitrary
+%exclusion of reads with less than 5 counts, the main purpose of which
+%is to reduce the computational burden. When all barcodes were kept,
+%the average running time of starcode was 10.27 seconds, \textit{versus}
+%$>$ 6 hours for the sequential algorithm. Using the Hamming distance,
+%as the authors originally suggested, decreased the average running time
+%of the sequential algorithm to about 3 hours and 30 minutes (and
+%the output differed from that of starcode). Note that in all the cases
+%mentioned above, starcode was run with a single core to compare the
+%algorithms based on similar computer resources. In conclusion, the
+%barcode clustering problem can be simplified by various tricks, but
+%starcode brings down the running time to nearly instantaneous, and
+%thereby obviates the need for such arbitrary heuristics.
 

### 2014-10-31 doc/plot-benchmark.R

@@ -1,0 +1,53 @@
+# Data
+clusters = c(1, 10, 100, 1000)
+starcode = c(1, 10, 100, 1000)
+seed = c(40, 363, 3701, 37596)
+cd.hit = c(49, 448, 3924, 29541)
+starcode_pairs = c(NA, 277168690, 116797218, 96383875)
+slidesort_pairs = c(NA, 263551040, 107392689, 87884477)
+cluster_sizes = c(1000, 10000, 100000, 1000000)
+starcode_time = c(1*60+54, 3*60+23, 7*60+43, 12*60+42)
+seed_time = c(51, 43, 37, 30)
+cd.hit_time = c(16, 14, 13, 9)
+slidesort_time = c(3*60+38, 36*60+47, 384*60+38, 3840*60)
+# Plot
+pdf("benchmark.pdf")
+par(mfrow=c(2,2))
+plot(log10(clusters), log10(starcode), type = "n",
+     ylim = c(0,max(log10(c(starcode, cd.hit, seed)))),
+     xlab = "Real number of clusters [log10]",
+     ylab = "Inferred number of clusters [log10]",
+     xaxp = c(0,3,3),
+     cex.lab = 1.3)
+abline(0,1, col=2)
+points(log10(clusters), log10(starcode), pch=1, type="b")
+points(log10(clusters), log10(seed), pch=16, type="b")
+points(log10(clusters), log10(cd.hit), pch=17, type="b")
+legend(x = "topleft", bty="n",
+     legend = c("starcode", "seed", "cd-hit"),
+     pch = c(1, 16, 17),
+     cex = 1.2)
+mtext("a", adj=-.05, cex=1.5, line=1)
+plot(slidesort_pairs/1e6, starcode_pairs/1e6, type = "b",
+     xlab = "Pairs in slidesort [millions]",
+     ylab = "Pairs in starcode [millions]",
+     cex.lab = 1.3)
+abline(0,1, col=2)
+mtext("b", adj=-.05, cex=1.5, line=1)
+plot(log10(cluster_sizes), starcode_time, type = "n",
+     ylim = c(1,max(log10(c(starcode_time, cd.hit_time,
+                      seed_time, slidesort_time)))),
+     xlab = "Cluster size [log10]",
+     ylab = "Running time [log10(seconds)]",
+     xaxp = c(3,6,3),
+     cex.lab = 1.3)
+points(log10(cluster_sizes), log10(starcode_time), pch=1, type="b")
[…]

### 2014-10-31 doc/plot-scalability.R

@@ -1,24 +1,30 @@
 # Data
-X1 = c(0,1,2,3,4,5)
-Y1 = c(0.465, 0.631, 0.776, 19.072, 2*60+20.767, 9*60+13.506)
-X2 = 10000*c(1,2,4,8,16,32,64)
-Y2 =c(0.294, 0.966, 3.374, 11.849, 39.074, 60+55.348, 4*60+55.366)
-X3 = c(1,2,4,6,8,10,12)
-Y3 = c(8*60+52, 4*60+20, 2*60+14, 1*60+29, 1*60+8, 58, 51)
-X4 = 10*c(1,2,3,4,5,6,7,8)
-Y4 = c(4*60+49, 8*60+41, 9*60+2, 9*60+18, 3*60+42, 60+6, 60+12, 60+19)
+X1 = c(0,1,2,3,4)
+Y1 = c(3.15, 4.108, 8.315, 5*60+8.83, 37*60+10.42)
+X2 = c(39062, 78125, 156250, 312500, 625000, 1250000, 2500000, 5000000)
+Y2 = c(1.735, 6.263, 20.573, 59.309, 2*60+40.555, 6*60+50.695,
+      15*60+28.397, 37*60+20.744)
+X3 = c(1,2,3,4,6,8,10,12,14,16)
+Y3 = c(4*60+48.256, 2*60+24.551, 60+37.476, 60+13.478, 48.613,
+       36.442, 28.651, 24.366, 21.665, 19.689)
+X4 = 10*c(1,2,3,4,5,6,8,12,16,22,28,34,42)
+Y4 = c(2*60+27.740, 4*60+37.854, 4*60+54.134, 5*60+8.237, 60+54.749,
+       17.325, 20.589, 23.390, 23.516, 25.902, 26.395, 27.595,
+       31.319)
 # Plot
 pdf("scalability.pdf")
 par(mfrow=c(2,2))
-plot(log(X2),log(Y2), type='b', xlab="Ln number of sequences", ylab="Ln Running time (seconds)", cex.lab=1.3)
-abline(lm(log(Y2)~log(X2)), col=2)
-text(x=13, y=5.65, "-16.7+1.7x", adj=1)
+plot(log10(X2),log10(Y2), type='b', xlab="Number of sequences [log10]", ylab="Running time [log10(seconds)]", cex.lab=1.3)
+fit = lm(log10(Y2)~log10(X2))
+abline(fit, col=2)
+text(x=6.2, y=3, adj=1,
[…]

### 2014-11-01 src/_starcode.h
Added check for proper DNA letters in input files.

@@ -57,6 +57,24 @@
 #define min(a,b) (((a) < (b)) ? (a) : (b))
 #define max(a,b) (((a) > (b)) ? (a) : (b))
 
+static const int valid_DNA_char[256] = { 
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,1,0,1,0,0,0,1,0,0,0,0,0,0,1,0,
+   0,0,0,0,1,0,0,0,0,0,0,0,0,0,0,0,
+   0,1,0,1,0,0,0,1,0,0,0,0,0,0,1,0,
+   0,0,0,0,1,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
+};
 
 struct useq_t;
 struct c_t;

### 2014-11-01 src/starcode.c

@@ -853,17 +853,21 @@ read_rawseq
       else {
          seq = copy;
       }
-      if (strlen(seq) > MAXBRCDLEN) {
+      size_t seqlen = strlen(seq);
+      if (seqlen > MAXBRCDLEN) {
          fprintf(stderr, "max sequence length exceeded (%d)\n",
                MAXBRCDLEN);
          fprintf(stderr, "offending sequence:\n%s\n", seq);
          abort();
       }
-      useq_t *new = new_useq(count, seq, NULL);
-      if (new == NULL) {
-         alert();
-         krash();
+      for (size_t i = 0 ; i < seqlen ; i++) {
+         if (!valid_DNA_char[(int)seq[i]]) {
+            fprintf(stderr, "invalid input\n");
+            fprintf(stderr, "offending sequence:\n%s\n", seq);
+            abort();
+         }
       }
+      useq_t *new = new_useq(count, seq, NULL);
       push(new, &uSQ);
    }
 
@@ -899,11 +903,19 @@ read_fasta
       if (line[nread-1] == '\n') line[nread-1] = '\0';
 
       if (lineno %2 == 0) {
-         if (strlen(line) > MAXBRCDLEN) {
+         size_t seqlen = strlen(line);
+         if (seqlen > MAXBRCDLEN) {
             fprintf(stderr, "max sequence length exceeded (%d)\n",
                   MAXBRCDLEN);
             fprintf(stderr, "offending sequence:\n%s\n", line);
             abort();
+         }
+         for (size_t i = 0 ; i < seqlen ; i++) {
+            if (!valid_DNA_char[(int)line[i]]) {
+               fprintf(stderr, "invalid input\n");
+               fprintf(stderr, "offending sequence:\n%s\n", line);
+               abort();
+            }
          }
          useq_t *new = new_useq(1, line, header);
          if (new == NULL) {
@@ -958,11 +970,19 @@ read_fastq
          strncpy(header, line, M);
       }
       else if (lineno % 4 == 2) {
-         if (strlen(line) > MAXBRCDLEN) {
+         size_t seqlen = strlen(line);
+         if (seqlen > MAXBRCDLEN) {
             fprintf(stderr, "max sequence length exceeded (%d)\n",
                   MAXBRCDLEN);
             fprintf(stderr, "offending sequence:\n%s\n", line);
             abort();
+         }
+         for (size_t i = 0 ; i < seqlen ; i++) {
[…]

### 2014-11-03 doc/document.tex
Updated benchmark figure. Text of the manuscript in progress.

@@
 Needleman-Wunsch (NW) algorithm \citep{pmid5420325}. In the
-original algorithm (Figure~\ref{fig:NW}a), the Levenshtein
+original algorithm (Figure \ref{fig:NW}a), the Levenshtein
 distance between two sequences is found by applying a recurrence
@@
 Instead of computing all the terms of the edit matrix, it is initialized
-as shown on Figure~\ref{fig:NW}b and only the terms around the diagonal
+as shown on Figure \ref{fig:NW}b and only the terms around the diagonal
 are computed. If a diagonal term has a value greater than $\tau$,
@@
 are updated row-wise while a depth-first search traverses the trie
-(Figure~\ref{fig:trie}). Every time a node is visited,
+(Figure \ref{fig:trie}). Every time a node is visited,
 a row is computed, and every time the search backtracks, a row is
@@
 path in the trie. The query GTTGCA is written at the top of the matrix,
-which is initialized as shown on Figure~\ref{fig:NW}b. The trie is
+which is initialized as shown on Figure \ref{fig:NW}b. The trie is
 traversed by a depth-first search (red path). At each
@@
 node a combination of row and column terms that form an angle shape,
-looking like a horizontally flipped L (Figure~\ref{fig:poucet}).
+looking like a horizontally flipped L (Figure \ref{fig:poucet}).
 Using this structure, the computation intermediates stored in a
@@
 \caption{Poucet search algorithm. The algorithm proceeds with the
-same principles as shown on Figure~\ref{fig:trie} with the difference
+same principles as shown on Figure \ref{fig:trie} with the difference
 that the edit matrix is not updated row-wise, but along a horizontally
@@
 that no match can be missed.
-(Figure~\ref{fig:filtration}). The rightmost segment is looked up
+(Figure \ref{fig:filtration}). The rightmost segment is looked up
 in the $\tau$+1-th index, the second rightmost segment and the
@@
 canonical sequence (sequence transferring their read counts to
-different canonicals are discarded).
+different canonicals are discarded). Note that the radius of the
+clusters can be higher than the maximum distance used for matching.
 
@@
 
-\subsection{Basic performance}
+\subsection{Presentation and basic performance}
 \label{performance}
-We measured basic performance and scalability metrics of starcode on a
-dataset of pseudo random sequences (Figure~\ref{fig:perf}). The standard
-configuration consists of a set of 1,000,000 sequences of length 40 running
-on 1 thread and with a maximum Levenshtein distance of 3. To test the
-scalability as a function of a single parameter, only the parameter under
-study was modified whereas the others were kept constant.
+Starcode is a general purpose DNA sequence clustering tool with a
+strong focus on error correction. Errors are assumed to be mismatches,
+insertions or deletions (the implementation presented here matches
+sequences with up to 8 errors). The input sequences can be single
+or paired-end reads, with an upper limit of 1024 nucleotides (512
+for paired-end). Sequences may be of variable length, they may
+be trimmed and filtered for quality or not. Input formats may be raw
+sequence, fasta or fastq (in which case starcode ignores the quality).
+By default, starcode outputs sequence clusters, it can also be used
+to filter out redundant sequences from input files. By default,
+clustering is performed under the assumption that divergence occurs
+from experimental errors (sequencing errors, PCR mutations
+\textit{etc}), a more general algorithm is also available for other
+clustering problems (an example of which is given in section
+\ref{section:motifs}).
 
-Figure~\ref{fig:perf}a shows the running time of starcode as a
-function of the number of input sequences $n$. In double logarithmic scale
-the trend is a straight line with slope 1.5, suggesting that the
-running time complexity of starcode is lower than quadratic (unlike
-the naive implementation of all-pairs search).
-Figure~\ref{fig:perf}b shows that the running time grows exponentially
+We show the basic performance and scalability of starcode on a
+dataset of pseudo random sequences (Figure \ref{fig:perf}). The standard
+configuration is a set of 1,000,000 sequences of length 40, running
+on 1 thread and with a maximum Levenshtein distance of 3. Only one
+parameter is modified whereas the others were kept constant.
+
+Figure \ref{fig:perf}a shows the running time of starcode as a
+function of the number of input sequences $n$. In double logarithmic
+scale the trend is a straight line with slope 1.5, suggesting that the
+running time complexity of starcode is lower than quadratic (the
+naive implementation of all-pairs search). Note that the sequences of
+this dataset have no match, see section \ref{section:benchmark} for
+an evaluation of the performance on more realistic datasets.
+Figure \ref{fig:perf}b shows that the running time grows exponentially
 as a function of the maximum Levenshtein distance used for clustering.
@@
 As a function of the sequence length, the running time first increases 
-but then plummets and stays low (Figure~\ref{fig:perf}c). Beyond a
+but then plummets and stays low (Figure \ref{fig:perf}c). Beyond a
 threshold length, the filtering algorithm starts to be efficient, and
 most of the queries are resolved without searching the trie.
-Finally, we show the scalability of starcode with increasing
-number of threads in Figure~\ref{fig:perf}d. The search algorithm is fully
+Finally, we show the scalability of starcode with increasing number
+of threads in Figure \ref{fig:perf}d. The search algorithm is fully
 parallel and the relative performance increases linearly up to 12 threads.
@@
 \subsection{Benchmark}
+\label{section:benchmark}
 We benchmarked starcode against the sequence clustering algorithms
@@
 outcome of the test is summarized in
-Figure~\ref{fig:benchmark}.
+Figure \ref{fig:benchmark}.
 
@@
 instructions to reproduce our observations, see
-http://github.com/gui11aume/starcode/misc). In conclusion, starcode
-was the only tool to achieve perfect precision.
+http://github.com/gui11aume/starcode/misc).
+
+The running time of the different tools as a function of the size
+of the clusters is shown on Figure \ref{fig:benchmark}c. For seed
+and cd-hit it decreases as the cluster size increases, whereas
+for starcode and slidesort it increases. In spite of this trend,
+the performance of starcode remains competitive, even for clusters
+of 1 million sequences. The memory usage is shown in Figure
+\ref{fig:benchmark}c.
+In conclusion, starcode was the only tool to achieve perfect precision
+on these datasets. It also maintains competitive performance in
+running time and memory usage.
 
@@
 %% Biological datasets %%
-
+The performance on artificial data is not always in agreement with
+the performance on experimental datasets. In typical experiments,
+the sizes of the clusters may be uneven, and the reads may contain
+near constant regions that may degrade the performance.
 We benchmarked sequence clustering algorithms on the problem of
@@
 \subsection{Identifying enriched sequence motifs}
+\label{section:motifs}
 Sequence motifs are thought to play an important role in DNA metabolism.
@@
 running time will also be shorter thanks to the lookup table
-search (Figure~\ref{fig:perf}c).
+search (Figure \ref{fig:perf}c).
 
@@
 distance will have to be increased, leading to exponentially longer
-running times (Figure~\ref{fig:perf}b). However, for the imporant
+running times (Figure \ref{fig:perf}b). However, for the imporant
 practical case that the divergence is driven by sequencing errors,

### 2014-11-03 doc/plot-benchmark.R

@@ -3,16 +3,21 @@ clusters = c(1, 10, 100, 1000)
 starcode = c(1, 10, 100, 1000)
 seed = c(40, 363, 3701, 37596)
 cd.hit = c(49, 448, 3924, 29541)
-starcode_pairs = c(NA, 277168690, 116797218, 96383875)
-slidesort_pairs = c(NA, 263551040, 107392689, 87884477)
+starcode_pairs = c(1828809738, 277168690, 116797218, 96383875)
+slidesort_pairs = c(1722705247, 263551040, 107392689, 87884477)
 cluster_sizes = c(1000, 10000, 100000, 1000000)
 starcode_time = c(1*60+54, 3*60+23, 7*60+43, 12*60+42)
 seed_time = c(51, 43, 37, 30)
 cd.hit_time = c(16, 14, 13, 9)
-slidesort_time = c(3*60+38, 36*60+47, 384*60+38, 3840*60)
+slidesort_time = c(3*60+38, 36*60+47, 384*60+38, 2369*60+1)
+starcode_mem = c(2606224, 2420160, 1733536, 1383892)
+slidesort_mem = c(388912, 415732, 410948, 371692)
+seed_mem = c(371692, 2320388, 2199632, 2071132)
+cd.hit_mem = c(260992, 249052, 247916, 247660)
 # Plot
 pdf("benchmark.pdf")
 par(mfrow=c(2,2))
+
 plot(log10(clusters), log10(starcode), type = "n",
      ylim = c(0,max(log10(c(starcode, cd.hit, seed)))),
      xlab = "Real number of clusters [log10]",
@@ -28,12 +33,16 @@ legend(x = "topleft", bty="n",
      pch = c(1, 16, 17),
      cex = 1.2)
 mtext("a", adj=-.05, cex=1.5, line=1)
-plot(slidesort_pairs/1e6, starcode_pairs/1e6, type = "b",
-     xlab = "Pairs in slidesort [millions]",
-     ylab = "Pairs in starcode [millions]",
+
+barplot(slidesort_pairs/starcode_pairs,
+     names.arg=log10(clusters),
+     xlab = "Number of clusters [log10]",
+     ylab = "Pairs found [slidesort/starcode]",
+     ylim=c(0,1),
      cex.lab = 1.3)
-abline(0,1, col=2)
+abline(1,0, col=2)
 mtext("b", adj=-.05, cex=1.5, line=1)
+
[…]

### 2014-11-03 doc/document.tex
Document: Added command-line tables.

@@
 
-\subsection{Benchmark conditions}
-All the tests were performed on a 16-core dual-processor
-Intel Xeon E5-2687W v2 system with 256~GB of DDR3-RAM at 1866~Mhz
-
 \end{methods}
@@
 \section{Results}
+
+\subsection{Benchmark conditions}
+All the tests were performed on a 16-core dual-processor
+Intel Xeon E5-2687W v2 system with 256~GB of DDR3-RAM at
+1866~Mhz. Command-line parameters were set equivalently in all
+softwares to run in single-core mode allowing up to 3 mismatches for
+input sequences of length 50. Execution options used in simulation and
+real datasets are summarized in Table \ref{tab:params1} and Table
+\ref{tab:params2}, respectively.
+
+\begin{table}[h]
+\centering
+\caption{Software execution options used in simulation benchmark.}
+%\resizebox{\columnwidth}{!}{
+\begin{tabular}{l|l}
+    \bf{software} & \bf{execution options}\\
+    \hline
+    starcode & \texttt{starcode -d3}\\
+    slidesort & \texttt{slidesort\_v2 -d 3 -t E -c DNA}\\
+    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
+    SEED & \texttt{SEED --mismatch 3}\\
+\end{tabular}
+%}
+\label{tab:params1}
+\end{table}
+
+\begin{table}[h]
+\centering
+\caption{Software execution options used in real data benchmark.}
+%\resizebox{\columnwidth}{!}{
+\begin{tabular}{l|l}
+    \bf{software} & \bf{execution options}\\
+    \hline
+    starcode & \texttt{starcode -d3}\\ 
+    slidesort & \texttt{slidesort\_v2 -d 3 -u -t E -c DNA}\\
+    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
+    SEED & \texttt{SEED --mismatch 3 --shift 3}\\ 
+\end{tabular}
+%}
+\label{tab:params2}
+\end{table}
 
@@
 
-%\begin{table}[h]
-%\centering
-%\caption{Software execution parameters (benchmark 1)}
-%\resizebox{\columnwidth}{!}{
-%\begin{tabular}{|l|l|}
-%    \hline
-%    starcode & \texttt{-d3 -t12 -s}\\ \hline
-%    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 7 -c .925 -T 12 -M 0 -r 0 -A 40}\\ \hline
-%\end{tabular}
-%}
-%\label{tab:params1}
-%\end{table}
-
 %% Biological datasets %%
@@
 All the datasets are Illumina reads.}
-\begin{tabular}{|l|l|l|l|}
-    \hline
+\begin{tabular}{l|l c r}
     \textit{dataset}   & \textit{read count} &
     \textit{read length} & \textit{type} \\ \hline
-    dataset 1 & 16,457,527   & 50  & single      \\ \hline
-    dataset 2 & 127,675,537  & 50  & single      \\ \hline
-    dataset 3 & 2,460,226    & 100  & paired-end \\ \hline
+    dataset 1 & 16,457,527   & 50  & single      \\
+    dataset 2 & 127,675,537  & 50  & single      \\
+    dataset 3 & 2,460,226    & 100  & paired-end \\
 \end{tabular}
@@
 that the software cannot be used for this dataset.}
-\begin{tabular}{|l|l|l|l|}
-    \hline
+\begin{tabular}{l|c c c}
     \textit{software}   & \textit{dataset 1} &
       \textit{dataset 2} & \textit{dataset 3} \\ \hline
-    starcode   & 171         & 2,898       & 44    \\ \hline
-    seed       & 135,959     & 60,374      & -     \\ \hline
-    sidesort   & 632,168     & $>$ 10 days & -     \\ \hline
-    rainbow    & -           & -           & 306   \\ \hline
-    cd-hit-est & $>$ 10 days & 512,591     & -     \\ \hline
+    starcode   & 171         & 2,898       & 44    \\
+    seed       & 135,959     & 60,374      & -     \\
+    sidesort   & 632,168     & $>$ 10 days & -     \\
+    rainbow    & -           & -           & 306   \\
+    cd-hit-est & $>$ 10 days & 512,591     & -     \\
 \end{tabular}
@@
 \caption{Memory usage (in Gb).}
-\begin{tabular}{|l|l|l|l|}
-    \hline
+\begin{tabular}{l|c c c}
     \textit{software}   & \textit{dataset 1} &
       \textit{dataset 2} & \textit{dataset 3} \\ \hline
-    starcode   & 9.4 & 30.9 & 5.2 \\ \hline
-    seed       & 4.1 & 53.9 & -   \\ \hline
-    sidesort   & 1.9 & 13.9 & -   \\ \hline
-    rainbow    & -   & -  & 0.5   \\ \hline
-    cd-hit-est & 3.8 & 28.5 & -   \\ \hline
+    starcode   & 9.4 & 30.9 & 5.2 \\
+    seed       & 4.1 & 53.9 & -   \\
+    sidesort   & 1.9 & 13.9 & -   \\
+    rainbow    & -   & -  & 0.5   \\
+    cd-hit-est & 3.8 & 28.5 & -   \\
 \end{tabular}

### 2014-11-04 doc/document.tex
Working on the discussion (in progress).

@@
 always have 5 times or lower read count than the canonical sequence.
-Otherwise, sequences are more likely unrelated, or both derived from
-the same canonical sequence.
+Otherwise, sequences are more likely unrelated, or both are derived
+from the same canonical sequence.
 
 For other sequence clustering problems, starcode implements a
-multi-purpose algorithm called ``sphere clustering''
-\citep{pmid23953119}. In sphere clustering, sequences are sorted by
+multi-purpose algorithm called ``sphere clustering''.
+In sphere clustering, sequences are sorted by
 frequency of occurrence. Starting from the most
@@
 belong to only one cluster.
-
-\end{methods}
-
-\section{Results}
 
@@
     slidesort & \texttt{slidesort\_v2 -d 3 -t E -c DNA}\\
-    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
-    SEED & \texttt{SEED --mismatch 3}\\
+    cd-hit 4.6.1 & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
+    seed & \texttt{SEED --mismatch 3}\\
 \end{tabular}
@@
     slidesort & \texttt{slidesort\_v2 -d 3 -u -t E -c DNA}\\
-    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
-    SEED & \texttt{SEED --mismatch 3 --shift 3}\\ 
+    cd-hit 4.6.1 & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
+    seed & \texttt{SEED --mismatch 3 --shift 3}\\ 
 \end{tabular}
@@
 \end{table}
+
+\end{methods}
+
+\section{Results}
 
@@
 be trimmed and filtered for quality or not. Input formats may be raw
-sequence, fasta or fastq (in which case starcode ignores the quality).
+sequence, FASTA or FASTQ (in which case starcode ignores the quality).
 By default, starcode outputs sequence clusters, it can also be used
@@
 from experimental errors (sequencing errors, PCR mutations
-\textit{etc}), a more general algorithm is also available for other
+\textit{etc}.), a more general algorithm is also available for other
 clustering problems (an example of which is given in section
@@
 exclusively on paired-end reads, while the other tools run on single
-reads.
+reads, for this reason all the tools could not be run no the same
+dataset.
 
@@
 most 1 indel. The number of satellites per cluster ranged from
-999900 to 900. Evaluating the exactness of slidesort on these datasets
+999900 to 900. Rainbow was not tested on this benchmark because
+the generated data is single reads. In addition, evaluating the
+exactness of slidesort 
 was problematic because the number of 3-matches in each dataset is
@@
 \centerline{\includegraphics[scale=0.47]{benchmark.pdf}}
-\caption{Accuracy of the sequence clustering and all-pairs search
-algorithms. Starcode, seed and cd-hit were bencmarked on 4 datasets
-with the same number of sequences, but different number of clusters
-(see main text). \textbf{a} Starcode identifies the correct number
-of clusters, while seed and cd-hit identify about 40 false positives
-per true positive. \textbf{b} Slidesort identifies 5-10\% less
-pairs than starcode. In both panels, the red line is first
-bisector.}
+\caption{Benchmark results on artificial datasets of known cluster
+structure (see main text). \textbf{a} Accuracy measured by the number
+of identified clusters. Starcode identifies the correct
+number of clusters, while seed and cd-hit identify about 40 false
+positives per true positive. The red line is the first bisector and
+indicates perfect results. \textbf{b} Accuracy measured by the
+number of identified pairs. Slidesort identifies 5-10\% less pairs
+than starcode. The red line indicates a ratio of 1. \textbf{c}
+Running time of the different tools. As the size of the clusters
+the running time of starcode increases but remains competitive.
+\textbf{d} Memory usage of the different tools. The memory usage
+of starcode decreases as the size of the cluster increases.}
 \label{fig:benchmark}
@@
 of 1 million sequences. The memory usage is shown in Figure
-\ref{fig:benchmark}c.
+\ref{fig:benchmark}c. The smallest memory footprint is achieved by
+slidesort and cd-hit. The memory usage of starcode is the highest
+for clusters of size 1000, but it decreases and becomes lower than
+the memory usage of seed as the size of the cluster increases.
+
 In conclusion, starcode was the only tool to achieve perfect precision
-on these datasets. It also maintains competitive performance in
-running time and memory usage.
+on these datasets. It also maintains a competitive performance in
+terms of running time and memory usage.
 
@@
 The performance on artificial data is not always in agreement with
-the performance on experimental datasets. In typical experiments,
-the sizes of the clusters may be uneven, and the reads may contain
-near constant regions that may degrade the performance.
+the performance on experimental datasets. Typical experiments present
+additional difficulties. For instance,
+the sizes of the clusters may be uneven and the reads may contain near
+constant regions that may degrade the performance of the algorithm.
 We benchmarked sequence clustering algorithms on the problem of
@@
 while dataset 2 contains 4 clusters with more than 1 million sequences.
-Dataset 3 consist of Illumina paired-end reads.
+Dataset 3 consist of paired-end reads.
 
@@
 
-The running times of starcode, seed, slidesort, rainbow and cd-hit-est
+The running times of starcode, seed, slidesort, rainbow and cd-hit
 are summarized in Table~\ref{tab:biobench_runtime}. Starcode was faster
@@
 greater, respectively. Rainbow came in second position on dataset 3,
-but it is the only other tool able cluster paired-end reads. We did not
+but it is the only other tool able to cluster paired-end reads. We did not
 record the exact running times past 10 days since this is several
@@
 than the best tool, while on dataset 2 it used significantly less
-memory than seed and approximately as much as cd-hit. A large portion
-of the memory usage of starcode is dedicated to the trie structures
-used in the poucet search, which explains why starcode is in general
-more memory-demanding than alternative tools.
+memory than seed and approximately as much as cd-hit.
 
@@
 \centering
-\caption{Memory usage (in Gb).}
+\caption{Memory usage (in GB).}
 \begin{tabular}{l|c c c}
@@
 
-%\subsection{Clustering TRIP barcodes}
-%In the course of setting up the TRIP technology in our laboratory
-%\citep{pmid23953119}, we realized the need to develop efficient
-%algorithms to cluster similar sequences. Briefly, the principle of TRIP
-%(Thousands of Reporters Integrated in Parallel) is to tag reporter
-%transcripts with random barcodes and measure the abundance of barcodes
-%in the RNA as a proxy for gene expression. There is no reference to
-%match aberrant barcodes against, because the tagging sequences are
-%unknown. Instead, barcodes are matched against each other and clustered
-%by similarity to infer canonical sequences.
-%
-%We tested the efficiency of starcode on the TRIP dataset from
-%\cite{pmid23953119}. In the experiment labeled mPGKA, we identified
-%24.1 million (91\%) barcode-containing reads out of 26.6 million,
-%consisting of approximately 223,000 and 220,000 unique barcode sequences
-%for PCR replicates 1 and 2, respectively. Following the authors, we kept
-%only the barcodes with at least 5 counts and performed clustering as
-%described: ``First we sorted barcodes according to their counts. Then,
-%for each barcode (starting from the most frequent one), we identified
-%and removed all its mutant versions, defined as barcodes within a Hamming
-%distance of 2.'' We implemented this method, here on referred to as the
-%``sequential algorithm'' with the function call
-%\texttt{stringDist(method='lv', maxDist=2)} from R the package
-%\texttt{stringDist} \citep{R}.  When replacing the Hamming distance by
-%the Levenshtein
-%distance, starcode produced exactly the same output as the sequential
-%algorithm. The running time of starcode averaged over the replicates was
-%2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which
-%represents a 75-fold speedup.
-%
-%The performance of the sequential algorithm relies on the arbitrary
-%exclusion of reads with less than 5 counts, the main purpose of which
-%is to reduce the computational burden. When all barcodes were kept,
-%the average running time of starcode was 10.27 seconds, \textit{versus}
-%$>$ 6 hours for the sequential algorithm. Using the Hamming distance,
-%as the authors originally suggested, decreased the average running time
-%of the sequential algorithm to about 3 hours and 30 minutes (and
-%the output differed from that of starcode). Note that in all the cases
-%mentioned above, starcode was run with a single core to compare the
-%algorithms based on similar computer resources. In conclusion, the
-%barcode clustering problem can be simplified by various tricks, but
-%starcode brings down the running time to nearly instantaneous, and
-%thereby obviates the need for such arbitrary heuristics.
-
 \subsection{Identifying enriched sequence motifs}
 \label{section:motifs}
+As a sequence clustering algorithm, starcode can also be used for
+other applications, such as the identification of enriched motifs.
 Sequence motifs are thought to play an important role in DNA metabolism.
@@
 computationally difficult for long motifs (above 12-13 nucleotides)
-because of the combinatorial scaling. But as motifs become longer, the
-problem of identifying abundant inexact matches becomes similar to
-barcode clustering. We reasoned that starcode could also be used for
-the task of identifying biologically meaningful sequence motifs.
+because of the combinatorial scaling.
+
+%But as motifs become longer, the
+%problem of identifying abundant inexact matches becomes similar to
+%barcode clustering. We reasoned that starcode could also be used for
+%the task of identifying biologically meaningful sequence motifs.
 
@@
 consisting of 2.77 million unique sequences. Clustering the 12-mers
-with starcode within a Levenshtein distance of 2 took less than 45
-seconds with 12 threads. We identified the known DNA uptake sequence
+with starcode within a Levenshtein distance of 2,
+%took less than 45
+%seconds with 12 threads. We identified the known DNA uptake sequence
+we identified the known DNA uptake sequence
 of \textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant
@@
 Levenshtein distance 2. The score of the most enriched 10-mer is thus
-the sum of the ranks of all 10-mers within this distance. Clustering
-the 6.3 million extracted 10-mers with 12 threads took about 20 seconds
-for each replicate.  The most enriched 10-mers were AGGACACGGA,
+the sum of the ranks of all 10-mers within this distance.
+%Clustering
+%the 6.3 million extracted 10-mers with 12 threads took about 20 seconds
+%for each replicate.  The most enriched 10-mers were AGGACACGGA,
+Among the 6 replicates, the most enriched 10-mers were AGGACACGGA,
 AGGACACGGA, AGGACGGAGG, AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except
@@
 \section{Discussion and conclusion}
-Through the parallel poucet search algorithm, starcode implements an
-exact sequence clustering algorithm that can be faster than popular
-heuristics. By design, starcode is tailored to process high throughput
-sequencing data on multi-core platforms. Our benchmark shows that
+Starocde is a solid algorithm for sequence clustering based on
+all-pairs matching. It achieves high precision, and on experimental
+datasets it can be faster than popular heuristics. By design,
+starcode is tailored to process high throughput sequencing data
+on multi-core platforms.
+Our benchmark shows that
 starcode achieves perfect clustering on short random sequences in less

### 2014-11-04 doc/plot-benchmark.R

@@ -12,7 +12,7 @@ cd.hit_time = c(16, 14, 13, 9)
 slidesort_time = c(3*60+38, 36*60+47, 384*60+38, 2369*60+1)
 starcode_mem = c(2606224, 2420160, 1733536, 1383892)
 slidesort_mem = c(388912, 415732, 410948, 371692)
-seed_mem = c(371692, 2320388, 2199632, 2071132)
+seed_mem = c(2361448, 2320388, 2199632, 2071132)
 cd.hit_mem = c(260992, 249052, 247916, 247660)
 # Plot
 pdf("benchmark.pdf")

### 2014-11-04 doc/document.tex
Finished discussion.

@@
 starcode is tailored to process high throughput sequencing data
-on multi-core platforms.
-Our benchmark shows that
-starcode achieves perfect clustering on short random sequences in less
-time than what is considered acceptable for heuristic searches. We also
-show that starcode outperforms next generation read mappers in this
-context. The software tools used for this benchmark are usually not
-used for barcode or random sequence clustering. In this respect,
-starcode fills a need arising from the development of barcoding
-technologies.
+on multi-core platforms with sufficient amount of memory. Due to its
+superior precision and faster running time, it fills a gap among
+available software, by allowing to take full advantage of middle
+to high end hardware.
 
-The speed of starcode also makes it useful for other clustering
-tasks, such as identifying enriched motifs in microbial genomes and
-in experimental data. Here we have given two examples of such
+The speed and precision of starcode also makes it useful for other
+clustering tasks, such as identifying enriched motifs in microbial
+genomes and in experimental data. Here we have given two examples of such
 applications. In the first, we recover a known enriched 12-mer in the
@@
 
-The current version of starcode has been primarily optimized for speed.
-The memory footprint depends on the number of sequences to cluster
-(because the sequences of the input set are loaded in memory as a set of
-of tries) and on the mean number of matches per sequence. Every match is
-stored in memory until the clustering phase, which may represent a large
-overhead if the dataset is dense. As counterintuitive as it may seem,
-long queries will usually impose a lower memory footprint because the
-matches between the sequences are more sparse. We have shown that the
-running time will also be shorter thanks to the lookup table
-search (Figure \ref{fig:perf}c).
-
-Perhaps the most surprising element of this study is that an exact
-algorithm can compete with extremely fast heuristics. This will no
-longer be true when clustering divergent sequences because the Levenshtein
-distance will have to be increased, leading to exponentially longer
-running times (Figure \ref{fig:perf}b). However, for the imporant
-practical case that the divergence is driven by sequencing errors,
-starcode illustrates that there is still room for algorithmic innovations
-that can outperform heuristics. The idea of the poucet search seems
-simple in retrospect, yet it is a powerful way to tap into the data
-structuration provided by string sorting. This principle could find some
-applications in algorithms used in various fields.
+One of the reasons why starcode appears to be faster than alternative
+tools is that it is designed to cluster relatively similar sequences.
+When clustering related but
+divergent sequences, the Levenshtein distance will have to be increased,
+leading to exponentially longer running times (Figure \ref{fig:perf}b).
+However, for the imporant practical case of correcting errors introduced
+by sequencing, starcode illustrates that there is still room for
+developing algorithms that are both faster and more accurate than
+the current state of the art.
 

### 2014-11-05 doc/document.tex
Updated manuscript.

@@
 to performing a pairwise comparison of all the barcodes,
-which is unfeasible for excessive computationally complexity.
+which is unfeasible for excessive computational complexity.
 
@@
 and the PacBio platform has a 15\% error rate consisting of insertions
-and deletions \citep{pmid19023044}. The throughput of such technologies
+and deletions \citep{pmid19023044}. The enormous throughput of such technologies
 has recently created additional needs to develop efficient error
@@
 the graph construction phase. Matching is carried out by lossless
-filtration, followed by a trie search.  The novelty of the algorithm
-is the poucet strategy, which uses the redundancy of alphabetically
-sorted sequences to avoid unnecessary recomputations and gain speed.
+filtration, followed by an exhaustive search on the branches of a
+prefix trie.  The novelty of the algorithm is the poucet strategy,
+which uses the redundancy of alphabetically sorted sequences to avoid
+unnecessary recomputations and gain speed.
 
 In this article we present and benchmark starcode. We show that on
-real biological datasets, starcode is faster than existing sequence
-clustering software. Even though starcode was designed for error
-correction, we show that it can be used for other problems. As an
-illustration, we use it to identify enriched motifs in a bacterial
-genome and in protein-RNA interaction experiments.
+real biological datasets, starcode is orders of magnitude faster than
+existing sequence clustering software. Even though starcode was
+designed for error correction, we also show that it can be used for
+other problems. As an illustration, we use it to identify enriched
+motifs in a bacterial genome and in protein-RNA interaction
+experiments.
 
@@
 search. To this end, starcode uses a partition
-approach similar to that described by \cite{WuManber1992}. If the
-query is partitioned in $\tau+1$ segments, every $\tau$-match
-contains at least one of those (assuming that all the segments
-have length at least $\tau$). Indeed, there are at most $\tau$
-differences between the query and the match, so at least one segment
-is unmodified. In the $\tau$-match, the position of the shared
-segment is at most $\tau$ nucleotides on the left or on the right
-of its position in the query.
+approach similar to that described by \cite{WuManber1992}. The query
+is initially partitioned in $\tau+1$ segments. Assuming that all the
+segments have length at least $\tau$, then every $\tau$-match present
+in the database will contain at least a verbatim copy of one of the
+query segments. Indeed, there are at most $\tau$ editions between
+the query and the match to be distributed in $\tau+1$ regions, so at
+least one segment is unmodified. Due to potential insertions and
+deletions in the preceding segments, the shared segment may be shifted
+up to $\tau$ nucleotides on the left (all insertions) or on the right
+(all deletions) from its original position in the query.
 
-These observations are the basis of a filtration method that proceeds
-as follows. Every time a sequence is added to the trie, its segments
-are added to $\tau+1$ different indexes. More specifically, the
-first $\tau$ nucleotides of the sequence are removed, and the rest
-of the sequence is partitioned in $\tau+1$ segments of sizes
-differing by at most 1 (the longer segments always in 3' for
-consistency). The first fragments are added to the first index,
-the second fragments to the second index \textit{etc}.
+These observations are the basis of a filtration method with 100\%
+specificity. More precisely, the segments are defined as follows:
+the first $\tau$ nucleotides of the sequence are removed, and the rest
+of the sequence is partitioned in $\tau+1$ segments of sizes differing
+by at most 1 (the longer segments always in 3' for consistency). Every
+time a sequence is added to the trie, it is partitioned and its
+segments are added to $\tau+1$ different indexes. The first fragments
+are added to the first index, the second fragments to the second index,
+\textit{etc}. Before the search, the query is partitioned in the same
+way and its segments are looked up in the indexes. In case no match is
+found, this query has no $\tau$-match in the current database,
+therefore the trie search is omitted. Conversely, if at least one
+segment is found, the trie search must performed.
 
-Before the search, segments of the query are looked up in
-the indexes, and in case no match is found, the trie search is
-omitted because this query has no $\tau$-match. If at least one
-segment is found, the trie search is performed. As mentioned above,
-segments shared between the query and a $\tau$-match can be shifted
-up to $\tau$ nucleotides. For this reason, shifted segments are
-looked up in the indexes according to the following scheme ensuring
-that no match can be missed.
-(Figure \ref{fig:filtration}). The rightmost segment is looked up
-in the $\tau$+1-th index, the second rightmost segment and the
-contiguous segments shifted by 1 nucleotides are are looked up in the
-$\tau$-th index and so on until the first segment and its
-contiguous segments shifted by up to $\tau$ nucleotides are
-looked up in the first index.
+As mentioned above, segments shared between the query and a
+$\tau$-match may be found shifted up to $\tau$ nucleotides. For this
+reason, shifted segments of the query are looked up in the indexes
+according to the scheme of Figure \ref{fig:filtration}, which ensures
+that no match can be missed: the rightmost segment is looked up in the
+$\tau$+1-th index, the second rightmost segment and the contiguous
+segments shifted by 1 nucleotides are are looked up in the $\tau$-th
+index and so on, until the first segment and its contiguous segments
+shifted by up to $\tau$ nucleotides are looked up in the first index.
 
@@
     \State \textsc{sort} sequences alphabetically
-    \State $k \gets$ \textsc{compute} lookup word lengths
+    \State $k \gets$ \textsc{compute} filter segment lengths
     \State $trie \gets$ \textsc{create} an empty trie of height $height$
@@
     \State $seq \gets$ \textsc{get} next sequence
-    \If{at least one $k$-mer of $seq$ is in the lookup table}
+    \If{at least one $k$-mer of $seq$ is in the filter index}
     \State $seed \gets$ \textsc{length} of shared prefix
@@
     \State \textsc{insert} $seq$ path in $trie$
-    \State \textsc{insert} $seq$ $k$-mers into the lookup table
+    \State \textsc{insert} $seq$ $k$-mers into the filter index
     \EndFor
@@
 Queries are sorted and partitioned in contiguous blocks.
-The matching step then proceeds in two phases. In the first, a
+The matching step then proceeds in two phases. In the build phase, a
 distinct trie is built from the sequences of each block according
 to the algorithm described above. In the second, all the sequence
-blocks are queried against the tries. If the queries are partitioned
-in $N$ blocks, the first phase consists of $N$ seek and construct
-jobs, wile the second consists of $N(N-1)/2$ query jobs. In each
-phase, the jobs show little dependency on each other, so the
-matching algorithm can be efficiently parallelized provided $N$
-is larger than the number of independent threads.
+blocks are queried against all the other tries. If the queries are
+partitioned in $N$ blocks, the first phase consists of $N$ seek and
+construct jobs, wile the second consists of $N(N-1)/2$ query jobs. In
+each phase, the jobs show no dependency on each other, so the
+matching algorithm can be efficiently parallelized provided $N$ is
+larger than the number of independent threads.
 
@@
 softwares to run in single-core mode allowing up to 3 mismatches for
-input sequences of length 50. Execution options used in simulation and
-real datasets are summarized in Table \ref{tab:params1} and Table
-\ref{tab:params2}, respectively.
+input sequences of length 50. Tables \ref{tab:params1} and
+\ref{tab:params2} summarize the execution options used in simulation
+and real data sets, respectively.
 
@@
 \begin{tabular}{l|l}
-    \bf{software} & \bf{execution options}\\
+    \bf{software} & \bf{command-line options}\\
     \hline
@@
     slidesort & \texttt{slidesort\_v2 -d 3 -t E -c DNA}\\
-    cd-hit 4.6.1 & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
+    cd-hit-est & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
     seed & \texttt{SEED --mismatch 3}\\
@@
     slidesort & \texttt{slidesort\_v2 -d 3 -u -t E -c DNA}\\
-    cd-hit 4.6.1 & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
+    cd-hit-est & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
     seed & \texttt{SEED --mismatch 3 --shift 3}\\ 
+    rainbow & \texttt{rainbow cluster -m 3}
 \end{tabular}
@@
 for paired-end). Sequences may be of variable length, they may
-be trimmed and filtered for quality or not. Input formats may be raw
-sequence, FASTA or FASTQ (in which case starcode ignores the quality).
-By default, starcode outputs sequence clusters, it can also be used
-to filter out redundant sequences from input files. By default,
-clustering is performed under the assumption that divergence occurs
-from experimental errors (sequencing errors, PCR mutations
-\textit{etc}.), a more general algorithm is also available for other
-clustering problems (an example of which is given in section
-\ref{section:motifs}).
+be trimmed and filtered for quality or not. File formats compatible
+with starcode are: raw sequence, raw sequence with count, FASTA or
+FASTQ (in which case starcode ignores the quality). Starcode has two
+output formats: cluster and non-redundant. In cluster mode, starcode
+returns detailed information of the clustering results, i.e. canonical
+sequences, cluster sizes and the complete list of their constituent
+sequences. In non-redundant mode, only the canonical sequences are
+printed. The second mode is especially useful to filter out redundant
+sequences from input files. By default, clustering is performed under
+the assumption that divergence occurs from experimental errors
+(sequencing errors, PCR mutations \textit{etc}.), a more general
+algorithm is also available for other clustering problems (an example
+of which is given in section \ref{section:motifs}).
 
@@
 configuration is a set of 1,000,000 sequences of length 40, running
-on 1 thread and with a maximum Levenshtein distance of 3. Only one
-parameter is modified whereas the others were kept constant.
+on 1 thread and with a maximum Levenshtein distance of 3. In each
+test, only one parameter is modified while the others are kept constant.
 
@@
 that the input reading and clustering steps are brief but not parallel,
-the second is that there is insufficient memory bandwidth to satisfy
-the increased demand of memory accesses due to hardware limitations.
+the second is due to hardware limitations, i.e. there is insufficient
+memory bandwidth to satisfy the increased demand of parallel memory accesses.
 
@@
 intensive step of the sequence clustering problem. Rainbow runs
-exclusively on paired-end reads, while the other tools run on single
+exclusively on paired-end reads, whereas the other tools run on single
 reads, for this reason all the tools could not be run no the same
@@
 identify approximately 40 false clusters per true cluster on all
-the datasets. We also observed that slidesort found 5-10\% less
-3-matches than starcode on all the datasets. We were surprised by
+the datasets (Figure \ref{fig:benchmark}a). We also observed that
+slidesort found 5-10\% less 3-matches than starcode on all the
+datasets (Figure \ref{fig:benchmark}b). We were surprised by 
 this result because slidesort is claimed to be an exact algorithm.
 However, this was clearly not the case when we ran additional tests
-on smaller datasets where naive pairwise comparisons is feasible
+on smaller datasets where naive pairwise comparisons are feasible
 (the starcode repository contains such a dataset along with the
@@
 The running time of the different tools as a function of the size
-of the clusters is shown on Figure \ref{fig:benchmark}c. For seed
-and cd-hit it decreases as the cluster size increases, whereas
-for starcode and slidesort it increases. In spite of this trend,
-the performance of starcode remains competitive, even for clusters
-of 1 million sequences. The memory usage is shown in Figure
+of the clusters is shown on Figure \ref{fig:benchmark}c. The running
+time of slidesort and starcode show a linear and sub-linear increase,
+respectively. Seed and cd-hit run approximately in constant time
+regardless of the cluster size. In spite of this trend, the
+performance of starcode remains competitive, even for clusters of 1
+million sequences. The memory usage is shown in Figure
 \ref{fig:benchmark}c. The smallest memory footprint is achieved by
-slidesort and cd-hit. The memory usage of starcode is the highest
-for clusters of size 1000, but it decreases and becomes lower than
-the memory usage of seed as the size of the cluster increases.
+slidesort and cd-hit, with a maximum difference of an order of
+magnitude with respect to the other tools. Comparing with
+slidesort is not completely fair since it does not hold in memory the
+full graph necessary for clustering, which is by far the most memory
+demanding structure. The memory usage of starcode is the highest for
+clusters of size 1000, but it decreases and becomes lower than the
+memory usage of seed as the size of the cluster increases.
 
 In conclusion, starcode was the only tool to achieve perfect precision
-on these datasets. It also maintains a competitive performance in
-terms of running time and memory usage.
+on these datasets at a price of increased memory
+footprint. Considering the exactness of the output, starcode maintains
+a competitive performance in terms of running time.
 
@@
 the sizes of the clusters may be uneven and the reads may contain near
-constant regions that may degrade the performance of the algorithm.
-We benchmarked sequence clustering algorithms on the problem of
-clustering TRIP barcodes \citep{pmid23953119}. Briefly, the principle
-of TRIP (Thousands of Reporters Integrated in Parallel) is to tag
-reporter transcripts with random barcodes and measure the abundance
-of barcodes in the RNA as a proxy for gene expression. There is no
-reference to match aberrant barcodes against, because the tagging
-sequences are unknown.
+constant regions that usually degrade the performance of filter-based
+algorithms. We benchmarked sequence clustering algorithms on the
+problem of clustering TRIP barcodes \citep{pmid23953119}. Briefly, the
+principle of TRIP (Thousands of Reporters Integrated in Parallel) is
+to tag reporter transcripts with random barcodes and measure the
+abundance of barcodes in the RNA as a proxy for gene expression. There
+is no reference to match aberrant barcodes against, because the
+tagging sequences are unknown.
 
@@
 while dataset 2 contains 4 clusters with more than 1 million sequences.
-Dataset 3 consist of paired-end reads.
+Dataset 3 consists of paired-end reads.
 
@@
     dataset 2 & 127,675,537  & 50  & single      \\
-    dataset 3 & 2,460,226    & 100  & paired-end \\
+    dataset 3 & 2,460,226    & 50+50  & paired-end \\
 \end{tabular}

### 2014-11-06 doc/document.tex
Updated benchmark with TRiP data and results text.

@@
 of the clusters is shown on Figure \ref{fig:benchmark}c. The running
-time of slidesort and starcode show a linear and sub-linear increase,
+time of slidesort and starcode show a linear and sub-linear trends,
 respectively. Seed and cd-hit run approximately in constant time
-regardless of the cluster size. In spite of this trend, the
+regardless of the cluster size. In spite of this result, the
 performance of starcode remains competitive, even for clusters of 1
@@
 slidesort and cd-hit, with a maximum difference of an order of
-magnitude with respect to the other tools. Comparing with
+magnitude with respect to the other tools. Note that the comparison with
 slidesort is not completely fair since it does not hold in memory the
-full graph necessary for clustering, which is by far the most memory
-demanding structure. The memory usage of starcode is the highest for
+full graph necessary for clustering (which is by far the most memory
+demanding structure). The memory usage of starcode is the highest for
 clusters of size 1000, but it decreases and becomes lower than the
@@
 The basic properties of the datasets used for benchmarking are
-summarized in Table~\ref{tab:biobench_properties}. Datasets 1 and 2
-consist of Illumina single reads; they differ by the total read count
-and by the empirical cluster sizes. According to the output of starcode,
-the largest clusters of dataset 1 contain approximately 2,000 sequences,
-while dataset 2 contains 4 clusters with more than 1 million sequences.
-Dataset 3 consists of paired-end reads.
-
+summarized in Table~\ref{tab:biobench_properties}. Dataset 1
+(SRR950457) has been pre-processed to extract the barcode and remove
+the constant part of the reads. Only barcodes between 15 and 17
+nucleotides were included in the file. Dataset 2 (SRRHENCHI) consists
+of raw Illumina single reads. These datasets differ by the read size,
+the total read count and by the empirical cluster sizes. According to
+the output of starcode, the largest clusters of dataset 1 contain
+approximately 70,000 sequences, whereas dataset 2 contains 4 clusters
+with more than 1 million sequences. Dataset 3 (SRR950477) has been
+included to benchmark starcode against rainbow in paired-end
+clustering mode.
 \begin{table}[h]
@@
     \textit{read length} & \textit{type} \\ \hline
-    dataset 1 & 16,457,527   & 50  & single      \\
-    dataset 2 & 127,675,537  & 50  & single      \\
-    dataset 3 & 2,460,226    & 50+50  & paired-end \\
+    SRR950457 & 6,542,309    & 16     & single     \\
+    SRRxxxxxx & 127,675,537  & 50     & single     \\
+    SRR950477 & 2,460,226    & 50+50  & paired-end \\
 \end{tabular}
@@
 The running times of starcode, seed, slidesort, rainbow and cd-hit
-are summarized in Table~\ref{tab:biobench_runtime}. Starcode was faster
-than the other tools on all the datasets. Seed came in second position
-on datasets 1 and 2, with a running time approximately 800 and 20 times
-greater, respectively. Rainbow came in second position on dataset 3,
-but it is the only other tool able to cluster paired-end reads. We did not
-record the exact running times past 10 days since this is several
-orders of magnitude higher than the running time of starcode.
+are summarized in Table~\ref{tab:biobench_runtime}. We accommodated the
+distance threshold for the first dataset to compensate for the reduced
+sequence length. Both starcode and slidesort were executed with
+the option '-d 2' and the identity for cd-hit was set to '-c 0.85'. We
+were not able to run seed on dataset 1 due to limitations on the
+minimum input sequence length. Starcode was faster than the other
+tools on all the datasets. CD-HIT and seed came in second position
+with a running time approximately 35 and 20 times greater on datasets
+1 and 2, respectively. Rainbow was nearly an order of magnitude slower
+in the job of clustering paired-end reads. We did not record the exact
+running times past 10 days since this is several orders of magnitude
+higher than the running time of starcode.
 
@@
 \begin{tabular}{l|c c c}
-    \textit{software}   & \textit{dataset 1} &
-      \textit{dataset 2} & \textit{dataset 3} \\ \hline
-    starcode   & 171         & 2,898       & 44    \\
-    seed       & 135,959     & 60,374      & -     \\
-    sidesort   & 632,168     & $>$ 10 days & -     \\
+    \textit{software}   & \textit{SRR950457} &
+      \textit{SRRxxxxxx} & \textit{SRR950477} \\ \hline
+    starcode   & 5           & 2,898       & 44    \\
+    seed       & -           & 60,374      & -     \\
+    sidesort   & 4,055       & $>$ 10 days & -     \\
     rainbow    & -           & -           & 306   \\
-    cd-hit-est & $>$ 10 days & 512,591     & -     \\
+    cd-hit-est & 170         & 512,591     & -     \\
 \end{tabular}
@@
 shown in Table~\ref{tab:biobench_memory}. The values represent the
-peak memory usage throughout the run on the datasets described above.
-On datasets 1 and 3, starcode had a significantly larger memory usage
-than the best tool, while on dataset 2 it used significantly less
-memory than seed and approximately as much as cd-hit.
+peak memory usage throughout the run on the datasets described
+above. On short reads (dataset 1), starcode outperforms the other
+tools taking advantage of the trie compaction. On dataset 3, starcode
+had a significantly larger memory usage than rainbow. Starcode and
+cd-hit used the same amount of memory on dataset 2. Both needed twice
+as much memory as slidesort, which runs in advantage because it does
+not need to keep the complete graph during the all-pair comparison
+step to perform future clustering.
 
@@
       \textit{dataset 2} & \textit{dataset 3} \\ \hline
-    starcode   & 9.4 & 30.9 & 5.2 \\
-    seed       & 4.1 & 53.9 & -   \\
-    sidesort   & 1.9 & 13.9 & -   \\
+    starcode   & 0.65 & 30.9 & 5.2 \\
+    seed       & -    & 53.9 & -   \\
+    sidesort   & 1.30 & 13.9 & -   \\
     rainbow    & -    & -    & 0.5 \\
-    cd-hit-est & 3.8 & 28.5 & -   \\
+    cd-hit-est & 0.80 & 28.5 & -   \\
 \end{tabular}

### 2014-11-06 doc/document.tex
Manuscript: corrected some typos.

@@
 exclusively on paired-end reads, whereas the other tools run on single
-reads, for this reason all the tools could not be run no the same
+reads, for this reason all the tools could not be run on the same
 dataset.
@@
 999900 to 900. Rainbow was not tested on this benchmark because
-the generated data is single reads. In addition, evaluating the
+the generated data are single reads. In addition, evaluating the
 exactness of slidesort 
@@
 of the clusters is shown on Figure \ref{fig:benchmark}c. The running
-time of slidesort and starcode show a linear and sub-linear trends,
+time of slidesort and starcode show linear and sub-linear trends,
 respectively. Seed and cd-hit run approximately in constant time
@@
 the constant part of the reads. Only barcodes between 15 and 17
-nucleotides were included in the file. Dataset 2 (SRRHENCHI) consists
+nucleotides were included in the file. Dataset 2 (SRRxxxxxx) consists
 of raw Illumina single reads. These datasets differ by the read size,
@@
 were not able to run seed on dataset 1 due to limitations on the
-minimum input sequence length. Starcode was faster than the other
+minimum sequence length. Starcode was significantly faster than the other
 tools on all the datasets. CD-HIT and seed came in second position
@@
 \begin{tabular}{l|c c c}
-    \textit{software}   & \textit{dataset 1} &
-      \textit{dataset 2} & \textit{dataset 3} \\ \hline
+    \textit{software}   & \textit{SRR950457} &
+      \textit{SRRxxxxxx} & \textit{SRR950477} \\ \hline
     starcode   & 0.65 & 30.9 & 5.2 \\

### 2014-11-07 doc/document.tex
Manuscript: Updated HC dataset name.

@@
 the constant part of the reads. Only barcodes between 15 and 17
-nucleotides were included in the file. Dataset 2 (SRRxxxxxx) consists
+nucleotides were included in the file. Dataset 2 (PRJEB7686) consists
 of raw Illumina single reads. These datasets differ by the read size,
@@
     SRR950457 & 6,542,309    & 16     & single     \\
-    SRRxxxxxx & 127,675,537  & 50     & single     \\
+    PRJEB7686 & 127,675,537  & 50     & single     \\
     SRR950477 & 2,460,226    & 50+50  & paired-end \\
@@
     \textit{software}   & \textit{SRR950457} &
-      \textit{SRRxxxxxx} & \textit{SRR950477} \\ \hline
+      \textit{PRJEB7686} & \textit{SRR950477} \\ \hline
     starcode   & 5           & 2,898       & 44    \\
@@
     \textit{software}   & \textit{SRR950457} &
-      \textit{SRRxxxxxx} & \textit{SRR950477} \\ \hline
+      \textit{PRJEB7686} & \textit{SRR950477} \\ \hline
     starcode   & 0.65 & 30.9 & 5.2 \\
@@
 \section{Discussion and conclusion}
-Starocde is a solid algorithm for sequence clustering based on
+Starcode is a solid algorithm for sequence clustering based on
 all-pairs matching. It achieves high precision, and on experimental
@@
 The use of a clustering method to tackle this problem is unusual, but it
-it illustrates the potential advantages of distance-based approaches.
+illustrates the potential advantages of distance-based approaches.
 

### 2014-11-07 doc/document.tex
Working on document.tex (preparing for merge conflicts).

--- 2014-11-06 doc/document.tex
@@
 of those applications, sequencing errors need to be corrected.
-This is particularly important when using random DNA barcodes to
-trace and quantify transcripts or lineages, since spurious barcodes
-introduce inexistent entities. The reference population of random
-barcodes is unknown and in this case, error correction amounts
-to performing a pairwise comparison of all the barcodes,
-which is unfeasible for excessive computational complexity.
+This is particularly important when sequencing reads from an
+unknwon reference such as random DNA barcodes.
+In this case, error correction can be
+done by performing a pairwise comparison of all the barcodes,
+which is computationally complex problem.
 
 \section{Results:}
-Here we address this problem and describe an exact algorithm to
+Here we address this challenge and describe an exact algorithm to
 determine which pairs of sequences lie within a given Levenshtein
 distance. For error correction or redundancy reduction purposes,
-matched pairs are then merged into clusters represented by a
-canonical sequence. The effiency of starcode is attributable to
+matched pairs are then merged into clusters of similar sequences.
+The effiency of starcode is attributable to
 the poucet search, a novel implementation of the Needleman-Wunsch
@@
 
-Sequencing technologies never achieve perfect precision. For instance,
-the Illumina platform \citep{pmid16056220} has a 1-2\% error rate
-consisting of substitutions \citep{pmid18660515, pmid21576222}
+All sequencing technologies have a certain degree of imprecision.
+For instance, the Illumina platform \citep{pmid16056220} has a 1-2\%
+error rate consisting of substitutions \citep{pmid18660515, pmid21576222}
 and the PacBio platform has a 15\% error rate consisting of insertions
-and deletions \citep{pmid19023044}. The enormous throughput of such technologies
-has recently created additional needs to develop efficient error
-correction algorithms.
+and deletions \citep{pmid19023044}. The enormous throughput of such
+technologies has recently created additional needs for developing
+efficient error correction algorithms.
 
@@
 
-Here we describe a sequence clustering algorithm focused on error
-correction. The algorithm is called ``starcode'' in reference to 
-clusters of random barcodes, which typically have a star shape.
-Starcode is based on all-pairs search, \textit{i.e.} all the pairs
+Here we describe a sequence clustering algorithm called ``starcode''
+in reference to clusters of random barcodes, which typically have a star
+shape.  Starcode is based on all-pairs search, \textit{i.e.} all the pairs
 of sequences below a given Levenshtein distance are identified during
@@
 be trimmed and filtered for quality or not. File formats compatible
-with starcode are: raw sequence, raw sequence with count, FASTA or
-FASTQ (in which case starcode ignores the quality). Starcode has two
-output formats: cluster and non-redundant. In cluster mode, starcode
+with starcode are raw sequence, raw sequence with count, FASTA or
+FASTQ (in which case starcode ignores the quality). Starcode either
 returns detailed information of the clustering results, i.e. canonical
 sequences, cluster sizes and the complete list of their constituent
-sequences. In non-redundant mode, only the canonical sequences are
-printed. The second mode is especially useful to filter out redundant
+sequences. Alternatively, only the canonical sequences are
+printed, which is useful to filter out redundant
 sequences from input files. By default, clustering is performed under
 the assumption that divergence occurs from experimental errors
-(sequencing errors, PCR mutations \textit{etc}.), a more general
+(sequencing errors, PCR mutations \textit{etc}.) and a more general
 algorithm is also available for other clustering problems (an example
@@
 the generated data are single reads. In addition, evaluating the
-exactness of slidesort 
-was problematic because the number of 3-matches in each dataset is
-not known (pairs of statellites in the same cluster may be 3-matches
-or not). For this reason we only compared the number of pairs found
-by starcode with the number of pairs found by slidesort. The 
-outcome of the test is summarized in
-Figure \ref{fig:benchmark}.
+exactness of slidesort was problematic because the number of
+3-matches in each dataset is not known (pairs of statellites
+in the same cluster may be 3-matches or not). For this reason
+we only compared the number of pairs found by starcode with
+the number of pairs found by slidesort. The outcome of the test
+is summarized in Figure \ref{fig:benchmark}.
 
@@
 on smaller datasets where naive pairwise comparisons are feasible
-(the starcode repository contains such a dataset along with the
-instructions to reproduce our observations, see
+(more information is available on the starcode repository the, see
 http://github.com/gui11aume/starcode/misc).
@@
 slidesort is not completely fair since it does not hold in memory the
-full graph necessary for clustering (which is by far the most memory
-demanding structure). The memory usage of starcode is the highest for
-clusters of size 1000, but it decreases and becomes lower than the
-memory usage of seed as the size of the cluster increases.
-
+full graph necessary for clustering. The memory usage of starcode is
+the highest for clusters of size 1000, but it decreases and becomes
+lower than the memory usage of seed as the size of the cluster increases.
 In conclusion, starcode was the only tool to achieve perfect precision
@@
 We would like to thank Maria Chatzou for her precious feedback on the
-preliminary version of this manuscript.
+preliminary version of this manuscript and Heng-Chang Chen for
+performing the Drosophila TRIP experiments used for benchmarking.
 

### 2014-11-07 doc/document.tex
Manuscript ready for resubmission.

@@
 We show the basic performance and scalability of starcode on a
-dataset of pseudo random sequences (Figure \ref{fig:perf}). The standard
+dataset of pseudorandom sequences (Figure \ref{fig:perf}). The standard
 configuration is a set of 1,000,000 sequences of length 40, running
@@
 All the datasets are Illumina reads.}
-\begin{tabular}{l|l c r}
+\begin{tabular}{l|c c c}
     \textit{dataset}   & \textit{read count} &
     \textit{read length} & \textit{type} \\ \hline
-    SRR950457 & 6,542,309    & 16     & single     \\
+    SRR950457 & 6,542,309    & $16\pm 1$ & single     \\
     PRJEB7686 & 127,675,537  & 50        & single     \\
-    SRR950477 & 2,460,226    & 50+50  & paired-end \\
+    SRR950477 & 2,460,226    & 100+100     & paired-end \\
 \end{tabular}
@@
 sequence length. Both starcode and slidesort were executed with
-the option '-d 2' and the identity for cd-hit was set to '-c 0.85'. We
+the option `\texttt{-d 2}' and the identity for cd-hit was set to
+`\texttt{-c 0.85}'. We
 were not able to run seed on dataset 1 due to limitations on the
 minimum sequence length. Starcode was significantly faster than the other
-tools on all the datasets. CD-HIT and seed came in second position
+tools on all the datasets. Seed and cd-hit came in second position
 with a running time approximately 35 and 20 times greater on datasets
@@
 \caption{Running time (in seconds) of the software on three biological
-datasets. Runs longer than 10 days were interrupted. A dash indicates
-that the software cannot be used for this dataset.}
+datasets. Exact running time was not recordeed past 10 days.
+A dash indicates that the software cannot be used for this dataset.}
 \begin{tabular}{l|c c c}
@@
 had a significantly larger memory usage than rainbow. Starcode and
-cd-hit used the same amount of memory on dataset 2. Both needed twice
-as much memory as slidesort, which runs in advantage because it does
-not need to keep the complete graph during the all-pair comparison
-step to perform future clustering.
+cd-hit used similar amount of memory on dataset 2. Both needed twice
+as much memory as slidesort, which has the advantage of not storing
+the complete graph during the all-pair comparison.
 
@@
 
+It is somehwat surprising that starcode is significantly faster
+than competing tools on experimental datasets, whereas seed and cd-hit
+are faster on artificial datasets. Starcode was developed ground up
+from TRIP experimental datasets and the poucet search was selected for
+giving the best empirical results. We speculate that the trie structure
+benefits from the entropy deficit that is typically observed in
+experimental data versus pseudorandom reads.
+
 The speed and precision of starcode also makes it useful for other
@@
 preliminary version of this manuscript and Heng-Chang Chen for
-performing the Drosophila TRIP experiments used for benchmarking.
+performing the \textit{Drosophila} TRIP experiments.
 
 \paragraph{Funding\textcolon}
-The research leading to these results has received funding from the
-Government of Catalonia (Dept. of Economy and Knowledge) and the Spanish
+This research was funded by the Government of Catalonia (Dept. of
+Economy and Knowledge) and the Spanish
 Ministry of Economy and Competitiveness (Centro de Excelencia Severo Ochoa

### 2015-01-18 doc/document.tex
Added software versions.

@@
     \hline
-    starcode & \texttt{starcode -d3}\\
-    slidesort & \texttt{slidesort\_v2 -d 3 -t E -c DNA}\\
-    cd-hit-est & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
-    seed & \texttt{SEED --mismatch 3}\\
+    starcode-1.0 & \texttt{starcode -d3}\\
+    slidesort-2 & \texttt{slidesort\_v2 -d 3 -t E -c DNA}\\
+    cd-hit-est-4.6.1 & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
+    seed-1.4.1 & \texttt{SEED --mismatch 3}\\
 \end{tabular}
@@
     \hline
-    starcode & \texttt{starcode -d3}\\ 
-    slidesort & \texttt{slidesort\_v2 -d 3 -u -t E -c DNA}\\
-    cd-hit-est & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
-    seed & \texttt{SEED --mismatch 3 --shift 3}\\ 
-    rainbow & \texttt{rainbow cluster -m 3}
+    starcode-1.0 & \texttt{starcode -d3}\\ 
+    slidesort-2 & \texttt{slidesort\_v2 -d 3 -u -t E -c DNA}\\
+    cd-hit-est-4.6.1 & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
+    seed-1.4.1 & \texttt{SEED --mismatch 3 --shift 3}\\ 
+    rainbow-2.0.3 & \texttt{rainbow cluster -m 3}
 \end{tabular}

### 2015-01-19 src/main-starcode.c
Added -r parameter to set the minimum parent-to-child cluster ratio.

@@ -38,6 +38,7 @@ char *USAGE =
 "    -d --dist: maximum Levenshtein distance (default auto)\n"
 "    -q --quiet: quiet output (default verbose)\n"
 "    -t --threads: number of concurrent threads (default 1)\n"
+"    -r --cluster-ratio: minumum cluster size ratio (default 5)\n"
 "    -s --sphere: sphere clustering (default message passing)\n"
 "    -v --version: display version and exit\n"
 "\n"
@@ -117,6 +118,7 @@ main(
    // Unset flags (value -1).
    int dist = -1;
    int threads = -1;
+   int cluster_ratio = -1;
 
    // Unset options (value 'UNSET').
    char * const UNSET = "unset";
@@ -135,6 +137,7 @@ main(
          {"sphere",            no_argument,       &sp_flag, 's'},
          {"version",           no_argument,              0, 'v'},
          {"dist",              required_argument,        0, 'd'},
+         {"cluster-ratio",     required_argument,        0, 'r'},
          {"help",              no_argument,              0, 'h'},
          {"input",             required_argument,        0, 'i'},
          {"input1",            required_argument,        0, '1'},
@@ -144,7 +147,7 @@ main(
          {0, 0, 0, 0}
       };
 
-      c = getopt_long(argc, argv, "1:2:d:hi:o:qst:v",
+      c = getopt_long(argc, argv, "1:2:d:hi:o:qst:r:v",
             long_options, &option_index);
  
       // Done parsing //
@@ -240,6 +243,22 @@ main(
          }
          else {
             fprintf(stderr, "error: --thread set more than once\n");
+            say_usage();
+            return EXIT_FAILURE;
+         }
+         break;
+
+      case 'r':
+         if (cluster_ratio < 0) {
+            cluster_ratio = atoi(optarg);
+            if (cluster_ratio < 1) {
+               fprintf(stderr, "error: --cluster-ratio must be numeric and greater than 0\n");
+               say_usage();
+               return EXIT_FAILURE;
+            }
+         }
+         else {
+            fprintf(stderr, "error: --cluster-ratio set more than once\n");
             say_usage();
             return EXIT_FAILURE;
          }
@@ -378,6 +397,7 @@ main(
 
    // Set remaining default options.
    if (threads < 0) threads = 1;
+   if (cluster_ratio < 0) cluster_ratio = 5;
 
    int exitcode =
    starcode(
@@ -389,6 +409,7 @@ main(
[…]

### 2015-01-19 src/starcode-private.h

@@ -38,9 +38,6 @@
 
 #define MAX_K_FOR_LOOKUP 14
 
-#define PARENT_TO_CHILD_FACTOR  5
-#define PARENT_TO_CHILD_MINIMUM 10
-
 #define BISECTION_START  1
 #define BISECTION_END    -1
 

### 2015-01-19 src/starcode.c

@@ -25,11 +25,11 @@
 #include "starcode-private.h"
 
 //    Global variables    //
-static FILE     * OUTPUTF1 = NULL;           // output file 1
-static FILE     * OUTPUTF2 = NULL;           // output file 2
-static format_t   FORMAT   = UNSET;          // input format
-static output_t   OUTPUTT  = DEFAULT_OUTPUT; // output type
-
+static FILE     * OUTPUTF1      = NULL;           // output file 1
+static FILE     * OUTPUTF2      = NULL;           // output file 2
+static format_t   FORMAT        = UNSET;          // input format
+static output_t   OUTPUTT       = DEFAULT_OUTPUT; // output type
+static int        CLUSTER_RATIO = 5;              // min parent-to-child ratio to link clusters.
 
 int
 starcode
@@ -42,6 +42,7 @@ starcode
    const int verbose,
    const int showclusters,
          int thrmax,
+         int parent_to_child,
    const int outputt
 )
 {
@@ -49,6 +50,7 @@ starcode
    OUTPUTF1 = outputf1;
    OUTPUTF2 = outputf2;
    OUTPUTT = outputt;
+   CLUSTER_RATIO = parent_to_child;
 
    if (verbose) {
       fprintf(stderr, "running starcode with %d thread%s\n",
@@ -464,7 +466,7 @@ do_query
                // pair if counts are on the same order of magnitude.
                int mincount = child->count;
                int maxcount = parent->count;
-               if (maxcount < PARENT_TO_CHILD_FACTOR * mincount) continue;
+               if (maxcount < CLUSTER_RATIO * mincount) continue;
                // The child is modified, use the child mutex.
                mutexid = match->count > query->count ?
                   job->queryid : job->trieid;

### 2015-01-19 src/starcode.h

@@ -46,6 +46,7 @@ int starcode(
    const int    verbose,
    const int    showclusters,
          int    maxthreads,
+         int    parent_to_child,
    const int    outputt
 );
 

### 2015-01-19 doc/document.tex
Corrected typos in manuscript and README.md.

@@
 exactness of slidesort was problematic because the number of
-3-matches in each dataset is not known (pairs of statellites
+3-matches in each dataset is not known (pairs of satellites
 in the same cluster may be 3-matches or not). For this reason
@@
     seed       & -           & 60,374      & -     \\
-    sidesort   & 4,055       & $>$ 10 days & -     \\
+    slidesort   & 4,055       & $>$ 10 days & -     \\
     rainbow    & -           & -           & 306   \\
@@
     seed       & -    & 53.9 & -   \\
-    sidesort   & 1.30 & 13.9 & -   \\
+    slidesort   & 1.30 & 13.9 & -   \\
     rainbow    & -    & -    & 0.5 \\

### 2015-01-19 misc/compute_Levenshtein_distances.py
Added example to prove that slidesort is not all-pairs search.

@@ -1,0 +1,46 @@
+#!/usr/bin/env python
+# -*- coding:utf-8 -*-
+
+import sys
+
+from itertools import combinations
+
+def levenshtein(s1, s2):
+    '''This implementation of the Levensthein distance is taken
+    from the following address (with the implicit assumption that
+    a public wiki version is less likely to have bugs than a private
+    version) http://en.wikibooks.org/wiki/Algorithm_Implementation/Strings/Levenshtein_distance#Python '''
+    if len(s1) < len(s2):
+        return levenshtein(s2, s1)
+ 
+    # len(s1) >= len(s2)
+    if len(s2) == 0:
+        return len(s1)
+ 
+    previous_row = range(len(s2) + 1)
+    for i, c1 in enumerate(s1):
+        current_row = [i + 1]
+        for j, c2 in enumerate(s2):
+            insertions = previous_row[j + 1] + 1 # j+1 instead of j since previous_row and current_row are one character longer
+            deletions = current_row[j] + 1       # than s2
+            substitutions = previous_row[j] + (c1 != c2)
+            current_row.append(min(insertions, deletions, substitutions))
+        previous_row = current_row
+ 
+    return previous_row[-1]
+
+if __name__ == '__main__':
+   # Read the fasta file (ignore headers).
+   sequences = []
+   with open('tiny.fasta') as f:
+      for line in f:
+         if line[0] == '>': continue
+         else: sequences.append(line.rstrip())
+
+   # Compare pairwise Levenshtein distances.
+   count = 0
+   for a,b in combinations(sequences,2):
+      d = levenshtein(a,b)
+      if d < 4:
+         count += 1
+         sys.stdout.write('%d. %s %s\n' % (count, a,b))

### 2015-01-19 src/main-starcode.c

@@ -36,10 +36,10 @@ char *USAGE =
 "\n"
 "  general options:\n"
 "    -d --dist: maximum Levenshtein distance (default auto)\n"
-"    -q --quiet: quiet output (default verbose)\n"
 "    -t --threads: number of concurrent threads (default 1)\n"
 "    -r --cluster-ratio: minumum cluster size ratio (default 5)\n"
 "    -s --sphere: sphere clustering (default message passing)\n"
+"    -q --quiet: quiet output (default verbose)\n"
 "    -v --version: display version and exit\n"
 "\n"
 "  input/output options (single file, default)\n"

### 2015-01-20 doc/document.tex
Manuscript updated.

@@
 sequencing error. This method uses message passing \citep{mackay}
-to identfy and count ``canonical'' sequences (also referred to as
-centroids in the clustering terminology). Each sequence transfers its
+to identify and count ``canonical'' sequences (also referred to as
+centroids in the clustering terminology). By default, each sequence transfers its
 read count to its closest $\tau$-match provided the latter has at least 5
@@
 Otherwise, sequences are more likely unrelated, or both are derived
-from the same canonical sequence.
+from the same canonical sequence. This behavior can be modified with
+the command-line option \emph{cluster-ratio} to allow for a more
+flexible or more strict clustering, e.g. to cluster unique input
+sequences together, \emph{cluster-ratio} must be set to 1.
 
@@
 \subsection{Presentation and basic performance}
-\label{performance}
+\label{section:performance}
 Starcode is a general purpose DNA sequence clustering tool with a
@@
 on 1 thread and with a maximum Levenshtein distance of 3. In each
-test, only one parameter is modified while the others are kept constant.
+test, only one parameter is modified while the others are kept
+constant. Since the clustering step does not require additional memory
+allocation and is significantly faster than all-pairs search, the
+performance results presented in Sections \ref{section:performance} and
+\ref{section:benchmark} apply for both message-passing and spheres
+clustering algorithms.
 
@@
 parallel and the relative performance increases linearly up to 12 threads.
-The bending observed thereafter has two sources. The first is 
+The bending observed thereafter has two sources; the first is 
 that the input reading and clustering steps are brief but not parallel,
@@
 on smaller datasets where naive pairwise comparisons are feasible
-(more information is available on the starcode repository the, see
-http://github.com/gui11aume/starcode/misc).
+(more information is available on the starcode repository, see
+http://github.com/gui11aume/starcode/tree/master/misc).
 

### 2015-01-25 doc/rebuttal_1.txt
Updated rebuttal and submitted (final) manuscript.

We would like to thank the reviewers for their constructive comments, which we found very useful to improve the software and the manuscript. We believe that the new version of the manuscript is more clear and focused. More importantly, the suggestions of the reviewers have helped us highlight the advantages of starcode over competing software.

Given the substantial amount of modifications requested by reviewers #2 and #3, we have completely rewritten the manuscript, except for section 3.3. It would be too cumbersome to list all the changes, but we quickly summarize the most conspicuous.
1. We have changed the title and no longer refer to starcode as an exact clustering algorithm.
2. We have removed the comparison with BWT-based short read aligners to replace it by a comparison with other sequence clustering algorithms.
3. We have added a figure in the methods section to clarify the lossless filtration procedure.

Below is a point-by-point response to the comments of the reviewers.

----------------------------------------
Reviewer: 1

Comments to the Author
I have read your manuscript about the algorithm you have written, and testing the available implementation. As far as I have seen with a small subset of a real dataset of 632624 42-mer SNV reads, when both starcode and cd-hit-est are fine tuned , both spent the same time for it. I realized that your algorithm takes much time just changing the Levenshtein distance from 3 to 4. So, would it be feasible to autotune the Levenshtein distance based on the average sequence size?

---------
RESPONSE:
We have also observed a very significant time difference between 3 and 4 errors. The main reason we did not implement this in the previous version is that the error rate depends on the experimental setup, which is known only to the end user. However, we have now decided to set the default parameters of starcode to correct errors introduced during sequencing, which allows us to guess a value for the maximum Levenshtein distance. The default value is now 2 + floor(median length/30), which is compatible with the error rate expected from Illumina sequencers. Experienced users can of course overrule this by the value that is more appropriate to their experimental setup.
---------

I have another question. Have you thought about using an adapted version of your clustering algorithm on the context of bisulphite sequencing data? In that context, you usually need a reference sequence to compare the methylation state of the cytosines in the reads. In that context, a methylated cytosine becomes a thymine and its complementary nucleotide, guanine, becomes an adenine. And reading your strategy I guess it would also work for this kind of data taking into account this "polymorphism".

---------
RESPONSE:
We would like to thank you for this suggestion, which we did not think about. Your comment makes us aware of the need to develop efficient algorithms for this type of data. However, we believe that adding this feature would stretch starcode too far from its core design. Libraries of shotgun bisulfite sequencing can be clustered by starcode, but the reads that do not have the same start and end position will most likely not cluster together because their Levenshtein distance will be high. As a result, we do not know what use this would be for the end user. It may be that we have misunderstood the application you have in mind. Should this be the case, please do not hesitate to provide an example dataset and to give a concrete proposal of how you would modify starcode in this context, we would then be happy to implement this additional feature, provided it is feasible.
---------

About the implementation, I have a suggestion. FASTQ format is much used on NGS pipelines, so your implementation could read such files.

---------
RESPONSE:
The FASTQ format is indeed the de facto standard. Starcode can now process single and paired-end FASTQ file formats.
---------

Reviewer: 2

Comments to the Author
This paper describes Starcode, a pipeline for discovering pairs of short sequences with small edit distance, followed by clustering. The basic approach is seed matching, followed by an optimized trie-based Needleman-Wunsch algorithm to discover similarities, followed by a heuristic clustering procedure.  A parallel version of the algorithm is benchmarked on one synthetic and one real data set against other DNA read clustering and short-read alignment tools, and then applied to barcode identification and motif finding tasks.

The methods in Starcode are generally strong.  While optimizing Smith-Waterman-like algorithms using a trie is not itself highly novel, the detailed strategy described has some attractive aspects (banding, sorting reads, and careful attention to avoiding recomputation). The initial seed-matching stage is guaranteed not to miss alignments of interest, but the detailed description in Section 2.3 is somewhat confusing, and in any case the method used is only the simplest of a large number of lossless filtration strategies.

---------
RESPONSE:
With your pointers about lossless filtration strategies, we have been able to trace the first mention of the segmentation approach to a classical paper by Wu and Manber, which is now cited in the text. We have rewritten the section explaining the filtration step and we have included a figure that, we hope, clarifies the process and makes this section more readable. We have looked into more elaborate filtration strategies, such as q-gram-based methods, but none of them seemed particularly adapted for trie indexes. Given that the simple segmentation approach of the current implementation seems to give satisfactory results, we decided to keep it. However, we are willing to improve starcode if possible, and we would be happy to implement any specific suggestion for a more powerful filtration method.
---------

The method is "exact" in the sense that every pairwise alignment of interest will be found, but the following clustering step is still necessarily heuristic, so it may not be fair to call the overall pipeline an exact algorithm.  The fact that the pipeline gets the clearly right answer on the synthetic data set is more attributable to the simplicity of the test case than any inherent merit of the clustering algorithm.

---------
RESPONSE:
Somewhat too focused on the pairing step, we did not realize that clustering algorithms cannot be exact. Thanks for the reality check. We have changed the title to "Starcode: sequence clustering based on all-pairs search", which we believe reflects better what starcode does. Also, we no longer refer to starcode as an exact algorithm, and we have taken care to use the 'exact' attribute only for the pairing step in the text.
---------

Overall, validation is somewhat limited.  The tests versus other tools confound the contributions of the fundamental algorithm and the parallelization strategy to overall performance; hence, it is difficult to assess whether one algorithm outperforms another on a single core, or whether the only performance differences are due to scalability of the tools' parallelization strategies.  Moreover, the competing algorithms were not tuned to solve the task at hand, but only to produce the best answer they could within a certain arbitrary time limit.

---------
RESPONSE:
We have changed the benchmark along those lines. All the tools are now benchmarked with a single thread. The only exception is in Figure 5d, where starcode is used with an increasing number of threads on the same dataset to show the performance scaling. We also present more fair and thorough tests, the competing tools were allowed to run until completion with the notable exception of slidesort. Slidesort is several orders of magnitudes slower than the other tools and that it would most likely not return before several weeks when run on the largest benchmark datasets. We have contacted the editor about this issue and after his response we have decided to interrupt the runs after 10 days. Given that starcode runs in less than 1 hour on the most difficult dataset, we hope that you will agree that this threshold is not unfair (even if we admit that it is arbitrary).
---------

More significantly, the competing tools are solving a different problem than Starcode, and the parametrizations of those tools were questionable, so the comparison of error rates, and possibly of running times, in Figure 5 is frankly dubious.  For example, BWA was told not to penalize mismatches at all (-B0) and to effectively disable its seeding stage (-k1), used a different scoring system for gaps, and in any case is not designed to be a complete algorithm.   Bowtie2 also uses a different scoring system, and its documentation states that the tools is not designed for -a mode, and that using this flag will make it run very slowly.   I note also that the short read mappers were only tested on the synthetic data set, while the actual short read data set was tested only with the generic clustering tools (CD-HIT and USEARCH).    None of the other tools were tested on the TRIP barcode recovery task.

---------
RESPONSE:
Our initial purpose was to test starcode against very fast algorithms to emphasize its speed. We realize that this is off the point so we have removed the comparison with BWT-based aligners. We have tried to test all the algorithms on all the datasets (artificial and experimental, including TRIP barcode recovery), which was not always possible because of the distinction between single and paired-end reads.
---------

Overall, the Starcode tool is likely a useful contribution in itself, particularly for tasks such as barcode recovery where the completeness of the pairwise alignment stage is important. However, the validation should be fairer to competing tools and should separate the contributions of the core algorithm versus parallelism.

MINOR COMMENTS:

* In the abstract, "computationally complexity" is a typo.

* In section 2.3, the variable "k" is used in at least two places where it seems that "tau" was meant.

---------
RESPONSE:
We took this information into account when updating the text.
---------

Reviewer: 3

Comments to the Author
GENERAL COMMENTS
This paper presents an exact algorithm for clustering nearly identical short read sequences. It addresses an important need in the NGS field. The algorithm is novel and interesting, and its performance results look promising. Thus, it is likely to be of interest to many users in the rapidly growing NGS field. However, there are several major concerns with this paper that need to be addressed.

MAJOR CONCERNS
(1) The authors compare their tool only against sequence clustering methods which have not been specifically designed for short read clustering, while ignoring the most relevant algorithms and software implementations designed for this application field, including SlideSort (Bioinformatics 27, 464–470), SEED (Bioinformatics 27, 2502–2509), Rainbow (Bioinformatics 28, 2732–2737) and many more. SlideSort is also an exact algorithm and it is not even mentioned by this publication. Instead of including the not very meaningful results from the BWT-based aligners, the authors need to make comparisons against tools that are intended for this application domain.

---------
RESPONSE:
We would like to thank you for pointing out the existence of these tools, which we were not aware of. We have included them in the benchmark. In the mean time, we have removed USEARCH. It turned out early on in the revision process that the free version of USEARCH imposes a limit of 4GB on memory usage. With this limitation, it was impossible to benchmark it on the experimental datasets. As mentioned in the answers to reviewer #2, the comparison with the BWT-based aligners has been removed as we have realized that it is not very meaningful.
In the course of the benchmark, it occurred to us that slidesort is not an exact algorithm, contrary to the claim of the original article. We have contacted the authors of slidesort about this issue. Their answer was that it could be a bug and that they would get back in touch with us in a few weeks. Since we are unsure about the way to deal with this issue, we provide in the revised manuscript a link to a dataset for which slidesort fails to identify all the pairs. Please let us know whether this is appropriate, or whether you recommend other actions.
---------

(2) While Starcodes shows some very promising performance, it would be important to add a section that summarizes for the reader what type of sequence clustering problems it addresses and which ones not, e.g. mismatches, gaps, upper/lower length limit of sequences, sequences of variable length (e.g. after adaptor/quality trimming), etc.

---------
RESPONSE:
The results section of the revised manuscript starts with a general description of starcode, along with a description of the input and output. We present starcode as a clustering algorithm for error correction because it was originally designed for this purpose, and also because we believe that this gives a concrete sense of what starcode does.
---------

(3) The current set of performance tests with simulated data misses to include clusters of extreme sizes, e.g. 100 million sequence where some clusters contain ~0.5 million sequences, which can greatly degrade the time performance of certain clustering algorithms. Those extreme cluster sizes are very common in real NGS data sets due to PCR artifacts,  adaptor contaminations, etc.

---------
RESPONSE:
We now test the sequence clustering tools on a hard experimental dataset (dataset 2 in the benchmark section), which contains slightly more than 127 million sequences, with 4 clusters over 1 million sequences.
---------

(4) Fig 4c shows the performance for a very narrow window of read lengths. It would be very interesting to include here reads of lengths of up to 300bp or longer because many NGS technologies produce now reads of this length.

---------
RESPONSE:
The analysis has been run up to 420 bp (the result is now shown in Fig 5c).
---------

(5) The topic of clustering paired-end (PE) read data should also be addressed or at least discussed somewhere.

---------
RESPONSE:
While implementing support for FASTQ files, we have also added the possibility to process paired-end files. The presentation section makes this explicit. This is also clear from the benchmark section.
---------

(6) Memory performance comparisons against other tools need to be included in the paper as it can be one of the biggest bottlenecks, even more than time performance.

---------
RESPONSE:
The benchmark section now includes the peak memory usage of all the tools.
---------

(7) The time performance comparisons against CD-HIT and USEARCH are incomplete. As described on page 4, the authors did not run the tests with these tools to completion (tests were sometimes stopped after an arbitrary time cutoff) which is unusual and misleading as it implies they could not be used for this task. Moreover, including in these comparisons NGS data sets with different properties and presenting the results in a table or chart would be more informative.

---------
RESPONSE:
The benchmark section is now more thorough and we ran the software until completion, to the exception of slidesort, as explained in the answers to the comments of reviewer #2. We now represent the data in a graphical way (Figure 6) and in tabular format (Tables 4 and 5).
---------

(8) The user manual page on github lists the input and output formats of Starcode. Support for FASTQ files is missing, which is by far the most common sequence format in the NGS field. It is also not clear to me how users are expected to process the outputs generated by Starcode. A typical routine would be the removal of the redundant sequences from the original FASTQ file. Adding support for this and potentially other outputs would greatly improve the usability of the software for non-expert users.

---------
RESPONSE:
As mentioned above, we have included support for FASTQ input files. The option --non-redundant now allows the user to remove redundant sequences from the input file, while maintaining the format (raw sequence, FASTA or FASTQ). In many applications, the counts are equally important as the sequences (for instance in transcriptomics and its derivatives). We are not aware of a popular file format specifying a sequence and a count, which is why we opted for the current default output format. If there is such a standard, we will be happy to change the default output of starcode (and of course we will implement other formats if the users of starcode request it).
---------

### 2015-01-25 doc/rebuttal_2.txt
Updated rebuttal and submitted (final) manuscript.

Dear editor, dear reviewers,

thank you once again for your constructive feedback. We have updated the manuscript according to your comments. Below is a brief clarification for each of the addressed issues.

For clarity, our responses to the points of the reviewers are bracketed by lines of dashes.

Reviewer: 1

Comments to the Author
The authors have addressed all my concerns from the initial review. The paper has greatly improved by including now comparisons against competing software tools, while the result of the benchmark comparisons are very strong.

Reviewer: 2

Comments to the Author
Meanwhile I was revising the manuscript, I realized that although you described the benchmarking scenario (hardware, run modes, flags, used datasets, etc...) quite well, you did not include the version numbers of the other algorithms/implementations you compared against. Bioinformatic software sometimes evolves too fast, and manuscript sentences which are currently true (or when the benchmarks were made) could not be reproduced in the future due software improvements, flag and behavior changes, bugfixes, etc... so version information is also a requirement.

---------
Thanks for the observation, we have included the versions of the benchmarked software in tables 1 and 2.
---------

Also, as starcode has in its implementation more than one clustering algorithm, I recommend you to include runs with this alternate algorithm with the adequate (if not all the) benchmarks.

---------
The most complex step of starcode is discovering whether each sequence pair falls within a given edit distance, which is orders of magnitude more expensive in terms of time and memory usage than any of the clustering algorithms. The performance benchmarks are therefore independent of the selected clustering strategy.

To point this out in the manuscript, we have added the following sentence in section 3.1:

"Since the clustering step does not require additional memory allocation and is significantly faster than all-pairs search, the performance results presented in Sections 3.1 and 3.2 apply for both message-passing and spheres clustering algorithms."
---------

In page 4, when you are describing the default clustering algorithm you explain its behavior, establishing that each sequence transfers its read count to its closest tau-match, provided the latter has at least 5 times more counts. And the reason it is 5 times is because no sequencing technology has an error rate higher than 20%, which is a very reasonable default. Perhaps it is a nonsense, but I think it would be useful to provide a parameter to adjust this value on runtime, so you can set up stricter or laxer error rates.

---------
This is indeed a good point. For instance, Oxford Nanopore seems to have an error rate higher than 20%. We have included a new command-line parameter (cluster-ratio) to set the minimum count ratio to cluster in message-passing mode. Additionally, we have included the following note in the manuscript (Section 2.6):

"This behavior can be modified with the command-line option cluster-ratio to allow for a more flexible or more strict clustering, e.g. to cluster unique input sequences together, cluster-ratio must be set to 1."
---------

Page 5, left column, line 56: the link http://github.com/gui11aume/starcode/misc to the supplementary material does not work

---------
We apologize for the issue, the link was incorrect and we accidentally deleted the folder at some point. The misc folder has been uploaded to the repository again and the link is now accessible. Also, the text has been updated to pont to the the proper page on Github, which is https://github.com/gui11aume/starcode/tree/master/misc.
---------

Page 5, left column, line 44: a typo, "statellites" should be "satellites"
Page 6, tables 4 and 5: a typo, "sidesort" should be "slidesort"

---------
The typos have been corrected in the updated version of the manuscript.
---------

### 2015-06-15 doc/starcode_tutorial.tex
First version of the tutorial.

\documentclass[12pt]{article}
\usepackage{url}
\title{Starcode tutorial}
\author{Guillaume Filion}

\begin{document}
\maketitle

\section{Build instructions}

Starcode runs on Mac and Linux as a command line application.
You can download the source code of Starcode from Github with
the following command on a standard terminal.

\begin{verbatim}
git clone git@github.com:gui11aume/starcode
\end{verbatim}

Note that this requires that you already have a Github account
and that the computer you are working on has an SSH  key registered
on Github. If this is not the case, follow the instructions from
\url{https://help.github.com/articles/generating-ssh-keys/}.

This should download a directory named \texttt{starcode}. To build
Starcode, execute the following.

\begin{verbatim}
cd starcode
make
\end{verbatim}

This should succeed on most Linux systems because \texttt{make} is
available by default. Should it not be the case, you can obtain it by
typing \texttt{sudo apt-get install make} on Ubuntu. On Mac, you need
to install XCode, which may take some time. First, you will need an
Apple ID, then you will need to download it from the developer
website of Apple \url{https://developer.apple.com/xcode/downloads/}.
Then, you may need to follow the instructions shown on the following
link to install the command line version of \texttt{make}
\url{http://stackoverflow.com/q/10265742/1248687}.

This should have created an executable called \texttt{starcode}. To
check that the building is successful, execute the following command
in the same directory.

\begin{verbatim}
./starcode test/test_file.txt
\end{verbatim}

You should obtain the following output.

\begin{verbatim}
running starcode with 1 thread
reading input files
raw format detected
sorting
setting dist to 2
progress: 100.00%
AGGGCTTACAAGTATAGGCC  7
CCTCATTATTTGTCGCAATG  7
GGGAGCCCACAGTAAGCGAA  7
TAGCCTGGTGCGACTGTCAT  7
TGCGCCAAGTACGATTTCCG  7
\end{verbatim}

\section{Starcode basics}

The previous example is a standard invocation of Starcode on sequence
data. The input file \texttt{test/test\_file.txt} is in raw format
(\textit{i.e.} plain text), it contains 35 lines, each with one
sequence of 20 nucleotides. By default, Starcode is verbose and prints
some information about the clustering process. In the example above,
Starcode used a single CPU, and it clustered the sequences within
edit distance 2 of each other.

The output starts after the line \texttt{progress: 100.00\%}. It
consists of five lines (one per cluster) where the sequence is the
centroid of the cluster (the most representative sequence) and the
number is its size (the number of sequences in the cluster).

We can make Starcode quiet (non verbose) with the \texttt{-q} option,
and we can set the clustering distance to 0 with \texttt{-d0}.

\begin{verbatim}
./starcode -q -d0 test/test_file.txt
\end{verbatim}

The output is the following.

\begin{verbatim}
AGGGCTTACAAGTATAGGCC  6
CCTCATTATTTGTCGCAATG  6
GGGAGCCCACAGTAAGCGAA  6
TAGCCTGGTGCGACTGTCAT  6
TGCGCCAAGTACGATTTCCG  6
AGGGGTTACAAGTCTAGGCC  1
CCTCATTATTTACCGCAATG  1
GGAAGCCCACAGCAAGCGAA  1
TAACCTGGTGCGACTGTTAT  1
TGCGCCAAGTAAGAATTCCG  1
\end{verbatim}

This time we obtain five clusters of size 6 and five clusters of size 1.
Each centroid of the clusters of size 1 has 2 mismatches with one
of the centroids of the clusters of size 6, which is why they form
separate clusters when the distance is less than 2.

As a side note, this example output illustrates that the clusters
are sorted first by size and then by alphabetical order of the
centroid.

\end{document}

### 2015-06-17 doc/tutorial.tex
Put the pdf of the tutorial in the root directory.

--- 2015-06-15 doc/starcode_tutorial.tex
@@
 This should succeed on most Linux systems because \texttt{make} is
-available by default. Should it not be the case, you can obtain it by
+available by default. If this is not the case, you can obtain it by
 typing \texttt{sudo apt-get install make} on Ubuntu. On Mac, you need
@@
 
-This should have created an executable called \texttt{starcode}. To
-check that the building is successful, execute the following command
-in the same directory.
+Calling \texttt{make} should create an executable called
+\texttt{starcode}. To check that the building is successful, execute
+the following command in the same directory.
 
@@
 
-You should obtain the following output.
+If you obtain the output shown below, then everything went fine
+and you are done with the build. If not, then something went
+wrong. In this case, you can explain how to reproduce the problem
+on \url{https://github.com/gui11aume/starcode/issues}.
 
@@
 The previous example is a standard invocation of Starcode on sequence
-data. The input file \texttt{test/test\_file.txt} is in raw format
+data. The input file \texttt{test/test\char`_file.txt} is in raw format
 (\textit{i.e.} plain text), it contains 35 lines, each with one

### 2015-11-17 src/starcode-private.h
Sorted output sequence ids and removed duplicates.

@@ -164,6 +164,7 @@ int        count_order (const void *, const void *);
 void       destroy_useq (useq_t *);
 void       destroy_lookup (lookup_t *);
 void     * do_query (void*);
+int        int_ascending (const void*, const void*);
 void       krash (void) __attribute__ ((__noreturn__));
 int        lut_insert (lookup_t *, useq_t *); 
 int        lut_search (lookup_t *, useq_t *); 

### 2015-11-17 src/starcode.c

@@ -192,12 +192,19 @@ starcode
             if (u->canonical != canonical) {
                // Print previous cluster seqIDs.
                if (showids) {
-                  if (canonical->nids > 1)
-                     fprintf(OUTPUTF1, "\t%u", canonical->seqid[0]);
-                  else
+                  int last = -1;
+                  if (canonical->nids > 1) {
+                     qsort(canonical->seqid, canonical->nids, sizeof(int), int_ascending);
+                     last = canonical->seqid[0];
+                     fprintf(OUTPUTF1, "\t%u", last);
+                  }
+                  else 
                      fprintf(OUTPUTF1, "\t%u", (unsigned int)(unsigned long)canonical->seqid);
-                  for (unsigned int k = 1; k < canonical->nids; k++)
-                     fprintf(OUTPUTF1, ",%u", canonical->seqid[k]);
+                  for (unsigned int k = 1; k < canonical->nids; k++) {
+                     if (canonical->seqid[k] == last) continue;
+                     last = canonical->seqid[k];
+                     fprintf(OUTPUTF1, ",%u", last);
+                  }
                }
                canonical = u->canonical;
                if (FORMAT == PE_FASTQ) {
@@ -233,12 +240,18 @@ starcode
 
          // Print last cluster seqIDs.
          if (showids) {
-            if (canonical->nids > 1)
-               fprintf(OUTPUTF1, "\t%u", canonical->seqid[0]);
-            else
+            int last = -1;
+            if (canonical->nids > 1) {
+               qsort(canonical->seqid, canonical->nids, sizeof(int), int_ascending);
+               last = canonical->seqid[0];
+               fprintf(OUTPUTF1, "\t%u", last);
+            } else
                fprintf(OUTPUTF1, "\t%u", (unsigned int)(unsigned long)canonical->seqid);
-            for (unsigned int k = 1; k < canonical->nids; k++)
-               fprintf(OUTPUTF1, ",%u", canonical->seqid[k]);
+            for (unsigned int k = 1; k < canonical->nids; k++) {
+               if (canonical->seqid[k] == last) continue;
+               last = canonical->seqid[k];
+               fprintf(OUTPUTF1, ",%u", last);
+            }
          }
          fprintf(OUTPUTF1, "\n");
 
@@ -285,12 +298,14 @@ starcode
 
[…]

### 2015-11-17 doc/tutorial.tex

@@
 
-The previous example is a standard invocation of Starcode on sequence
-data. The input file \texttt{test/test\char`_file.txt} is in raw format
-(\textit{i.e.} plain text), it contains 35 lines, each with one
-sequence of 20 nucleotides. By default, Starcode is verbose and prints
-some information about the clustering process. In the example above,
-Starcode used a single CPU, and it clustered the sequences within
-edit distance 2 of each other.
+Starcode is a sequence clustering algorithm. It takes a bunch of
+sequences, finds which are similar and pools them together. Starcode
+considers that two sequences are similar if their edit distance
+(also known as the Levenshtein distance) is low, \textit{i.e.}
+if they almost identical, except for a few mismatches, insertions
+and deletions.
 
-The output starts after the line \texttt{progress: 100.00\%}. It
-consists of five lines (one per cluster) where the sequence is the
-centroid of the cluster (the most representative sequence) and the
-number is its size (the number of sequences in the cluster).
+The previous example is a standard run of Starcode on sequence
+data. By default, Starcode is verbose and gives some information on
+how the run is performed before it outputs the clusters. 
+In this case, we learn that Starcode used a single CPU, that
+the input file \texttt{test/test\char`_file.txt} is in raw format
+(\textit{i.e.} plain text), and that the threshold distance for
+similarity search was set to 2. The output propoer starts after
+the line \texttt{progress: 100.00\%}. It consists of five lines
+(one per cluster) where the sequence is the centroid of the cluster
+(the irepresentative sequence) and the number is its size (the
+number of sequences in the cluster).
+
+The input file \texttt{test/test\char`_file.txt} contains 35 sequences
+that Starcode arranged in 5 clusters of 7 sequences each.
 

### 2015-11-17 src/main-starcode.c
Terminal call without arguments/pipe prints usage.

@@ -135,6 +135,11 @@ main(
    char * input2  = UNSET;
    char * output  = UNSET;
 
+   if (argc == 1 && isatty(0)) {
+      say_usage();
+      return EXIT_SUCCESS;
+   }
+
    int c;
    while (1) {
       int option_index = 0;

### 2015-11-17 src/main-starcode.c
Reorganized tests. Minor change to help message.

@@ -59,10 +59,11 @@ char *USAGE =
 "  output format options\n"
 "       --print-clusters: outputs cluster compositions\n"
 "       --non-redundant: remove redundant sequences\n"
-"       --seq-id: print sequence numbers\n";
+"       --seq-id: also print sequence id numbers (1-based)\n";
 
 void say_usage(void) { fprintf(stderr, "%s\n", USAGE); }
 void say_version(void) { fprintf(stderr, VERSION "\n"); }
+
 void SIGSEGV_handler(int sig) {
    void *array[10];
    size_t size;

### 2015-11-17 src/trie-private.h

@@ -1,106 +1,0 @@
-/*
-** Copyright 2014 Guillaume Filion, Eduard Valera Zorita and Pol Cusco.
-**
-** File authors:
-**  Guillaume Filion     (guillaume.filion@gmail.com)
-**  Eduard Valera Zorita (ezorita@mit.edu)
-**
-** License: 
-**  This program is free software: you can redistribute it and/or modify
-**  it under the terms of the GNU General Public License as published by
-**  the Free Software Foundation, either version 3 of the License, or
-**  (at your option) any later version.
-**
-**  This program is distributed in the hope that it will be useful,
-**  but WITHOUT ANY WARRANTY; without even the implied warranty of
-**  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-**  GNU General Public License for more details.
-**
-**  You should have received a copy of the GNU General Public License
-**  along with this program.  If not, see <http://www.gnu.org/licenses/>.
-**
-*/
-
-#ifndef _STARCODE_TRIE_PRIVATE_HEADER
-#define _STARCODE_TRIE_PRIVATE_HEADER
-
-#define PAD 5              // Position of padding nodes.
-#define EOS -1             // End Of String, for 'dash()'.
-// Translation tables between letters and numbers.
-static const char untranslate[7] = "NACGT N";
-// Translation table to insert nodes in the trie.
-//          ' ': PAD (5)
-//     'a', 'A': 1
-//     'c', 'C': 2
-//     'g', 'G': 3
-//     't', 'T': 4
-static const int translate[256] = { 
-   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
-   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
- PAD,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
-   0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,
-   0,1,0,2,0,0,0,3,0,0,0,0,0,0,0,0,
[…]

### 2015-11-17 test/tests_starcode.c

@@ -1,0 +1,993 @@
+#include "unittest.h"
+#include "starcode.c"
+
+static const char untranslate[7] = "NACGT N";
+
+void
+test_starcode_1
+// Basic tests of useq.
+(void)
+{
+   // Regular initialization without info.
+   useq_t *u = new_useq(1, "some sequence", NULL);
+   test_assert_critical(u != NULL);
+   test_assert(u->count == 1);
+   test_assert(strcmp(u->seq, "some sequence") == 0);
+   test_assert(u->info == NULL);
+   test_assert(u->matches == NULL);
+   test_assert(u->canonical == NULL);
+   destroy_useq(u);
+
+   // Regular initialization with info.
+   u = new_useq(1, "some sequence", "some info");
+   test_assert_critical(u != NULL);
+   test_assert(u->count == 1);
+   test_assert(strcmp(u->seq, "some sequence") == 0);
+   test_assert(strcmp(u->info, "some info") == 0);
+   test_assert(u->matches == NULL);
+   test_assert(u->canonical == NULL);
+   destroy_useq(u);
+
+   // Initialize with negative value and \0 string.
+   u = new_useq(-1, "", NULL);
+   test_assert_critical(u != NULL);
+   test_assert(u->count == -1);
+   test_assert(strcmp(u->seq, "") == 0);
+   test_assert(u->info == NULL);
+   test_assert(u->matches == NULL);
+   test_assert(u->canonical == NULL);
+   destroy_useq(u);
+
+   // Initialize with NULL string.
+   u = new_useq(0, NULL, NULL);
+   test_assert(u == NULL);
+
+}
+
+
+void
+test_starcode_2
+(void)
+// Test addmatch().
+{
+
+   useq_t *u1 = new_useq(12983, "string 1", NULL);
+   test_assert_critical(u1 != NULL);
+   test_assert(u1->matches == NULL);
+   useq_t *u2 = new_useq(-20838, "string 2", NULL);
+   test_assert_critical(u2 != NULL);
+   test_assert(u2->matches == NULL);
+
+   // Add match to u1.
+   addmatch(u1, u2, 1, 2);
+
+   test_assert(u2->matches == NULL);
[…]

### 2015-11-17 test/tests_trie.c

@@ -1,8 +1,5 @@
-#include "trie.h"
-#include "trie-private.h"
-#include "starcode.h"
-#include "starcode-private.h"
 #include "unittest.h"
+#include "trie.c"
 
 // Globals //
 int LEAF_NODE;
@@ -1658,1037 +1655,5 @@ test_mem_6
 
    unredirect_stderr();
    reset_alloc();
-
-}
-
-void
-test_starcode_1
-// Basic tests of useq.
-(void)
-{
-   // Regular initialization without info.
-   useq_t *u = new_useq(1, "some sequence", NULL);
-   test_assert_critical(u != NULL);
-   test_assert(u->count == 1);
-   test_assert(strcmp(u->seq, "some sequence") == 0);
-   test_assert(u->info == NULL);
-   test_assert(u->matches == NULL);
-   test_assert(u->canonical == NULL);
-   destroy_useq(u);
-
-   // Regular initialization with info.
-   u = new_useq(1, "some sequence", "some info");
-   test_assert_critical(u != NULL);
-   test_assert(u->count == 1);
-   test_assert(strcmp(u->seq, "some sequence") == 0);
-   test_assert(strcmp(u->info, "some info") == 0);
-   test_assert(u->matches == NULL);
-   test_assert(u->canonical == NULL);
-   destroy_useq(u);
-
-   // Initialize with negative value and \0 string.
-   u = new_useq(-1, "", NULL);
-   test_assert_critical(u != NULL);
-   test_assert(u->count == -1);
-   test_assert(strcmp(u->seq, "") == 0);
-   test_assert(u->info == NULL);
-   test_assert(u->matches == NULL);
-   test_assert(u->canonical == NULL);
-   destroy_useq(u);
-
-   // Initialize with NULL string.
-   u = new_useq(0, NULL, NULL);
-   test_assert(u == NULL);
-
-}
-
-
-void
-test_starcode_2
-(void)
-// Test addmatch().
-{
-
-   useq_t *u1 = new_useq(12983, "string 1", NULL);
-   test_assert_critical(u1 != NULL);
-   test_assert(u1->matches == NULL);
-   useq_t *u2 = new_useq(-20838, "string 2", NULL);
[…]

### 2015-11-17 doc/tutorial.tex
Updated the tutorial.

@@
 \begin{verbatim}
-./starcode test/test_file.txt
+./starcode
 \end{verbatim}
@@
 on \url{https://github.com/gui11aume/starcode/issues}.
+
+\begin{verbatim}
+Usage:  starcode [options]
+
+  general options:
+    -d --dist: maximum Levenshtein distance (default auto)
+    -t --threads: number of concurrent threads (default 1)
+    -r --cluster-ratio: minumum cluster size ratio (default 5)
+    -s --sphere: sphere clustering (default message passing)
+    -q --quiet: quiet output (default verbose)
+    -v --version: display version and exit
+
+  input/output options (single file, default)
+    -i --input: input file (default stdin)
+    -o --output: output file (default stdout)
+
+  input options (paired-end fastq files)
+    -1 --input1: input file 1
+    -2 --input2: input file 2
+
+  output format options
+       --print-clusters: outputs cluster compositions
+       --non-redundant: remove redundant sequences
+       --seq-id: also print sequence id numbers (1-based)
+\end{verbatim}
+
+\section{Starcode basics}
+
+Starcode is a sequence clustering algorithm. It takes a bunch of
+sequences, finds which are similar and pools them together. Starcode
+considers that two sequences are similar if their edit distance
+(also known as the Levenshtein distance) is low, \textit{i.e.}
+if they almost identical, except for a few mismatches, insertions
+and deletions.
+
+Let us get started with a simple example. At the command line,
+issue the command below.
+
+\begin{verbatim}
+./starcode test/test_file.txt
+\end{verbatim}
+
+In principle, you should obtain the following output.
 
@@
 
-\section{Starcode basics}
-
-Starcode is a sequence clustering algorithm. It takes a bunch of
-sequences, finds which are similar and pools them together. Starcode
-considers that two sequences are similar if their edit distance
-(also known as the Levenshtein distance) is low, \textit{i.e.}
-if they almost identical, except for a few mismatches, insertions
-and deletions.
-
-The previous example is a standard run of Starcode on sequence
-data. By default, Starcode is verbose and gives some information on
+By default, Starcode is verbose and gives some information on
 how the run is performed before it outputs the clusters. 
@@
 
+\section{Clustering options}
+
+Coming soon...
+
+\section{Documenting clusters}
+
+Starcode also allows you to track the sequences that end up in a
+given cluster. Here, `sequences' can be considered as either input
+sequences, or distinct nucleotide sequences. The option
+\texttt{--print-clusters} shows the distinct nucleotide sequences
+that constitute the cluster. Below is and example.
+
+\begin{verbatim}
+./starcode -q --print-clusters test/test_file.txt
+AGGGCTTACAAGTATAGGCC  7 AGGGCTTACAAGTATAGGCC,AGGGGTTACAAGTCTAGGCC
+CCTCATTATTTGTCGCAATG  7 CCTCATTATTTACCGCAATG,CCTCATTATTTGTCGCAATG
+GGGAGCCCACAGTAAGCGAA  7 GGAAGCCCACAGCAAGCGAA,GGGAGCCCACAGTAAGCGAA
+TAGCCTGGTGCGACTGTCAT  7 TAACCTGGTGCGACTGTTAT,TAGCCTGGTGCGACTGTCAT
+TGCGCCAAGTACGATTTCCG  7 TGCGCCAAGTAAGAATTCCG,TGCGCCAAGTACGATTTCCG
+\end{verbatim}
+
+With this type of output, we know that the first cluster
+consists of two distinct sequences, namely the centroid itself
+and \texttt{AGGGGTTACAAGTCTAGGCC}, which together amount to 7
+sequences. This kind of output is useful when you want to convert
+sequences to their centroid.
+
+As mentioned above, `sequences' can also be viewed as input sequences
+in which case they represent different entities, even when their
+nucleotide sequences are the same. These sequences are typically
+different reads from a sequencing experiment. The option
+\texttt{--seq-id} prints the ID numbers of the input sequences that
+end up in each cluster. Here is an example.
+
+\begin{verbatim}
+./starcode -q --seq-id test/test_file.txt
+AGGGCTTACAAGTATAGGCC  7 1,4,5,13,17,20,22
+CCTCATTATTTGTCGCAATG  7 3,14,19,27,30,31,32
+GGGAGCCCACAGTAAGCGAA  7 6,7,10,12,15,24,25
+TAGCCTGGTGCGACTGTCAT  7 8,9,16,21,28,33,35
+TGCGCCAAGTACGATTTCCG  7 2,11,18,23,26,29,34
+\end{verbatim}
+
+With this type of output, we know that the first cluster consists of
+input sequences number 1, 4, 5, 13, 17,20 and 22. This kind of input
+is useful when sequences have different origins and you want to
+know the composition of a cluster relative to the different sources.
+Note that the first input sequence is given ID number 1 and not 0.
+
+Note that options \texttt{--print-clusters} and \texttt{--seq-id} are
+not mutually exclusive, and they may be used together. These options
+have only mild effects on the speed and the memory footprint of
+Starcode.
+
+\section{Input formats}
+
+Starcode accepts FASTA files, FASTQ files and plain text files.
+The directory \texttt{test} contains some example files. We have
+already seen what happens on plain text files in the previous
+examples. For FASTA files, Starcode is run as before. The FASTA
+headers are simply ignored.
+
+\begin{verbatim}
+./starcode -q test/test_file.fasta
+AGGGCTTACAAGTATAGGCC  3
+CCTCATTATTTGTCGCAATG  1
+TGCGCCAAGTACGATTTCCG  1
+\end{verbatim}
+
+FASTQ files present a complication because they can be paired.
+In the case of a single FASTQ file, Starcode is run as before.
+The FASTQ headers and the quality are ignored.
+
+\begin{verbatim}
+./starcode -q test/test_file1.fastq
+AGGGCTTACAAGTATAGGCC  3
+CCTCATTATTTGTCGCAATG  1
+TGCGCCAAGTACGATTTCCG  1
+\end{verbatim}
+
+When there are two paired FASTQ files, you must tell Starcode to
+expect two input files. The first way is to use the \texttt{-1}
+flag for the first file and the \texttt{-2} for the second.
+For more clarity, you can use the longer form \texttt{--input1}
+and \texttt{input2} instead. Below is an example.
+
+\begin{verbatim}
+./starcode -q -1 test/test_file1.fastq -2 test/test_file2.fastq
+AGGGCTTACAAGTATAGGCC/AAGGGCTTACAAGTATAGGC 3
+CCTCATTATTTGTCGCAATG/ACCTCATTATTTGTCGCAAT 1
+TGCGCCAAGTACGATTTCCG/ATGCGCCAAGTACGATTTCC 1
+\end{verbatim}
+
+As you can see, the output also changes. The two paired sequences
+are lumped next to each other and each centroid now consists of
+a pair of sequences. It is important to mention that in this
+mode, the edit distance between two pairs of sequences is the sum
+of the pairwise edit distances. For instance, if the edit distance
+is set to 3 with \texttt{-d3}, the pair \texttt{AAA/TTT} is a
+neighbor of the pair \texttt{CAA/GGT}, but not of the pair
+\texttt{CCC/GGG}.
+
+\section{Getting a non redundant sample}
+
+By default, Starcode clusters the input sequences and outputs the
+clusters. In some application, it is interesting to obtain a non
+redundant sample of the input. In general, it is easy to do so by
+collecting the centroid sequences, but because Starcode discards
+extra information from FASTA and FASTQ files, we provide the
+option \texttt{--non-redundant} for cases that the format has to
+be respected.
+
+\begin{verbatim}
+./starcode -q --non-redundant test/test_file1.fastq
+@seq1/1
+AGGGCTTACAAGTATAGGCC
++
+BBBBBBBBBBBBBBBBBBBB
+@seq3/1
+CCTCATTATTTGTCGCAATG
++
+BBBBBBBBBBBBBBBBBBBB
+@seq2/1
+TGCGCCAAGTACGATTTCCG
++
+BBBBBBBBBBBBBBBBBBBB
+\end{verbatim}
+
+When two FASTQ files are used with this option, nothing is
+printed on the screen. Instead, two new FASTQ files are created.
+Note that the sequences are still sorted in alphabetical order,
+so their order will usually not be the same as the original input.
+
 \end{document}

### 2015-11-18 src/main-starcode.c
Added documentation for seq-id option.

@@ -59,7 +59,7 @@ char *USAGE =
 "  output format options\n"
 "       --print-clusters: outputs cluster compositions\n"
 "       --non-redundant: remove redundant sequences\n"
-"       --seq-id: print sequence numbers\n";
+"       --seq-id: print sequence numbers (1-based)\n";
 
 void say_usage(void) { fprintf(stderr, "%s\n", USAGE); }
 void say_version(void) { fprintf(stderr, VERSION "\n"); }
